### 2015-12-08 ms.tex
Convert to latex, build via remake
For the moment using style from MEE

\documentclass[a4paper,11pt]{article}
\usepackage[osf]{mathpazo}
\usepackage{ms}
\usepackage[]{natbib}
\raggedright

\newcommand{\livingdata}{\texttt{living\_data}}
\newcommand{\smurl}[1]{{\footnotesize\url{#1}}}
\usepackage{graphicx}

\title{{\livingdata}: A package for distributing dynamic and versioned
data}
\author{Richard G. FitzJohn$^1$, and fans$^2$}
\affiliation{
$^1$ Somewhere in the UK\\
$^2$ Everywhere else\\

}
\date{}

\bibliographystyle{mee}

\usepackage[title,titletoc,toc]{appendix}

\mstype{Applications Note}
\runninghead{The {\livingdata} package}
\keywords{}

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item Data is cool
\end{enumerate}

\subsection{Introduction}

In recent years the idea of the data publication has become established
across many different scientific disciplines. Researchers can publish
the data associated with a publication (or even not associated with a
publication) in one of a number of open access locations, such as the
Dryad Data Repository, Figshare, or (less ideally) as supplementary
files to the original publication. These stable data then are freely and
easily available. Researchers get credit for the production (or
assembly) of that data, and the model facilitates future, creative uses
by the research community. As this new model has taken hold, one new
problem that has emerged: many of the most useful datasets are often not
as static as we once throught they were. The static model works very
well for reproducing a \emph{particular} analysis in a \emph{particular}
paper, which is crucial and important for reproducibility.
{[}\textbf{MWP}: I think we should highlight the fact that there is a
range of types of data from one-off studies of a single population at a
single time (where Dryad is a perfectly suitable option) to large
databases (think GenBank) that require expert database management, good
APIs, etc. I think this approach fills a niche somewhere in the middle.
I realize this is mentioned below but think this would perhaps be useful
upfront to set the scope of the problem we want to address.{]}

From a scientific discovery perspective, however, this is not the most
useful model. Many, although not all, of the most useful datasets are
snapshots of existing quantitative knowledge at a given time. For
example, meta-analyses are built on the existing data for a given
question at a given time. When a new study is published the static
dataset from the last meta-analysis immediately becomes obsolete--to
keep up-to-date that new study should be added. However, if we have a
static data distribution system, this is impossible. Another problem
with the static model is the discovery of small errors. For any much
medium-sized database that is repeatedly used, errors will be found, and
there needs to be some organized way to correct them.

\subsection{Versioning data}

Almost all useful databases are living--they \emph{should} get better
through time either by the addition of new data or by improving the
quality of existing data. For example a dataset on plant characteristics
might become obsolete as taxonomic changes occur. This might only
require a small amount of work and a minor change, but if there is no
way to distribute that new dataset, then that work has to be re-done by
every user of that dataset. Moreover, those required changes adds steps
between the canonical dataset and the new analysis (and publication),
making reproducibility of the new analysis more difficult than it needs
to be. If the canonical dataset is versioned--and versions are
accessible--the up-to-date data can be distributed, used, and analyses
based on these data can be reproduced.

\subsection{One solution for small-to-medium sized
datasets}

Versioned data is laudable goal, but implementation is difficult.

For finding the right tool to version data, the size of the dataset is
an important consideration. ``Big data'' requires a separate set of
hardware and software tools (see Amazon web services); examples of big
data include genetic datasets, global datasets species observations, and
remote sensing data. This size of this data is bigger than individual
research groups and often requires a specialized governmental or
non-governmental institution with full-time staff to curate and maintain
those research resources.

We were particularly interested in smaller and more specialized
problems--the type of datasets that a given research group might
specialize in. These are the types of research questions that lend
themselves to periodic analysis (or meta-analyses): - does chocolate
consumption correlate with obesity? - does the rate of spread of
epidemics correlate with transmission rates? - how do plants partition
resources in different conditions? - does the body size of animals
correlate with range size? - which birds have been observed in an annual
survey in a given place?

{[}\textbf{MWP:} I think the focus on questions is perhaps misplaced. We
should instead focus on the qualities of the data that we are interested
in.{]}

Each of those questions might be the captured in a dataset that could be
one lab's or NGO's expertise--or more commonly a compendium of work by
many different groups, harvested from the original researchers or from
the literature. In many cases, the key data for these questions will not
be ``big data''--they'll be small to possibly medium sized. And the
questions may drive a given research line, but are not in most cases
general enough to establish infra-structure for. But crucially the data
themselves will change over time. And currently there is no
infrastructure for organizing and distributing these types of ``living
datasets''.

So what we need in this case is a low-cost (both money and time) way of
making versions of a dataset and delivering it to potential users, along
with notes on what has changed and why since the previous version.
Ideally periodically versions can be deposited and given a digital
object identifier (DOI) at a major research archive, but in between
major versions, small changes can be made, errors can be corrected, and
the most recent version can be distributed to users. Idealy there would
also be a way for users to easily flag errors for the currator to then
correct in the next release. As many of the users of small-to-medium
sized datasets use the open source language R, we also prioritized easy
of use from within R.

For small datasets, it is possible to use the same tools developed for
big data: there are a number of high-setup-cost options (stand-alone
websites, APIs), which have advantages if you have the resources to
develop and maintain them. We wanted a relatively low cost way (both in
money and time).

\subsection{Our system}

Our approach builds on the theory of ``semantic versioning'' developed
for software distribution; this system uses the familiar 1.0.0 notation.
We argue that data versioning is a similar problem, and we are not the
first to make this argument (INSERT LINKS HERE). Applying semantic
versioning to data, a change from 1.0.0 to 1.0.1 implies a tiny changes,
for example a minor error correction. A change from 1.0.0 to 1.1.0
implies a major change, for example adding a new study to a meta-analysis
dataset. A change from 1.0.0 to 2.0.0 implies a very major revision, for
example improving the entire structure of the dataset and adding new
columns.

One of the key problems in switching from a static to a dynamic data
world, is that it's currently hard to set up and maintain a system like
this. Ideally using a versioned database should be easy for both the
user and the data maintainer. This is especially crucial is the data
maintainers are busy. Moreover, old versions need to be preserved in a
stable and accessible way.

To try and achieve to above goals, we built a system to distribute
versions of our datasets. We think that with this infrastructure, it is
relatively easy to set up and maintain a versioned system with any
small-to-medium sized dataset.

In our system, the data is downloaded only once per computer, available
for offline access, easy to port to other platforms, and can get a doi
when desired (i.e.~when it's used for a publication). All old versions
are accessible.

\subsection{Some worked examples}

So far we have three:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \texttt{Biomass And Allometry Database} \citep{Falster-2015}
\item
  \texttt{taxonlookup} \citep{Pennell-2015a}
\item
  \texttt{plantgrowthform}
\end{enumerate}

Each of these is a versioned dataset, designed for a particular research
goal, and also explicitly flexible such that they can grow in the
future. In all three cases, the user needs to install a few R packages.
The data versions themselves are stored as Github Releases (as well as
if desired on DataDryad or FigShare), then any of the versions of the
data can be downloaded and used with only one of \texttt{R} code.

\subsection{How much expertise is needed to use a versioned
dataset?}

Very very basic \texttt{R}. With 4 lines of \texttt{R} code, you can
have the data downloaded and loaded into R and ready for use.

\subsection{How much expertise is needed to set a system like this
up?}

Basically there are two general skills required: 1) understanding
version control via \texttt{git} and 2) writing a very simple R package.
From the prolific Hadley Wickham there is a
\href{http://r-pkgs.had.co.nz/git.html}{very clear introduction to git}
and \href{http://r-pkgs.had.co.nz/intro.html}{an instruction to the
basics of R packages}.

With those two skills the rest should be relatively easy. To set up a
living dataset you need to provide:
\begin{enumerate}
  \item a github repository name (e.g.~wcornwell/taxonlookup, traitecoevo/baad.data).
  \item A single file that will be the versioned dataset (e.g. \texttt{plant\_lookup.csv} or \texttt{baad\_data.zip}).
  \item A function to load that data into R:(e.g.~a simple wrapper around \texttt{read.csv}).
\end{enumerate}

Then as the data changes (hopefully improves) with time, update the data
file and create a github release with a new version number (preferably
using \href{http://semver.org/}{semantic versioning}). All the versions
will be available to the user. And versions can be given a unique DOI
when appropriate, e.g.~when they are used in a publication.

\subsection{Conclusion}

Because it builds on existing infra-structure for distributing software,
this system requires very little (or no) money to set up and only a
small amount of programming expertise. The current infrastructure will
not work for big or continually updated data (e.g.~weather stations or
satellite downloads). For a set of problems that need periodically
updated small-to-medium sized data (e.g.~most meta-analysis problems),
this might be a useful approach.

\bibliography{refs}

\end{document}

### 2015-12-08 R/build.R

@@ -1,0 +1,68 @@
+## TODO: move this into remake TODO: options for saying what command is being
+## run
+
+latex_build_clean <- function(...) {
+  latex_build(..., clean=TRUE)
+}
+
+latex_build <- function(filename, bibliography = NULL, chdir = TRUE, interaction = "nonstopmode",
+  max_attempts = 5L, clean = FALSE, engine = "pdflatex") {
+  if (chdir && dirname(filename) != "") {
+    owd <- setwd(dirname(filename))
+    on.exit(setwd(owd))
+    filename <- basename(filename)
+  }
+
+  res <- run_latex(filename, interaction, engine)
+  if (engine == "xelatex") {
+    res <- run_latex(filename, interaction, engine)
+  }
+  if (!is.null(bibliography)) {
+    run_bibtex(filename)
+    res <- run_latex(filename, interaction, engine)
+  }
+
+  pat <- c("Rerun to get cross-references right", "Rerun to get citations correct",
+    "Rerun to get outlines right")
+  isin <- function(p, x) {
+    any(grepl(p, x))
+  }
+  for (i in seq_len(max_attempts)) {
+    if (any(vapply(pat, isin, logical(1), res))) {
+      res <- run_latex(filename, interaction, engine)
+    } else {
+      break
+    }
+  }
+
+  if (clean) {
+    latex_clean(filename)
+  }
+
+  invisible(NULL)
+}
+
+latex_clean <- function(filename) {
+  filebase <- sub(".tex$", "", filename)
+  exts <- c(".log", ".aux", ".bbl", ".blg", ".fls", ".out", ".snm", ".nav", ".tdo",
+    ".toc")
+  aux <- paste0(filebase, exts)
+  file.remove(aux[file.exists(aux)])
+}
+
+run_latex <- function(filename, interaction = "nonstopmode", engine = "pdflatex") {
+  args <- c(paste0("-interaction=", interaction), "-halt-on-error", filename)
+  call_system(Sys_which(engine), args)
+}
+
+run_bibtex <- function(filename) {
+  call_system(Sys_which("bibtex"), sub(".tex$", "", filename))
+}
+
+Sys_which <- function(x) {
+  ret <- Sys.which(x)
+  if (ret == "") {
+    stop(sprintf("%s not found in $PATH", x))
+  }
+  ret
+}

### 2016-05-31 ms.tex
changed intro paragraph

@@
 
-\newcommand{\livingdata}{\texttt{living\_data}}
+\newcommand{\datastorr}{\texttt{datastorr}}
 \newcommand{\smurl}[1]{{\footnotesize\url{#1}}}
@@
 
-\title{{\livingdata}: A package for distributing dynamic and versioned
+\title{{\datastorr}: A package for distributing dynamic and versioned
 data}
-\author{Richard G. FitzJohn$^1$, and fans$^2$}
+\author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$, Matthew
+  W. Pennell$^3$, and William K. Cornwells$^4$}
 \affiliation{
@@
 $^2$ Everywhere else\\
+$^3$ Department of Zoology and Biodiversity Research Centre,
+University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
 
@@
 \mstype{Applications Note}
-\runninghead{The {\livingdata} package}
+\runninghead{The {\datastorr} package}
 \keywords{}
@@
 
-In recent years the idea of the data publication has become established
-across many different scientific disciplines. Researchers can publish
-the data associated with a publication (or even not associated with a
-publication) in one of a number of open access locations, such as the
-Dryad Data Repository, Figshare, or (less ideally) as supplementary
-files to the original publication. These stable data then are freely and
-easily available. Researchers get credit for the production (or
-assembly) of that data, and the model facilitates future, creative uses
-by the research community. As this new model has taken hold, one new
-problem that has emerged: many of the most useful datasets are often not
-as static as we once throught they were. The static model works very
-well for reproducing a \emph{particular} analysis in a \emph{particular}
-paper, which is crucial and important for reproducibility.
-{[}\textbf{MWP}: I think we should highlight the fact that there is a
-range of types of data from one-off studies of a single population at a
-single time (where Dryad is a perfectly suitable option) to large
-databases (think GenBank) that require expert database management, good
-APIs, etc. I think this approach fills a niche somewhere in the middle.
-I realize this is mentioned below but think this would perhaps be useful
-upfront to set the scope of the problem we want to address.{]}
+Data is quickly becoming a first-class scientific product.
+Funding bodies, publishers, and scientific social norms are increasingly both requiring data to be published alongside research publications and recognizing the value of producing and sharing data for its own sake.
+This has been a rapid, and exciting, sea change and we as a community are still catching up and trying to discern how best to disseminate data and credit data creators and aggregators.
+One thing that is clear is that there is not a one-size-fits-all solution to this problem as data is incredibly heterogeneous.
+On one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}).
+On the other end are repositories for storing a wide variety of static, stand-alone datasets (\texttt{Dryad}, \texttt{Figshare}, and less optimally, as supplements to research papers.
+The static model works very well for reproducing a \emph{particular} analysis in a \emph{particular} paper, which is crucial and important for reproducibility.
+While both of these classes of data are well-served by current infrastructure, we argue that there is a middle ground that does not fit either of these paradigms very well.
 
-From a scientific discovery perspective, however, this is not the most
-useful model. Many, although not all, of the most useful datasets are
-snapshots of existing quantitative knowledge at a given time. For
-example, meta-analyses are built on the existing data for a given
-question at a given time. When a new study is published the static
-dataset from the last meta-analysis immediately becomes obsolete--to
-keep up-to-date that new study should be added. However, if we have a
-static data distribution system, this is impossible. Another problem
-with the static model is the discovery of small errors. For any much
-medium-sized database that is repeatedly used, errors will be found, and
-there needs to be some organized way to correct them.
+Consider the case of the data from an ecological meta-analysis.
+It is only a snapshot of our existing knowledge at a given time.
+When a new study is published the static dataset from the last meta-analysis immediately becomes obsolete --- to keep up-to-date that new study should be added.
+However, if we have a static data distribution system, this is impossible.
+Another problem with the static model is the discovery of small errors.
+For any much medium-sized database that is repeatedly used, errors will be found, and there needs to be some organized way to correct them.
+And while a centralized GenBank-esque repository would be useful for updating or adding to the data from this meta-analysis, the infrastructure investment (designing maintaining, and hosting the database) is clearly overkill for most projects and is a very substantial barrier to entry.
 

### 2017-04-24 ms.tex
Add tables listing features, main aims

@@
 \author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$, Matthew
-  W. Pennell$^3$, and William K. Cornwells$^4$}
+  W. Pennell$^3$, and William K. Cornwell$^2$}
 \affiliation{
 $^1$ Somewhere in the UK\\
-$^2$ Everywhere else\\
+$^2$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
+University of New South Wales, Sydney, NSW 2052, Australia\\
 $^3$ Department of Zoology and Biodiversity Research Centre,
@@
 \end{enumerate}
+
+\section{Overview}
+
+Main aims of paper:
+
+\begin{enumerate}
+  \item Recognise need for versioned datasets
+  \item Provide workflow / infrastructure for people to distribute and access small-medium sized data with minimal setup and cost, by leveraging emerging technologies
+\end{enumerate}
+
+% Access point might change, e.g. github goes out of business
+
+% Workflows for dataset assembly: One or two packages? More complex like baad has two packages. Taxonlookup is all in one
+% Licensnes? Currently using MIT (taxonlookup), BSD\_2 (baad.data)
 
@@
 
+\section{Table}
+
+\begin{table}[h!]
+\centering
+\caption{Alternative frameworks for managing a living database}
+{\footnotesize
+\vspace{1cm}
+  \begin{tabular}{p{5cm}p{5cm}p{5cm}}
+  \hline
+  \textbf{Feature} & \textbf{Traditional web database} & \textbf{Lightweight versioned data}\\ \hline
+   Data storage     & SQL           &  git       \\
+   Webserver        & custom        &  Maintained by github.com\\
+   Frontend         & Ruby on rails &  datastorr \\
+   Ease of setup    & Easy          & Moderate \\ % server, backups, bandwidth
+   Data size        & Small-Very large & Up to 1Gb\\
+   Cost             & Ongoing       & Free \\
+   Bandwidth        & Pro rata      & Managed by github.com\\
+   Maintainer skills& Ruby, php     & R, git \\
+   User skills      & Web           & R \\
+   Versioning       & Hard          & Easy \\
+   DOI minting      & Manual        & Automatic \\
+  \hline
+  \end{tabular}
+  }
+\label{tab:sql_v_versioneddata}
+\end{table}
+
+\newpage
+
+\begin{table}[h!]
+\centering
+\caption{Main technologies leveraged to store and distribute "Lightweight versioned data".}
+{\footnotesize
+\vspace{1cm}
+  \begin{tabular}{p{5cm}p{10cm}}
+  \hline
+  \textbf{Technology} & \textbf{Description} \\\hline
+   git & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data\\
+   github.com & A commercial webplatform for sharing, visualising and managing `git' repositories. Includes issue tracking.\\
+   R     &  Open source statsitcial and data processing language. \\
+   CRAN  &  Open source repository of packages for the R language \\
+   datastorr & R package used to fetch github releases. \\
+   \hline
+  \end{tabular}
+  }
+\label{tab:technologies}
+\end{table}
+
+\newpage
+
+\begin{table}[h!]
+\centering
+\caption{Groups of users and their requirements}
+{\footnotesize
+\vspace{1cm}
+  \begin{tabular}{p{2cm}p{5cm}p{7cm}}
+  \hline
+  \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
+  Maintainer & Create and distribute versioned datasets & Low technical overhead \\
+    & & Easy workflow for releasing new versions \\
+    & & Long term preservation \\
+    & & Easy to crowd-source contributions \\
+    & & Low cost \\
+    & & Low maintenance \\
+  Contributor & Contribute to future versions & Add new data \\
+    & & Report errors in existing data  \\
+  Users (all) & Easy access to releases & Introduction \& overview \\
+    & & Long term stability \\
+    & & Clear path for user to become contributor \\
+  Users (computational, machines) & Scripted access to releases & Easy installation\\
+    & & Long term stability \\
+  \hline
+  \end{tabular}
+  }
+\label{tab:user_requirements}
+\end{table}
+
+\newpage
+
+\begin{table}[h!]
+\centering
+\caption{Alternative platforms for minting DOIs from github repositories}
+{\footnotesize
+\vspace{1cm}
+  \begin{tabular}{p{3cm}p{8cm}}
+  \hline
+  \textbf{Provider} & \textbf{Features} \\ \hline
+  \multicolumn{2}{l}{Zenodo: zenodo.org}\\
+    & Automated DOI for each github release \\
+    & Funded and maintained by CERN and OpenAIRE for long term preservation \\
+    & Only includes files embedded in git repo, not built  by code\\
+  \multicolumn{2}{l}{Figshare: figshare.com}\\
+    & Builds data package based on github release \\
+    & Allows you to upload additional files, e.g. datasets built by code \\
+    & Integrated with several journals, e.g. Nature family, PNAS \\
+    & Some limited potential for versioned DOIs\\
+  \hline
+  \end{tabular}
+  }
+\label{tab:doi_minting}
+\end{table}
+
+\newpage
+
 \bibliography{refs}

### 2017-04-24 R/build.R
Embed call_system command (replaces callr)

@@ -66,3 +66,86 @@ Sys_which <- function(x) {
   }
   ret
 }
+
+##' Function imported from callr package;  makes it easy to call a
+##' system command from R and have it behave.
+##'
+##' This function uses \code{system2} to call a system command fairly
+##' portably.  What it adds is a particular way of dealing with
+##' errors.  \code{call_system} runs the command \code{command} with
+##' arguments \code{args} (and with optionally set environment
+##' variables \code{env}) and hides \emph{all} produced output to
+##' stdout and stderr.  If the command fails (currently any nonzero
+##' exit code is counted as a failure) then \code{call_system} will
+##' throw an R error giving
+##' \itemize{
+##' \item the full string of the command run
+##' \item the exit code of the command
+##' \item any \code{errmsg} attribute that might have been returned
+##' \item all output that the program produced to either stdout and
+##' stderr
+##' }
+##'
+##' This means that a successful invocation of a program produces no
+##' output while the unsuccessful invocation throws an error and
+##' prints all information to the screen (though this is delayed until
+##' failure happens).
+##'
+##'
+##' \code{call_system} also returns the contents of both stderr and
+##' stdout \emph{invisibly} so that it can be inspected if needed.
+##'
+##' The function \code{run_system} does the same thing and will be
+##' removed as soon as code that depends on it is out of use.
+##'
+##' @title Run a system command, stopping on error
+##' @param command The system command to be invoked, as a character
+##' string.  \code{\link{Sys.which}} is useful here.
+##' @param args A character vector of arguments to \code{command}
+##' @param env A character vector of name=value pairs to be set as
+##' environment variables (see \code{\link{system2}}).
+##' @param max_lines Maximum number of lines of program output to
+##' print with the error message.  We may prune further to get the
+##' error message under \code{getOption("warn.length")}, however.
+##' @param p Fraction of the error message to show from the tail of
+##' the output if truncating on error (default is 20\% lines are head,
+##' 80\% is tail).
+##' @param stdout,stderr Passed to \code{system2}.  Set one of these
+##' to \code{FALSE} to avoid capturing output from that stream.  Setting
+##' both to \code{FALSE} is not recommended.
+##' @export
[…]

### 2017-04-24 ms.tex
some work on tables

@@
 \item
+  \texttt{coraltraits}
+\item
   \texttt{Biomass And Allometry Database} \citep{Falster-2015}
@@
 \centering
-\caption{Alternative frameworks for managing a living database}
+\caption{Alternative frameworks for setting up and managing a living database}
 {\footnotesize
 \vspace{1cm}
-  \begin{tabular}{p{5cm}p{5cm}p{5cm}}
+  \begin{tabular}{p{2.5cm}p{3.5cm}p{3.5cm}p{4cm}}
   \hline
-  \textbf{Feature} & \textbf{Traditional web database} & \textbf{Lightweight versioned data}\\ \hline
-   Data storage     & SQL           &  git       \\
-   Webserver        & custom        &  Maintained by github.com\\
-   Frontend         & Ruby on rails &  datastorr \\
-   Ease of setup    & Easy          & Moderate \\ % server, backups, bandwidth
-   Data size        & Small-Very large & Up to 1Gb\\
-   Cost             & Ongoing       & Free \\
-   Bandwidth        & Pro rata      & Managed by github.com\\
-   Maintainer skills& Ruby, php     & R, git \\
-   User skills      & Web           & R \\
-   Versioning       & Hard          & Easy \\
-   DOI minting      & Manual        & Automatic \\
+  \textbf{Feature} & \textbf{Static data}& \textbf{Web database} & \textbf{Lightweight versioned data}\\ 
   \hline
+   Webserver        & dataDryad or others & custom                          &  Maintained by github.com\\
+   Backend          & none & SQL + Ruby-on-rails 			& git + datastorr \\
+   User access      & Web browser & Web browser 				    & R \\
+   Ease of setup    & Very easy & Moderate 							& Easy\\ % server, backups, bandwidth
+   Data size        &Up to several Gb & Small-Very large 				& Up to 1Gb\\
+   Cost             &Varies & Varies  						& Free \\
+   Bandwidth        &managed by dataDryad & Pro rata 						& Managed by github.com\\
+   Maintainer skills &none & Ruby + php 					& R + git \\
+   User skills      &Web browsing& Web browsing  					& basic R \\
+   Versioning       &None& Hard 							& Easy \\
+   DOI minting      &Automatic & Many steps 					& Automatic \\
+  \hline 
+  \\
+
   \end{tabular}
@@
    datastorr & R package used to fetch github releases. \\
+   DOI & Digital object identifier, which refers a user to a single stable digital object \\ 
+   DOI ``minting'' & creation of DOI's \\
    \hline
@@
     & & Long term preservation \\
-    & & Easy to crowd-source contributions \\
-    & & Low cost \\
-    & & Low maintenance \\
+    & & Easy to crowd-source error checking and contributions \\
+    & & Low initial cost \\
+    & & Low on-going maintenance \\
   Contributor & Contribute to future versions & Add new data \\
@@
     & & Long term stability \\
-    & & Clear path for user to become contributor \\
-  Users (computational, machines) & Scripted access to releases & Easy installation\\
+    & & Clear path for users to become contributors \\
+  Users (computational, machines) & Build reproducible products on these data & Scripted access to releases\\
+    & &Easy installation\\
     & & Long term stability \\

### 2017-04-26 ms.tex
Edit text throughout

@@
 \newcommand{\smurl}[1]{{\footnotesize\url{#1}}}
+\newcommand{\ghsmurl}[1]{{\footnotesize\href{}{#1}}}
+
 \usepackage{graphicx}
 
-\title{{\datastorr}: A package for distributing dynamic and versioned
-data}
+\title{Leveraging open platforms to manage and distribute streams of versioned data}
 \author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$, Matthew
@@
 \affiliation{
-$^1$ Somewhere in the UK\\
-$^2$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
+$^1$ Imperial College, London, United Kingdom\\
+$^2$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
 University of New South Wales, Sydney, NSW 2052, Australia\\
-$^3$ Department of Zoology and Biodiversity Research Centre,
+$^3$ Department of Zoology and Biodiversity Research Centre,\\
 University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
@@
 
-\mstype{Applications Note}
-\runninghead{The {\datastorr} package}
+\mstype{Article Note}
+\runninghead{Lightweight versioned data}
 \keywords{}
@@
 % Workflows for dataset assembly: One or two packages? More complex like baad has two packages. Taxonlookup is all in one
-% Licensnes? Currently using MIT (taxonlookup), BSD\_2 (baad.data)
 
@@
 Funding bodies, publishers, and scientific social norms are increasingly both requiring data to be published alongside research publications and recognizing the value of producing and sharing data for its own sake.
-This has been a rapid, and exciting, sea change and we as a community are still catching up and trying to discern how best to disseminate data and credit data creators and aggregators.
-One thing that is clear is that there is not a one-size-fits-all solution to this problem as data is incredibly heterogeneous.
-On one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}).
-On the other end are repositories for storing a wide variety of static, stand-alone datasets (\texttt{Dryad}, \texttt{Figshare}, and less optimally, as supplements to research papers.
-The static model works very well for reproducing a \emph{particular} analysis in a \emph{particular} paper, which is crucial and important for reproducibility.
-While both of these classes of data are well-served by current infrastructure, we argue that there is a middle ground that does not fit either of these paradigms very well.
+This has been a rapid, and exciting, sea change and the scientific community are still catching up and trying to discern how best to disseminate data and credit data creators and aggregators.
+One thing that is clear is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous.
+On one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}).
+On the other end are repositories for storing a wide variety of static, stand-alone datasets (\texttt{Dryad}, \texttt{Figshare}, and less optimally, as supplements to research papers)(Table \ref{tab:sql_v_versioneddata}).
+While both of these classes of data are well-served by current infrastructure, we suggest there is a substantial middle ground that is not well served by either of the existing paradigms, where a lightweight system for serving versioned datasets is needed.
 
-Consider the case of the data from an ecological meta-analysis.
-It is only a snapshot of our existing knowledge at a given time.
-When a new study is published the static dataset from the last meta-analysis immediately becomes obsolete --- to keep up-to-date that new study should be added.
-However, if we have a static data distribution system, this is impossible.
-Another problem with the static model is the discovery of small errors.
-For any much medium-sized database that is repeatedly used, errors will be found, and there needs to be some organized way to correct them.
-And while a centralized GenBank-esque repository would be useful for updating or adding to the data from this meta-analysis, the infrastructure investment (designing maintaining, and hosting the database) is clearly overkill for most projects and is a very substantial barrier to entry.
+In this article we outline why versioned datasets are needed, and provide a relatively easy to use, yet robust and free tool-chain for researchers to maintain, distribute, and access small-medium sized datasets, by leveraging the best of emerging technologies in data science (Table \ref{tab:technologies}), together with a new software package called \texttt{datastorr}.
+While our workflow emphasises particular technologies (Table \ref{tab:technologies}), the principles are general and could easily be ported to other platforms.
 
-\subsection{Versioning data}
+\subsection{Why versioning of data is needed}
 
@@
 quality of existing data. For example a dataset on plant characteristics
-might become obsolete as taxonomic changes occur. This might only
-require a small amount of work and a minor change, but if there is no
-way to distribute that new dataset, then that work has to be re-done by
-every user of that dataset. Moreover, those required changes adds steps
-between the canonical dataset and the new analysis (and publication),
+might become obsolete as taxonomic changes occur. Updating the dataset might only
+require a small amount of work, but if there is no
+way to distribute the updated dataset, then that work has to be re-done by
+every user of that dataset. Moreover, those required changes add steps
+between the canonical (or master) dataset and the new analysis (and publication),
 making reproducibility of the new analysis more difficult than it needs
@@
 accessible--the up-to-date data can be distributed, used, and analyses
-based on these data can be reproduced.
+based on any of the released versions data can be reproduced.
 
-\subsection{One solution for small-to-medium sized
-datasets}
+\subsection{Why versioning of data has been difficult}
 
-Versioned data is laudable goal, but implementation is difficult.
+Providing versioned data is laudable goal, but implementation has been
+difficult. The static model of data distribution widely used in science, where
+datasets are archived either as supplementary materials for a journal article
+or under its own DOI, is simply not designed to provide streams of versioned
+data (Table \ref{tab:sql_v_versioneddata}). A particular issue with the static
+model is that data releases are mostly triggered by publication of academic
+papers. But for a living dataset, there may be multiple releases made in
+between papers. This is particularly the case with the discovery of small
+errors. For any medium-sized dataset that is put to reuse, errors will
+inevitably be found, and there needs to be some easy  way to correct them and
+distribute the updated dataset.
 
-For finding the right tool to version data, the size of the dataset is
-an important consideration. ``Big data'' requires a separate set of
-hardware and software tools (see Amazon web services); examples of big
-data include genetic datasets, global datasets species observations, and
-remote sensing data. This size of this data is bigger than individual
-research groups and often requires a specialized governmental or
-non-governmental institution with full-time staff to curate and maintain
-those research resources.
+To overcome the limitations of static datasets, many research groups have
+setup dedicated webservers to host and distribute data. In these systems,
+datasets areas maintained in a structure database, such as SQL, and a front
+end to this database is written, often using a platform such as Ruby on rails.
+When designed well, such systems are very useful for updating or adding to a
+dataset. However, the infrastructure investment (designing maintaining, and
+hosting the database) is clearly overkill for most projects and is a very
+substantial barrier to entry. And the system requires ongoing funds to keep it
+going, and possibly maintenance. Moreover, the existence of a live webserver,
+does not in itself guarantee ongoing access to different versions of a
+dataset.
 
-We were particularly interested in smaller and more specialized
-problems--the type of datasets that a given research group might
-specialize in. These are the types of research questions that lend
-themselves to periodic analysis (or meta-analyses): - does chocolate
-consumption correlate with obesity? - does the rate of spread of
-epidemics correlate with transmission rates? - how do plants partition
-resources in different conditions? - does the body size of animals
-correlate with range size? - which birds have been observed in an annual
-survey in a given place?
+When considering the right tool to store and deliver versioned data, the size
+of the dataset is another essential consideration. ``Big data'' brings with it
+specific needs and so requires a specific set of hardware and software tools.
+Examples of big data include genetic datasets, global datasets of species
+observations, and remote sensing data. This size of this data is bigger than
+that which individual research groups can maintain and often requires a
+specialized governmental or non-governmental institution with full-time staff
+to curate and maintain those research resources. We were particularly
+interested in smaller and more specialized problems--the type of datasets that
+a given research group might specialize in. In most of these cases, the data
+collected will not be ``big'' but rather small to medium sized. Such data
+support research projects, but are not in most cases general enough to
+establish infra-structure for. But crucially the data themselves will change
+over time. And currently there is no infrastructure for organizing and
+distributing these types of ``living datasets''.
 
-{[}\textbf{MWP:} I think the focus on questions is perhaps misplaced. We
-should instead focus on the qualities of the data that we are interested
-in.{]}
-
-Each of those questions might be the captured in a dataset that could be
-one lab's or NGO's expertise--or more commonly a compendium of work by
-many different groups, harvested from the original researchers or from
-the literature. In many cases, the key data for these questions will not
-be ``big data''--they'll be small to possibly medium sized. And the
-questions may drive a given research line, but are not in most cases
-general enough to establish infra-structure for. But crucially the data
-themselves will change over time. And currently there is no
-infrastructure for organizing and distributing these types of ``living
-datasets''.
-
-So what we need in this case is a low-cost (both money and time) way of
+So what is needed is simple tool-chain for
 making versions of a dataset and delivering it to potential users, along
 with notes on what has changed and why since the previous version.
-Ideally periodically versions can be deposited and given a digital
-object identifier (DOI) at a major research archive, but in between
-major versions, small changes can be made, errors can be corrected, and
-the most recent version can be distributed to users. Idealy there would
-also be a way for users to easily flag errors for the currator to then
-correct in the next release. As many of the users of small-to-medium
-sized datasets use the open source language R, we also prioritized easy
-of use from within R.
+Table \ref{tab:user_requirements} outlines what we consider requirements for
+three groups of people interacting with this system: maintainers,
+contributors, users. The solution is ideally low-cost (both in money and
+time), sophisticated, yet requiring little skill to setup and use.  Ideally
+there would also be a way for users to also become contributors, by flagging
+errors for the curator to correct in the next release, or even for the user to
+fix themselves.
 
-For small datasets, it is possible to use the same tools developed for
-big data: there are a number of high-setup-cost options (stand-alone
-websites, APIs), which have advantages if you have the resources to
-develop and maintain them. We wanted a relatively low cost way (both in
-money and time).
+% For small datasets, it is simply not possible to use the same tools developed for
+% big data: there are a number of high-setup-cost options (stand-alone
+% websites, APIs), which have advantages if you have the resources to
+% develop and maintain them. We wanted a relatively low cost way (both in
+% money and time).
 
-\subsection{Our system}
+\section{A lightweight, cheap and scalable workflow for versioned data}
 
-Our approach builds on the theory of ``semantic versioning'' developed
-for software distribution; this system uses the familiar 1.0.0 notation.
-We argue that data versioning is a similar problem, and we are not the
-first to make this argument (INSERT LINKS HERE). Applying semantic
-versioning to data, a change from 1.0.0 to 1.0.1 implies a tiny changes,
-for example a minor error correction. A change from 1.0.0 to 1.1.0
-implies a major change, for example adding a new study to a meta-analysis
-dataset. A change from 1.0.0 to 2.0.0 implies a very major revision, for
-example improving the entire structure of the dataset and adding new
-columns.
+With the growth of open-source tools and collaborative platforms, we believe
+all the features needed for delivering the vision outlined above are now
+present. The workflow we describe below builds off existing platforms and
+combines them in new ways, and offers the potential for
+widespread uptake at little or no cost.
+
+In brief, the workflow we present borrows best-practices for software
+development  and applies them to the challenge of maintaining and distributing
+versioned data. Software developers maintain a core set of code which produces
+a binary executable file that can be installed on a users local  computers. In
+our system, data developers maintain a core set of files, which produce an
+organised dataset that can be similarly be ``installed'' on a users local
+computer. In either the development of software or data, successive versions
+are released over time (called ``releases''). The similarity in workflow
+between sofwtare and data then allows us to deploy the same technological
+platforms that are used in software development, for the development and
+distribution of data. Importantly, the system we describe uses well-
+established and tools, which ensures high-level performance and stability.
+
+The core technologies used are summarised in Table \ref{tab:technologies} and
+described further below.
+
+\subsection{Version control via `git'}
+
+\texttt{Git} is a version control system that tracks line-by-line changes in text
+files. Git is used extensively in software development, but is increasingly
+also being applied to scientific code and data management \citep{Ram-2013,
+Perkel-2016}. Some key features of git that make attractive for management of data
+include:
+\begin{itemize}
+  \item Tracks all changes in monitored files. These changes are visible to
+  anyone interacting with the repository.
+  \item Allows users to annotate changes (`commits') with informative messages
+  detailing the changes.
+  \item Is a distributed system, meaning anyone who downloads the repository
+  can access the full history and contribute changes.
+  \item Can easily handle file sizes of at least 50Mb and repositories up to 1Gb.
+\end{itemize}
+
+\subsection{Github platform for storing git repositories}
+
+The website Github (\smurl{www.github.com}) is a commercial platform for
+hosting and interacting with git repositories. In practice, each dataset
+should be a separate git repository on github (see Table \ref{tab:examples} for
+examples). Although mainly used for
+computer code, Github is now also being use to manage scientific data
+\citep{Perkel-2016}. Github has a double role in our workflow:
+\begin{enumerate}
+  \item It provides a platform for data contributors to sync their files
+  and correspond about changes in the dataset.
+  \item It hosts a stream of data releases, thus acting as a central
+  point for distribution of the data.
+\end{enumerate}
+
+Particular features of the Github platform that make it attractive for
+handling versioned data include:
+
+\begin{itemize}
+  \item \textbf{Issues}: Each repository has tab where contributors can document issues in the data.
+These issue tags then provide a centralised point for describing a problem and discussing potential solutions.
+  \item \textbf{Pull requests}: Much of the activity on Github is on opens-ource projects with many contributors.
+An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
+  \item \textbf{Releases}: Github is designed to host releases (in this case compiled dataset), associated with a particular commit to the repository.
+  \item \textbf{Tags}: Git commits and releases can be tagged with an informative label.
+  \item \textbf{It's free}: All open-source projects are hosted for free on Github.
+  \item \textbf{Quality}: The Github servers are designed for managing enetrpise level projects, and as such can deliver multiple requests for releases and have sophisticated capabilities.
+  \item \textbf{DOIs}: Github repos can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
+\end{itemize}
+
+\subsection{Semantic versioning}
+
+To address the issue of how successive releases should be labelled, we suggest
+applying the theory of ``semantic versioning'' developed for software
+distribution. This system uses (the potentially familiar) notation of form
+``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted
+(REFS), labelling of data versions presents a similar problem to the labelling
+of software releases: signalling to users the types of change that has
+occurred between successive versions. Applying semantic versioning to data, a
+change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error
+correction. A change from 1.0.0 to 1.1.0 implies a substantial enhancement,
+for example adding a new study to a meta-analysis dataset (while otherwise
+maintaining the same dataset structure). A change from 1.0.0 to 2.0.0 implies
+a very major revision, for example improving the entire structure of the
+dataset and adding new columns.
+
+\subsection{\texttt{R}, \texttt{CRAN}, \texttt{datastorr}}
+
+While Github can host a stream of versioned data releases, we also provide a
+way for users to retrieve the data. One method is to point and click using a
+web browser, then download the data from the Github releases tab. But
+computational users will want something more sophisticated, allowing automated
+retrieval of different versions. Github makes this possible by providing a
+thorough API from which releases can be accessed programatically. And ideally
+load it directly into their analysis platform. As many of the users of small-
+to-medium sized datasets use the open source language \texttt{R}, we also
+prioritized easy of use from within \texttt{R}. Moreover, the  Comprehensive R
+Archive Network (CRAN, \smurl{cran.r-project.org}) offers a facility for users
+to write their own packages, which are then archived and able to be installed
+by anyone running R.
+
+To distribute/retrieve versioned data releases from Github, we have developed
+a  work flow that enables dataset maintainers to write a customised R package
+for their dataset. This workflow uses another R package called
+\texttt{datastorr} to handle all the interactions with github. For example, the package
+\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \ghsmurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples).
+
+Specific features of the \texttt{datastorr} package that make it easy to use include:
+
+\begin{itemize}
+  \item \textbf{Speed}: The data packages contain only the code for retrieving the data, not the data itself, and so are quick to install.
+  \item \textbf{Minimal code}: Specific versions of a dataset can be downlaoded in just two lines of code. Installing the package (if not already installed) takes an additional line.
+  \item \textbf{Efficient}: On any given machine, each dataset is downloaded only once. It is then stored in a standard directory for R packages, and so is available for offline access.
+\end{itemize}
+
+\subsection{Make your own data package}
+
+Table \ref{tab:examples} gives details for XXX example datasets already using
+the distribution system described here. Each of these is a versioned dataset,
+designed for a particular research goal, and also explicitly flexible such
+that they can grow in the future. In all three cases, the user needs to
+install a few R packages. The data versions themselves are stored as Github
+Releases (as well as if desired on DataDryad or FigShare), then any of the
+versions of the data can be downloaded and used with only one of \texttt{R}
+code.
+
+\section{Discussion}
+
+Topics to consider:
+
+\begin{itemize}
+
+\item Addresses important need, leveraging open source platfroms.
+Shows power of open source: by writing single new package, can provide global distribution and access to data, with first class system for managing contents.
+\item Versioning and open access essential for reproducibility
+\item Versioning should enable better benchmarking of models, also because tests can be automated.
+\item We have focussed on particualr set of tools, but could do similar things with other platfroms or languages.
+....
+\end{itemize}
+
+....
+
+(old content unchanged)
 
@@
 
-\subsection{Some worked examples}
+\subsection{How much expertise is needed to setup and use?}
 
-So far we have three:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  \texttt{coraltraits}
-\item
-  \texttt{Biomass And Allometry Database} \citep{Falster-2015}
-\item
-  \texttt{taxonlookup} \citep{Pennell-2015a}
-\item
-  \texttt{plantgrowthform}
-\end{enumerate}
-
-Each of these is a versioned dataset, designed for a particular research
-goal, and also explicitly flexible such that they can grow in the
-future. In all three cases, the user needs to install a few R packages.
-The data versions themselves are stored as Github Releases (as well as
-if desired on DataDryad or FigShare), then any of the versions of the
-data can be downloaded and used with only one of \texttt{R} code.
-
-\subsection{How much expertise is needed to use a versioned
-dataset?}
-
-Very very basic \texttt{R}. With 4 lines of \texttt{R} code, you can
+To use a versioned dataset: Very very basic \texttt{R}. With 4 lines of \texttt{R} code, you can
 have the data downloaded and loaded into R and ready for use.
 
-\subsection{How much expertise is needed to set a system like this
-up?}
-
-Basically there are two general skills required: 1) understanding
+To set a system like this up: Basically there are two general skills required: 1) understanding
 version control via \texttt{git} and 2) writing a very simple R package.
@@
 \begin{enumerate}
-  \item a github repository name (e.g.~wcornwell/taxonlookup, traitecoevo/baad.data).
+  \item a github repository name (e.g.~`wcornwell/taxonlookup`, `traitecoevo/baad.data`).
   \item A single file that will be the versioned dataset (e.g. \texttt{plant\_lookup.csv} or \texttt{baad\_data.zip}).
@@
   \hline
-  \textbf{Feature} & \textbf{Static data}& \textbf{Web database} & \textbf{Lightweight versioned data}\\ 
+  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web database} & \textbf{Lightweight versioned data}\\
   \hline
-   Webserver        & dataDryad or others & custom                          &  Maintained by github.com\\
+   Webserver        & e.g. dataDryad & custom                          &  Maintained by github.com\\
    Backend          & none & SQL + Ruby-on-rails 			& git + datastorr \\
    User access      & Web browser & Web browser 				    & R \\
-   Ease of setup    & Very easy & Moderate 							& Easy\\ % server, backups, bandwidth
-   Data size        &Up to several Gb & Small-Very large 				& Up to 1Gb\\
-   Cost             &Varies & Varies  						& Free \\
+   Ease of setup    & Very easy & Hard 							& Easy\\ % server, backups, bandwidth
+   Data size        & Up to several Gb & Small-Very large 				& Up to 1Gb\\
+   Cost             & Varies & Varies  						& Free \\
    Bandwidth        &managed by dataDryad & Pro rata 						& Managed by github.com\\
@@
 \centering
-\caption{Main technologies leveraged to store and distribute "Lightweight versioned data".}
+\caption{Technologies used to maintain, store and distribute "Lightweight versioned data".}
 {\footnotesize
@@
    git & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data\\
-   github.com & A commercial webplatform for sharing, visualising and managing `git' repositories. Includes issue tracking.\\
-   R     &  Open source statsitcial and data processing language. \\
+   github.com & A commercial webplatform for sharing, visualising and managing `git' repositories. Includes issue tracking\\
+   R     &  Open source statsitcial and data processing language \\
    CRAN  &  Open source repository of packages for the R language \\
-   datastorr & R package used to fetch github releases. \\
+   datastorr & R package used to fetch versioned releases from github  \\
    DOI & Digital object identifier, which refers a user to a single stable digital object \\ 
-   DOI ``minting'' & creation of DOI's \\
+   DOI ``minting'' & creation of a DOI's \\
    \hline
@@
 
+\begin{table}[h!]
+\centering
+\caption{Example datasets using the lightweight versioned data workflow described in this paper.}
+{\footnotesize
+\vspace{1cm}
+
+  \begin{tabular}{p{7cm}p{4cm}p{2cm}}
+  \hline
+   \textbf{URL} & \textbf{R package} & \textbf{Description} \\ \hline
+  \ghsmurl{github.com/dfalster/baad/} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} \citep{Falster-2015} provides...\\
+  \ghsmurl{github.com/traitecoevo/taxonlookup} & \texttt{taxonlookup} & \texttt{taxonlookup} \citep{Pennell-2015a} provides...\\
+  \ghsmurl{github.com/...} & \texttt{plantgrowthform}  & \texttt{plantgrowthform} provides... \\
+  \ghsmurl{github.com/...} & \texttt{fungaltraits} & \texttt{fungaltraits} (in progress) provides... \\
+  \ghsmurl{github.com/...} & \texttt{13C} & \texttt{13C} (in progress) provides .... \\
+  \hline
+  \end{tabular}
+  }
+\label{tab:examples}
+\end{table}
+
+\newpage
+\section{Figures}
+
+\begin{figure}[!hb]
+\centering
+%\includegraphics[width=\linewidth]{figure.pdf}
+\caption{Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.
+\textbf{a)} 3 numbers indicate type of change.
+\textbf{b)} A typical release stream.}
+\label{fig:semantic}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!hb]
+\centering
+%\includegraphics[width=\linewidth]{figure.pdf}
+\caption{Overview of the technologies involved in distributing Lightweight versioned data.}
+\label{fig:technology_stack}
+\end{figure}
 \bibliography{refs}

### 2017-04-26 ms.tex
Clean up GH urls; remove break after  title page

@@
 \newcommand{\smurl}[1]{{\footnotesize\url{#1}}}
-\newcommand{\ghsmurl}[1]{{\footnotesize\href{}{#1}}}
+\newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{#1}}}
 
@@
 \itemsep1pt\parskip0pt\parsep0pt
-\item Data is cool
-\end{enumerate}
-
-\section{Overview}
-
-Main aims of paper:
-
-\begin{enumerate}
+\item Data is cool.
 \item Recognise need for versioned datasets
@@
 \end{enumerate}
-
-% Access point might change, e.g. github goes out of business
-
-% Workflows for dataset assembly: One or two packages? More complex like baad has two packages. Taxonlookup is all in one
 
@@
 \texttt{datastorr} to handle all the interactions with github. For example, the package
-\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \ghsmurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples).
+\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples).
 
@@
 
-\section{Table}
+\newpage
+
+\section{Tables}
 
@@
 \vspace{1cm}
-  \begin{tabular}{p{3cm}p{8cm}}
+  \begin{tabular}{p{4cm}p{8cm}}
   \hline
   \textbf{Provider} & \textbf{Features} \\ \hline
-  \multicolumn{2}{l}{Zenodo: zenodo.org}\\
-    & Automated DOI for each github release \\
+  Zenodo: \smurl{zenodo.org} & Automated DOI for each github release \\
     & Funded and maintained by CERN and OpenAIRE for long term preservation \\
     & Only includes files embedded in git repo, not built  by code\\
-  \multicolumn{2}{l}{Figshare: figshare.com}\\
-    & Builds data package based on github release \\
+  Figshare: \smurl{figshare.com} & Builds data package based on github release \\
     & Allows you to upload additional files, e.g. datasets built by code \\
@@
 
-  \begin{tabular}{p{7cm}p{4cm}p{2cm}}
+  \begin{tabular}{p{3.5cm}p{3cm}p{7cm}}
   \hline
-   \textbf{URL} & \textbf{R package} & \textbf{Description} \\ \hline
-  \ghsmurl{github.com/dfalster/baad/} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} \citep{Falster-2015} provides...\\
-  \ghsmurl{github.com/traitecoevo/taxonlookup} & \texttt{taxonlookup} & \texttt{taxonlookup} \citep{Pennell-2015a} provides...\\
+   \textbf{Github repo} & \textbf{R package} & \textbf{Description} \\ \hline
+  \ghsmurl{dfalster/baad/} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} \citep{Falster-2015} provides...\\
+  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & \texttt{taxonlookup} \citep{Pennell-2015a} provides...\\
   \ghsmurl{github.com/...} & \texttt{plantgrowthform}  & \texttt{plantgrowthform} provides... \\

### 2017-06-23 ms.tex
tried out a (slightly) modified framing in the introduction

@@
 \title{Leveraging open platforms to manage and distribute streams of versioned data}
-\author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$, Matthew
-  W. Pennell$^3$, and William K. Cornwell$^2$}
+\author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$,\\ Matthew
+  W. Pennell$^3$, and William K. Cornwell$^{2,*}$}
 \affiliation{
@@
 University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
-
+$^*$ Corresponding author: w.cornwell@unsw.edu.au\\
 }
@@
 
-Data is quickly becoming a first-class scientific product.
-Funding bodies, publishers, and scientific social norms are increasingly both requiring data to be published alongside research publications and recognizing the value of producing and sharing data for its own sake.
-This has been a rapid, and exciting, sea change and the scientific community are still catching up and trying to discern how best to disseminate data and credit data creators and aggregators.
+Data is now a first-class scientific product.
+Funding bodies, publishers, and scientific social norms are increasingly both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses.
+This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators [Cite some Open Science papers, both pro and con].
 One thing that is clear is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous.
@@
 On the other end are repositories for storing a wide variety of static, stand-alone datasets (\texttt{Dryad}, \texttt{Figshare}, and less optimally, as supplements to research papers)(Table \ref{tab:sql_v_versioneddata}).
-While both of these classes of data are well-served by current infrastructure, we suggest there is a substantial middle ground that is not well served by either of the existing paradigms, where a lightweight system for serving versioned datasets is needed.
+We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.
 
-In this article we outline why versioned datasets are needed, and provide a relatively easy to use, yet robust and free tool-chain for researchers to maintain, distribute, and access small-medium sized datasets, by leveraging the best of emerging technologies in data science (Table \ref{tab:technologies}), together with a new software package called \texttt{datastorr}.
-While our workflow emphasises particular technologies (Table \ref{tab:technologies}), the principles are general and could easily be ported to other platforms.
+In this article we i) introduce the concept of Lightweight Versioned Data (LVD) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be leveraged to help researchers maintain, distribute, and access small to medium sized datasets; and iii) introduce a new R package called \texttt{datastorr}, an implementation of a LVD system.
+While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
+Furthermore, our backgrounds in the life sciences (and ecology and evolution, in particular); however, we argue that the issues we identify and the solutions we put forward are likely to be relevant to a wide variety of domains, across the natural and social sciences.
 

### 2017-07-11 ms.tex
more editing of the text

@@
 quality of existing data. For example a dataset on plant characteristics
-might become obsolete as taxonomic changes occur. Updating the dataset might only
+might become outdated as taxonomic changes occur. Updating the dataset might only
 require a small amount of work, but if there is no
@@
 
-Providing versioned data is laudable goal, but implementation has been
-difficult. The static model of data distribution widely used in science, where
-datasets are archived either as supplementary materials for a journal article
+While the benefits of versioned data are fairly clear, determining how best to implement it has been anything but.
+The static model of data distribution widely used in science, where datasets are archived either as supplementary materials for a journal article
 or under its own DOI, is simply not designed to provide streams of versioned
@@
 hosting the database) is clearly overkill for most projects and is a very
-substantial barrier to entry. And the system requires ongoing funds to keep it
-going, and possibly maintenance. Moreover, the existence of a live webserver,
+substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many groups. Moreover, the existence of a live webserver,
 does not in itself guarantee ongoing access to different versions of a
@@
 
-In brief, the workflow we present borrows best-practices for software
+In brief, the workflow we present, which we call Lightweight Versioned Data (LVD), borrows best-practices for software
 development and applies them to the challenge of maintaining and distributing
 versioned data. Software developers maintain a core set of code which produces
-a binary executable file that can be installed on a users local  computers. In
-our system, data developers maintain a core set of files, which produce an
-organised dataset that can be similarly be ``installed'' on a users local
+a binary executable file that can be installed on a users local  computers. With a LVD system, data developers maintain a core set of files, which produce an
+organised dataset that can be similarly be ``installed'' on a user's local
 computer. In either the development of software or data, successive versions
@@
 platforms that are used in software development, for the development and
-distribution of data. Importantly, the system we describe uses well-
+distribution of data. Importantly, LVD uses well-
 established and tools, which ensures high-level performance and stability.
@@
 also being applied to scientific code and data management \citep{Ram-2013,
-Perkel-2016}. Some key features of git that make attractive for management of data
-include:
+Perkel-2016}. Git is attractive for data management because it: 
 \begin{itemize}
@@
 
-\subsection{Github platform for storing git repositories}
+\subsection{GitHub platform for storing git repositories}
 
-The website Github (\smurl{www.github.com}) is a commercial platform for
+The website GitHub (\smurl{www.github.com}) is a commercial platform for
 hosting and interacting with git repositories. In practice, each dataset
@@
 These issue tags then provide a centralised point for describing a problem and discussing potential solutions.
-  \item \textbf{Pull requests}: Much of the activity on Github is on opens-ource projects with many contributors.
+  \item \textbf{Pull requests}: Much of the activity on GitHub is on open-source projects with many contributors.
 An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
-  \item \textbf{Releases}: Github is designed to host releases (in this case compiled dataset), associated with a particular commit to the repository.
+  \item \textbf{Releases}: GitHub is designed to host releases (in this case compiled datasets), associated with a particular commit to the repository.
   \item \textbf{Tags}: Git commits and releases can be tagged with an informative label.
-  \item \textbf{It's free}: All open-source projects are hosted for free on Github.
-  \item \textbf{Quality}: The Github servers are designed for managing enetrpise level projects, and as such can deliver multiple requests for releases and have sophisticated capabilities.
+  \item \textbf{It's free}: All open-source projects are hosted for free on GitHub.
+  \item \textbf{Quality}: The GitHub servers are designed for managing enetrpise level projects, and as such can deliver multiple requests for releases and have sophisticated capabilities.
   \item \textbf{DOIs}: Github repos can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
@@
 correction. A change from 1.0.0 to 1.1.0 implies a substantial enhancement,
-for example adding a new study to a meta-analysis dataset (while otherwise
+for example adding a new study to a meta-analysis dataset (while otherwise[]
 maintaining the same dataset structure). A change from 1.0.0 to 2.0.0 implies
@@
 
-\subsection{\texttt{R}, \texttt{CRAN}, \texttt{datastorr}}
+\subsection{\texttt{datastorr} and dataset-specific R packages}
 
-While Github can host a stream of versioned data releases, we also provide a
-way for users to retrieve the data. One method is to point and click using a
-web browser, then download the data from the Github releases tab. But
-computational users will want something more sophisticated, allowing automated
-retrieval of different versions. Github makes this possible by providing a
-thorough API from which releases can be accessed programatically. And ideally
-load it directly into their analysis platform. As many of the users of small-
-to-medium sized datasets use the open source language \texttt{R}, we also
-prioritized easy of use from within \texttt{R}. Moreover, the  Comprehensive R
-Archive Network (CRAN, \smurl{cran.r-project.org}) offers a facility for users
-to write their own packages, which are then archived and able to be installed
-by anyone running R.
-
-To distribute/retrieve versioned data releases from Github, we have developed
-a  work flow that enables dataset maintainers to write a customised R package
-for their dataset. This workflow uses another R package called
-\texttt{datastorr} to handle all the interactions with github. For example, the package
+While GitHub can host a stream of versioned data releases, many researchers will likely want to be able to access this data programmatically, and ideally, load it directly into their analysis platform.
+Our solution is for researchers releasing a dataset to write a customized R package, specific to their dataset.
+For example, the package
 \texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples).
+To enable this, we have developed a new R package called \texttt{datastorr}, which does all the heavy lifting both in terms of creating templates for the data-specific R packages and for interacting with the GitHub API.
+With \texttt{datastorr}, setting up a new data-specific package is straightforward and requires minimal computational skills.
 
@@
 
-\subsection{Make your own data package}
-
-Table \ref{tab:examples} gives details for XXX example datasets already using
-the distribution system described here. Each of these is a versioned dataset,
-designed for a particular research goal, and also explicitly flexible such
-that they can grow in the future. In all three cases, the user needs to
-install a few R packages. The data versions themselves are stored as Github
-Releases (as well as if desired on DataDryad or FigShare), then any of the
-versions of the data can be downloaded and used with only one of \texttt{R}
-code.
+We have created \texttt{datastorr} to make LVD easy to implement and manage for \texttt{R} users.
+And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
+However, we can imagine alternative methods for interacting with GitHub and building templates; so while useful, adopting LVD does not critically depend on using \texttt{datastorr}.
+Furthermore, the concept of LVD is general and we would love to see implementations in \texttt{Python} or other languages.
 

### 2017-07-11 ms.tex
more editing

@@
 
+[MWP: moved this from discussion]
+To set up a
+living dataset you need to provide:
+\begin{enumerate}
+  \item a github repository name (e.g.~`wcornwell/taxonlookup`, `traitecoevo/baad.data`).
+  \item A single file that will be the versioned dataset (e.g. \texttt{plant\_lookup.csv} or \texttt{baad\_data.zip}).
+  \item A function to load that data into R:(e.g.~a simple wrapper around \texttt{read.csv}).
+\end{enumerate}
+
+Then as the data changes (hopefully improves) with time, update the data
+file and create a github release with a new version number (preferably
+using \href{http://semver.org/}{semantic versioning}). All the versions
+will be available to the user. And versions can be given a unique DOI
+when appropriate, e.g.~when they are used in a publication.
+
 We have created \texttt{datastorr} to make LVD easy to implement and manage for \texttt{R} users.
@@
 
-....
-
-(old content unchanged)
-
-One of the key problems in switching from a static to a dynamic data
-world, is that it's currently hard to set up and maintain a system like
-this. Ideally using a versioned database should be easy for both the
-user and the data maintainer. This is especially crucial is the data
-maintainers are busy. Moreover, old versions need to be preserved in a
-stable and accessible way.
-
-To try and achieve to above goals, we built a system to distribute
-versions of our datasets. We think that with this infrastructure, it is
-relatively easy to set up and maintain a versioned system with any
-small-to-medium sized dataset.
-
-In our system, the data is downloaded only once per computer, available
-for offline access, easy to port to other platforms, and can get a doi
-when desired (i.e.~when it's used for a publication). All old versions
-are accessible.
-
-\subsection{How much expertise is needed to setup and use?}
-
-To use a versioned dataset: Very very basic \texttt{R}. With 4 lines of \texttt{R} code, you can
-have the data downloaded and loaded into R and ready for use.
-
-To set a system like this up: Basically there are two general skills required: 1) understanding
-version control via \texttt{git} and 2) writing a very simple R package.
-From the prolific Hadley Wickham there is a
-\href{http://r-pkgs.had.co.nz/git.html}{very clear introduction to git}
-and \href{http://r-pkgs.had.co.nz/intro.html}{an instruction to the
-basics of R packages}.
-
-With those two skills the rest should be relatively easy. To set up a
-living dataset you need to provide:
-\begin{enumerate}
-  \item a github repository name (e.g.~`wcornwell/taxonlookup`, `traitecoevo/baad.data`).
-  \item A single file that will be the versioned dataset (e.g. \texttt{plant\_lookup.csv} or \texttt{baad\_data.zip}).
-  \item A function to load that data into R:(e.g.~a simple wrapper around \texttt{read.csv}).
-\end{enumerate}
-
-Then as the data changes (hopefully improves) with time, update the data
-file and create a github release with a new version number (preferably
-using \href{http://semver.org/}{semantic versioning}). All the versions
-will be available to the user. And versions can be given a unique DOI
-when appropriate, e.g.~when they are used in a publication.
+%(old content unchanged)
+%
+%One of the key problems in switching from a static to a dynamic data
+%world, is that it's currently hard to set up and maintain a system like
+%this. Ideally using a versioned database should be easy for both the
+%user and the data maintainer. This is especially crucial is the data
+%maintainers are busy. Moreover, old versions need to be preserved in a
+%stable and accessible way.
+%
+%To try and achieve to above goals, we built a system to distribute
+%versions of our datasets. We think that with this infrastructure, it is
+%relatively easy to set up and maintain a versioned system with any
+%small-to-medium sized dataset.
+%
+%In our system, the data is downloaded only once per computer, available
+%for offline access, easy to port to other platforms, and can get a doi
+%when desired (i.e.~when it's used for a publication). All old versions
+%are accessible.
+%
+%
+%
+%\subsection{How much expertise is needed to setup and use?}
+%
+%To use a versioned dataset: Very very basic \texttt{R}. With 4 lines of \texttt{R} code, you can
+%have the data downloaded and loaded into R and ready for use.
+%
+%To set a system like this up: Basically there are two general skills required: 1) understanding
+%version control via \texttt{git} and 2) writing a very simple R package.
+%From the prolific Hadley Wickham there is a
+%\href{http://r-pkgs.had.co.nz/git.html}{very clear introduction to git}
+%and \href{http://r-pkgs.had.co.nz/intro.html}{an instruction to the
+%basics of R packages}.
 

### 2017-07-18 ms.tex
New title, name for concept, and spelling
Closes #12 Closes #13

@@
 
-\title{Leveraging open platforms to manage and distribute streams of versioned data}
+\title{Versioned data: why it’s needed and how it can be achieved (easily and cheaply)}
 \author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$,\\ Matthew
@@
 
-In this article we i) introduce the concept of Lightweight Versioned Data (LVD) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be leveraged to help researchers maintain, distribute, and access small to medium sized datasets; and iii) introduce a new R package called \texttt{datastorr}, an implementation of a LVD system.
+In this article we i) introduce the concept of Versioned Data Delivery (VDD) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be leveraged to help researchers maintain, distribute, and access small to medium sized datasets; and iii) introduce a new R package called \texttt{datastorr}, an implementation of a VDD system.
 While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
@@
 
+DF: Bring out general features more. Right now git delivers this.
 With the growth of open-source tools and collaborative platforms, we believe
@@
 
-In brief, the workflow we present, which we call Lightweight Versioned Data (LVD), borrows best-practices for software
+In brief, the workflow we present, which we call Versioned Data Delivery (VDD), borrows best-practices for software
 development and applies them to the challenge of maintaining and distributing
 versioned data. Software developers maintain a core set of code which produces
-a binary executable file that can be installed on a users local  computers. With a LVD system, data developers maintain a core set of files, which produce an
+a binary executable file that can be installed on a users local  computers. With a VDD system, data developers maintain a core set of files, which produce an
 organised dataset that can be similarly be ``installed'' on a user's local
@@
 are released over time (called ``releases''). The similarity in workflow
-between sofwtare and data then allows us to deploy the same technological
+between software and data then allows us to deploy the same technological
 platforms that are used in software development, for the development and
-distribution of data. Importantly, LVD uses well-
+distribution of data. Importantly, VDD uses well-
 established and tools, which ensures high-level performance and stability.
@@
   \item \textbf{Speed}: The data packages contain only the code for retrieving the data, not the data itself, and so are quick to install.
-  \item \textbf{Minimal code}: Specific versions of a dataset can be downlaoded in just two lines of code. Installing the package (if not already installed) takes an additional line.
+  \item \textbf{Minimal code}: Specific versions of a dataset can be downloaded in just two lines of code. Installing the package (if not already installed) takes an additional line.
   \item \textbf{Efficient}: On any given machine, each dataset is downloaded only once. It is then stored in a standard directory for R packages, and so is available for offline access.
@@
 
-We have created \texttt{datastorr} to make LVD easy to implement and manage for \texttt{R} users.
+We have created \texttt{datastorr} to make VDD easy to implement and manage for \texttt{R} users.
 And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
-However, we can imagine alternative methods for interacting with GitHub and building templates; so while useful, adopting LVD does not critically depend on using \texttt{datastorr}.
-Furthermore, the concept of LVD is general and we would love to see implementations in \texttt{Python} or other languages.
+However, we can imagine alternative methods for interacting with GitHub and building templates; so while useful, adopting VDD does not critically depend on using \texttt{datastorr}.
+Furthermore, the concept of VDD is general and we would love to see implementations in \texttt{Python} or other languages.
 
@@
 
-\item Addresses important need, leveraging open source platfroms.
+\item Addresses important need, leveraging (use) open source platfroms.
 Shows power of open source: by writing single new package, can provide global distribution and access to data, with first class system for managing contents.
@@
 \item Versioning should enable better benchmarking of models, also because tests can be automated.
-\item We have focussed on particualr set of tools, but could do similar things with other platfroms or languages.
+\item We have focussed on particular set of tools, but could do similar things with other platforms or languages.
 ....

### 2017-07-19 ms.tex
work on datastorr section

@@
 \mstype{Article Note}
-\runninghead{Lightweight versioned data}
+\runninghead{Versioned data delivery}
 \keywords{}
@@
 hosting the database) is clearly overkill for most projects and is a very
-substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many groups. Moreover, the existence of a live webserver,
+substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many research groups. Moreover, the existence of a live webserver,
 does not in itself guarantee ongoing access to different versions of a
@@
 specialized governmental or non-governmental institution with full-time staff
-to curate and maintain those research resources. We were particularly
-interested in smaller and more specialized problems--the type of datasets that
-a given research group might specialize in. In most of these cases, the data
+to curate and maintain those research resources. Here we are particularly
+interested in smaller and more specialized problems (e.g., cases where a single 
+research group contributes most of the data). In most of these cases, the data
 collected will not be ``big'' but rather small to medium sized. Such data
@@
 \subsection{Version control via `git'}
+
+MP: I think these sections on git would benefit from being turned into paragraphs
 
@@
 
-While GitHub can host a stream of versioned data releases, many researchers will likely want to be able to access this data programmatically, and ideally, load it directly into their analysis platform.
+GitHub can be used right ``out of the box'' to created versioned data releases
+(ideally following semantic versioning standards). Both database contributors and end users
+can go onto the website and up- or download the latest (or a previous) version of a dataset.
+However, for many use cases, this workflow is not really that efficient. Ideally, researchers
+should be able to access the database programmatically. To this end, we offer a novel solution.
+We have written a R package, \texttt{datastorr}, that a) interacts with the GitHub API to pull 
+down versions of the dataset; and b) constructs the shell of a second, database specific R
+package, which can be used to access a given dataset. Using this system, a researcher can create
+and distribute a R package that facilitates access to her data with (very) minimal computational skills.
+
+%While GitHub can host a stream of versioned data releases, many researchers will likely want to be able to access this data programmatically, and ideally, load it directly into their analysis platform.
 Our solution is for researchers releasing a dataset to write a customized R package, specific to their dataset.
 For example, the package
-\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples).
+%\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples).
 To enable this, we have developed a new R package called \texttt{datastorr}, which does all the heavy lifting both in terms of creating templates for the data-specific R packages and for interacting with the GitHub API.

### 2017-07-20 ms.tex
more writing in the datastorr section

@@
 \mstype{Article Note}
-\runninghead{Versioned data delivery}
+\runninghead{Versioned Data Delivery}
 \keywords{}
@@
 GitHub can be used right ``out of the box'' to created versioned data releases
-(ideally following semantic versioning standards). Both database contributors and end users
-can go onto the website and up- or download the latest (or a previous) version of a dataset.
-However, for many use cases, this workflow is not really that efficient. Ideally, researchers
-should be able to access the database programmatically. To this end, we offer a novel solution.
-We have written a R package, \texttt{datastorr}, that a) interacts with the GitHub API to pull 
-down versions of the dataset; and b) constructs the shell of a second, database specific R
-package, which can be used to access a given dataset. Using this system, a researcher can create
-and distribute a R package that facilitates access to her data with (very) minimal computational skills.
+(ideally following semantic versioning standards).
+Both database contributors and end users can go onto the website and up- or download the latest (or a previous) version of a dataset.
+However, for many use cases, this workflow is not really that efficient and may create barriers for reproducibility.
+Ideally, researchers should be able to access the database programmatically.
+To this end, we offer a novel solution.
+We have written a R package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the GitHub API to pull down versions of the dataset; and b) constructs the shell of a second, database specific R package, which can be used to access the dataset.
+Using this system, a researcher can create and distribute a R package that facilitates access to her data with (very) minimal computational skills.
 
-%While GitHub can host a stream of versioned data releases, many researchers will likely want to be able to access this data programmatically, and ideally, load it directly into their analysis platform.
-Our solution is for researchers releasing a dataset to write a customized R package, specific to their dataset.
-For example, the package
-%\texttt{baad.data} can retrieve data releases from the Biomass and Allometry Database, stored at \smurl{github.com/dfalster/baad/} (see Table \ref{tab:examples} for more examples).
-To enable this, we have developed a new R package called \texttt{datastorr}, which does all the heavy lifting both in terms of creating templates for the data-specific R packages and for interacting with the GitHub API.
-With \texttt{datastorr}, setting up a new data-specific package is straightforward and requires minimal computational skills.
+For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/dfalster/baad.data}), which is an interface to the Biomass and Allometry Database (BAAD) stored at \smurl{github.com/dfalster/baad}.
+\texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}.
+And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's harddrive.
+When the function \texttt{baad.data::baad_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline).
+If a researcher requires an older version of BAAD, perhaps to reproduce an analysis, specifying the version number of the data as an argument in \texttt{baad.data::baad_data()} will retrieve and store this, without overwriting any other dataset versions they may have previously downloaded.
 
-Specific features of the \texttt{datastorr} package that make it easy to use include:
+Using \texttt{datastorr}, researchers can set up their own versioned database (and R package) simply by providing: 1) a github repository name (e.g., ``traitecoevo/taxonlookup''); 2) a single file that will contain the versioned data (e.g., ``taxonlookup.csv''); 3) and the function used to load this into \texttt{R} (e.g., \texttt{read.csv()}).
 
-\begin{itemize}
-  \item \textbf{Speed}: The data packages contain only the code for retrieving the data, not the data itself, and so are quick to install.
-  \item \textbf{Minimal code}: Specific versions of a dataset can be downloaded in just two lines of code. Installing the package (if not already installed) takes an additional line.
-  \item \textbf{Efficient}: On any given machine, each dataset is downloaded only once. It is then stored in a standard directory for R packages, and so is available for offline access.
-\end{itemize}
-
-[MWP: moved this from discussion]
-To set up a
-living dataset you need to provide:
-\begin{enumerate}
-  \item a github repository name (e.g.~`wcornwell/taxonlookup`, `traitecoevo/baad.data`).
-  \item A single file that will be the versioned dataset (e.g. \texttt{plant\_lookup.csv} or \texttt{baad\_data.zip}).
-  \item A function to load that data into R:(e.g.~a simple wrapper around \texttt{read.csv}).
-\end{enumerate}
-
-Then as the data changes (hopefully improves) with time, update the data
+Then as the data changes (and hopefully improves) with time, update the data
 file and create a github release with a new version number (preferably
@@
 And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
-However, we can imagine alternative methods for interacting with GitHub and building templates; so while useful, adopting VDD does not critically depend on using \texttt{datastorr}.
-Furthermore, the concept of VDD is general and we would love to see implementations in \texttt{Python} or other languages.
+However, we can imagine alternative methods for interacting with GitHub and building templates and adopting a VDD system does not critically depend on using \texttt{datastorr}.
+Indeed, we would love to see implementations in \texttt{Python} or other languages.
 

### 2017-08-03 ms.tex
pointing people where to see how to do use datastorr; leaving it outside of the ms allows the pkg to change as needed
Minor bug fixes

@@
 And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's harddrive.
-When the function \texttt{baad.data::baad_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline).
-If a researcher requires an older version of BAAD, perhaps to reproduce an analysis, specifying the version number of the data as an argument in \texttt{baad.data::baad_data()} will retrieve and store this, without overwriting any other dataset versions they may have previously downloaded.
+When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline).
+If a researcher requires an older version of BAAD, perhaps to reproduce an analysis, specifying the version number of the data as an argument in \texttt{baad.data::baad\_data()} will retrieve and store this, without overwriting any other dataset versions they may have previously downloaded.
 
 Using \texttt{datastorr}, researchers can set up their own versioned database (and R package) simply by providing: 1) a github repository name (e.g., ``traitecoevo/taxonlookup''); 2) a single file that will contain the versioned data (e.g., ``taxonlookup.csv''); 3) and the function used to load this into \texttt{R} (e.g., \texttt{read.csv()}).
+A tutorial explaining precisely how to set this up is available at \smurl{github.com/richfitz/datastorr}.
 

### 2017-08-03 ms.tex
Import Figure illustrating semantic versioning
For #10

Nice figure @wcornwell
some work on the text

@@
 \linenumbers
+
+% For Analyses and Articles, the main text (excluding abstract, Methods, references and figure legends) is approximately 3,000 words.
+The abstract is typically 100-170 words, unreferenced.
+An Introduction is followed by sections headed Results, Discussion, Methods.
+The Methods and Results should be divided by topical subheadings; the Discussion does not contain subheadings.
+The Methods should be followed by References, Data Citations (optional), Acknowledgements and a Competing interests statement.
+A Data Citation section should be included when externally hosted datasets are referred to in the work and each can be cited formally with a unique, stable identifier and repository name.
+
 \section{Summary}
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
-\item Data is cool.
-\item Recognise need for versioned datasets
-\item Provide workflow / infrastructure for people to distribute and access small-medium sized data with minimal setup and cost, by leveraging emerging technologies
-\end{enumerate}
 
-\subsection{Introduction}
+Sharing data has become a key part of reproducible workflows in science.
+To help this process, numerous linked tools have been developed to allow quick and easy data sharing.
+Thus far, however, data has been largely considered a fixed and unchanging product---the particular data file for a particular analysis can be quickly unloaded, given a digital identifier, and shared.
+While this is highly useful, this framework does not account for on-going scientific progress: for many important problems, epitomised by problems that lead to meta-analysis, datasets will continue to grow with time---more data points will be added, and new and more useful data structures will be created.
+Essentially this means that data, like science in general, progress with time.
+We suggest that a versioning system for data will facilitate this process.
+Moreover, while the idea of versioned data has been taken up by large data sharing organizations, we present new tools for creating, archiving, and distributing versioned data easily, quickly and cheaply.
+These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.
+
+\section{Introduction}
 
@@
 
-\section{A lightweight, cheap and scalable workflow for versioned data}
+\section{Results}
 
-DF: Bring out general features more. Right now git delivers this.
-With the growth of open-source tools and collaborative platforms, we believe
-all the features needed for delivering the vision outlined above are now
-present. The workflow we describe below builds off existing platforms and
-combines them in new ways, and offers the potential for
-widespread uptake at little or no cost.
+\subsection{A lightweight, cheap and scalable workflow for versioned data}
 
-In brief, the workflow we present, which we call Versioned Data Delivery (VDD), borrows best-practices for software
-development and applies them to the challenge of maintaining and distributing
-versioned data. Software developers maintain a core set of code which produces
-a binary executable file that can be installed on a users local  computers. With a VDD system, data developers maintain a core set of files, which produce an
-organised dataset that can be similarly be ``installed'' on a user's local
-computer. In either the development of software or data, successive versions
-are released over time (called ``releases''). The similarity in workflow
-between software and data then allows us to deploy the same technological
-platforms that are used in software development, for the development and
-distribution of data. Importantly, VDD uses well-
-established and tools, which ensures high-level performance and stability.
+% DF: Bring out general features more.
+Right now git delivers this.
+With the growth of open-source tools and collaborative platforms, we believe all the features needed for delivering the vision outlined above are now present.
+The workflow we describe below builds off existing platforms and combines them in new ways, and offers the potential for widespread uptake at little or no cost.
 
-The core technologies used are summarised in Table \ref{tab:technologies} and
-described further below.
+In brief, the workflow we present, which we call Versioned Data Delivery (VDD), borrows best-practices for software development and applies them to the challenge of maintaining and distributing versioned data.
+Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local  computers.
+With a VDD system, data developers maintain a core set of files, which produce an organised dataset that can be similarly be ``installed'' on a user's local computer.
+In either the development of software or data, successive versions are released over time (called ``releases'').
+The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
+Importantly, VDD uses well-established and tools, which ensures high-level performance and stability.
 
-\subsection{Version control via `git'}
+The core technologies used are summarised in Table \ref{tab:technologies} and described further below.
 
-MP: I think these sections on git would benefit from being turned into paragraphs
+\subsection{Version control}
 
-\texttt{Git} is a version control system that tracks line-by-line changes in text
-files. Git is used extensively in software development, but is increasingly
-also being applied to scientific code and data management \citep{Ram-2013,
-Perkel-2016}. Git is attractive for data management because it: 
-\begin{itemize}
-  \item Tracks all changes in monitored files. These changes are visible to
-  anyone interacting with the repository.
-  \item Allows users to annotate changes (`commits') with informative messages
-  detailing the changes.
-  \item Is a distributed system, meaning anyone who downloads the repository
-  can access the full history and contribute changes.
-  \item Can easily handle file sizes of at least 50Mb and repositories up to 1Gb.
-\end{itemize}
+Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.
+Essentially version control tracks line-by-line changes in text files and creates and maintains a history of those changes.
+Increasingly version control has been applied to scientific code and data management, especially for small to medium sized datasets \citep{Ram-2013, Perkel-2016}.
+Git is attractive for data management because it tracks all changes in monitored files.
+This change history is visible to anyone interacting with the repository.
+It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.
 
-\subsection{GitHub platform for storing git repositories}
+Because \texttt{git} was built from the ground up for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data.
+For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
+There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future (REF).
+This future system would be faster computationally, smaller in file size, and thus allow for larger datasets to be versioned efficiently.
+That said, the workflow we describe here currently works well with \texttt{git} and should work more smoothly with its future replacement.
 
-The website GitHub (\smurl{www.github.com}) is a commercial platform for
-hosting and interacting with git repositories. In practice, each dataset
-should be a separate git repository on github (see Table \ref{tab:examples} for
-examples). Although mainly used for
-computer code, Github is now also being use to manage scientific data
-\citep{Perkel-2016}. Github has a double role in our workflow:
-\begin{enumerate}
-  \item It provides a platform for data contributors to sync their files
-  and correspond about changes in the dataset.
-  \item It hosts a stream of data releases, thus acting as a central
-  point for distribution of the data.
-\end{enumerate}
+\subsection{Cloud platforms for storing and distributing version controlled data}
 
-Particular features of the Github platform that make it attractive for
-handling versioned data include:
+There are many current platforms for storing and distributing versioned controlled data for low or no cost.
+The website GitHub (\smurl{www.github.com}) is a commercial platform for hosting and interacting with git repositories.
+In practice, each dataset should be a separate git repository on github (see Table \ref{tab:examples} for examples).
+Although mainly used for computer code, Github is now also being use to manage scientific data \citep{Perkel-2016}.
+Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
 
-\begin{itemize}
-  \item \textbf{Issues}: Each repository has tab where contributors can document issues in the data.
-These issue tags then provide a centralised point for describing a problem and discussing potential solutions.
-  \item \textbf{Pull requests}: Much of the activity on GitHub is on open-source projects with many contributors.
+\subsection{Allowing for one or more data contributers}
+
+TODO: turn this into a paragraph
+
+\textbf{Pull requests}: Much of the activity on GitHub is on open-source projects with many contributors.
 An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
-  \item \textbf{Releases}: GitHub is designed to host releases (in this case compiled datasets), associated with a particular commit to the repository.
-  \item \textbf{Tags}: Git commits and releases can be tagged with an informative label.
-  \item \textbf{It's free}: All open-source projects are hosted for free on GitHub.
-  \item \textbf{Quality}: The GitHub servers are designed for managing enetrpise level projects, and as such can deliver multiple requests for releases and have sophisticated capabilities.
-  \item \textbf{DOIs}: Github repos can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
-\end{itemize}
 
-\subsection{Semantic versioning}
+\subsection{`Semantic' versioning}
 
-To address the issue of how successive releases should be labelled, we suggest
-applying the theory of ``semantic versioning'' developed for software
-distribution. This system uses (the potentially familiar) notation of form
-``1.0.0'' for successive versions (Fig \ref{fig:semantic}). As has been noted
-(REFS), labelling of data versions presents a similar problem to the labelling
-of software releases: signalling to users the types of change that has
-occurred between successive versions. Applying semantic versioning to data, a
-change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error
-correction. A change from 1.0.0 to 1.1.0 implies a substantial enhancement,
-for example adding a new study to a meta-analysis dataset (while otherwise[]
-maintaining the same dataset structure). A change from 1.0.0 to 2.0.0 implies
-a very major revision, for example improving the entire structure of the
-dataset and adding new columns.
+To fully realize the benefits of versioned data, the versioning system needs to be communicated to the user.
+Crucially a user should be immediately aware of the relationship among versions.
+Software development has dealt with a very similar problem, and we suggest there is some benefit in adopting the notation from that field.
+Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution.
+This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}).
+As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions.
+Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction.
+A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise[] maintaining the same dataset structure).
+A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.
+
+\subsection{Digital object identifiers}
+
+Tools for creating and archiving digital objects have seen rapid progress in recent years.
+At present, each version of the dataset can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
+Here we describe the particular integration of github and zenodo.
+
+TODO: FLESH THIS OUT AND ADD REFERENCE TO BLOG POST AND OTHER THINKING ON THIS
+
+\subsection{Distributing data to different user needs}
+
+At current there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.
+The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements})).
+GitHub has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).
+Those "releases" can be downloaded directly by users via point-and-click.
+Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.
+Releases can also be linked to code-based users via an R package which we describe in the following section.
 
@@
 
-GitHub can be used right ``out of the box'' to created versioned data releases
-(ideally following semantic versioning standards).
-Both database contributors and end users can go onto the website and up- or download the latest (or a previous) version of a dataset.
-However, for many use cases, this workflow is not really that efficient and may create barriers for reproducibility.
-Ideally, researchers should be able to access the database programmatically.
+For many use cases, a point-and-click workflow may create barriers for reproducibility.
+Ideally, users should be able to also able access the all versions of the database programmatically.
 To this end, we offer a novel solution.
@@
 A tutorial explaining precisely how to set this up is available at \smurl{github.com/richfitz/datastorr}.
+% is this t
 
@@
 using \href{http://semver.org/}{semantic versioning}). All the versions
-will be available to the user. And versions can be given a unique DOI
-when appropriate, e.g.~when they are used in a publication.
+will be available to the user.
 
@@
 
-\item Addresses important need, leveraging (use) open source platfroms.
+\item Addresses important need, leveraging (use) open source platforms.
 Shows power of open source: by writing single new package, can provide global distribution and access to data, with first class system for managing contents.
@@
   \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & \texttt{taxonlookup} \citep{Pennell-2015a} provides...\\
-  \ghsmurl{github.com/...} & \texttt{plantgrowthform}  & \texttt{plantgrowthform} provides... \\
   \ghsmurl{github.com/...} & \texttt{fungaltraits} & \texttt{fungaltraits} (in progress) provides... \\
-  \ghsmurl{github.com/...} & \texttt{13C} & \texttt{13C} (in progress) provides .... \\
   \hline

### 2017-08-03 ms.tex
beginning of a discussion

@@
 
-Topics to consider:
+Many of the key roadblocks preventing a switch from a static to a dynamic data world, were technological: in the past it took great deal of money and expertise to set up a VDD system.
+The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.
+We argue that by linking together the tools described above, researches can now produce, curate and distribute versioned data much more easily and cheaply than in the past.
 
-\begin{itemize}
-
-\item Addresses important need, leveraging (use) open source platforms.
+%Topics to consider:
+%
+%\begin{itemize}
+%
+%\item Addresses important need, leveraging (use) open source platforms.
 Shows power of open source: by writing single new package, can provide global distribution and access to data, with first class system for managing contents.
-\item Versioning and open access essential for reproducibility
-\item Versioning should enable better benchmarking of models, also because tests can be automated.
-\item We have focussed on particular set of tools, but could do similar things with other platforms or languages.
-....
-\end{itemize}
+%\item Versioning and open access essential for reproducibility
+%\item Versioning should enable better benchmarking of models, also because tests can be automated.
+%\item We have focussed on particular set of tools, but could do similar things with other platforms or languages.
+%....
+%\end{itemize}
 
@@
 \subsection{Conclusion}
-
-Because it builds on existing infra-structure for distributing software,
-this system requires very little (or no) money to set up and only a
-small amount of programming expertise. The current infrastructure will
-not work for big or continually updated data (e.g.~weather stations or
-satellite downloads). For a set of problems that need periodically
-updated small-to-medium sized data (e.g.~most meta-analysis problems),
-this might be a useful approach.
 
@@
 \centering
-%\includegraphics[width=\linewidth]{figure.pdf}
+\includegraphics[width=\linewidth]{figures/Figure1.pdf}
 \caption{Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.

### 2017-08-03 ms.tex
more writing

@@
 
-\subsection{Allowing for one or more data contributers}
+\subsection{Potential for collaboration: allowing for one or more data contributors}
 
@@
 \section{Discussion}
+% General problem. An example of solution. Key features
+
+The problem of tracking data through time and versions is likely to be widespread.
+There are many alternate solutions, with associated features and drawbacks.
+The current structure of scientific research is very distributed, with research groups across the world working on the same question.
+Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.
+As such we argue that there is a niche for a easy, cheap, and flexible VDD system.
 
@@
 The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.
-We argue that by linking together the tools described above, researches can now produce, curate and distribute versioned data much more easily and cheaply than in the past.
+We argue that by linking together the tools described above, researchers can now produce, curate and distribute versioned data much more easily and cheaply than in the past.
+
+The key features of a VDD system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized DOIs.
+Moreover, it is relatively easy to update and add new versions.
+Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.
+But nonetheless, workable systems are already possible.
 

### 2017-08-03 ms.tex
Figure illustrating technologies and users
For #11
Minor edits and more references

@@
 
-Data is now a first-class scientific product.
-Funding bodies, publishers, and scientific social norms are increasingly both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses.
-This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators [Cite some Open Science papers, both pro and con].
+Data is now a first-class scientific product, as is evidenced by the advent of this journal \cite{Editorial-2014}.
+Increasingly, funding bodies, publishers, and scientific social norms are both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
+This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013}.
 One thing that is clear is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous.
-On one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}).
-On the other end are repositories for storing a wide variety of static, stand-alone datasets (\texttt{Dryad}, \texttt{Figshare}, and less optimally, as supplements to research papers)(Table \ref{tab:sql_v_versioneddata}).
+At one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}).
+At the other end are repositories for storing a wide variety of static, stand-alone datasets (eg.
+\texttt{Dryad}, \texttt{Figshare}, and \texttt{Zenodo}; or less optimally, as supplements to research papers)(Table \ref{tab:sql_v_versioneddata}).
 We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.
 
-In this article we i) introduce the concept of Versioned Data Delivery (VDD) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be leveraged to help researchers maintain, distribute, and access small to medium sized datasets; and iii) introduce a new R package called \texttt{datastorr}, an implementation of a VDD system.
+In this article we i) introduce the concept of Versioned Data Delivery (VDD) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new R package called \texttt{datastorr}, an implementation of a VDD system.
 While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
-Furthermore, our backgrounds in the life sciences (and ecology and evolution, in particular); however, we argue that the issues we identify and the solutions we put forward are likely to be relevant to a wide variety of domains, across the natural and social sciences.
+While the examples cited here are in ecology and evolution; the issues identified and the solutions put forward are relevant to nearly all data-oriented scientific domains.
 
-\subsection{Why versioning of data is needed}
+\subsection{Why versioning of data is needed and difficult}
 
@@
 
-\subsection{Why versioning of data has been difficult}
-
-While the benefits of versioned data are fairly clear, determining how best to implement it has been anything but.
+While the benefits of versioned data are fairly clear, determining how to implement it has been contentious .
 The static model of data distribution widely used in science, where datasets are archived either as supplementary materials for a journal article
@@
 research group contributes most of the data). In most of these cases, the data
-collected will not be ``big'' but rather small to medium sized. Such data
+collected will not be ``big'' but rather small-to-medium sized. Such data
 support research projects, but are not in most cases general enough to
-establish infra-structure for. But crucially the data themselves will change
+establish bespoke infrastructure for. But crucially the data themselves will change
 over time. And currently there is no infrastructure for organizing and
@@
 
-In brief, the workflow we present, which we call Versioned Data Delivery (VDD), borrows best-practices for software development and applies them to the challenge of maintaining and distributing versioned data.
+In brief, the workflow we present, which we call Versioned Data Delivery (VDD), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
 Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local  computers.
@@
 The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
-Importantly, VDD uses well-established and tools, which ensures high-level performance and stability.
+Importantly, VDD uses well-established tools, ensuring high-level performance and stability.
 
@@
 Essentially version control tracks line-by-line changes in text files and creates and maintains a history of those changes.
-Increasingly version control has been applied to scientific code and data management, especially for small to medium sized datasets \citep{Ram-2013, Perkel-2016}.
+Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}.
 Git is attractive for data management because it tracks all changes in monitored files.
-This change history is visible to anyone interacting with the repository.
+This history is visible to anyone interacting with the repository.
 It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.
 
-Because \texttt{git} was built from the ground up for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data.
+Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.
 For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
@@
 
-To fully realize the benefits of versioned data, the versioning system needs to be communicated to the user.
+To fully realize the benefits of a versioned data, the VDD system needs to be communicated to the user.
 Crucially a user should be immediately aware of the relationship among versions.
-Software development has dealt with a very similar problem, and we suggest there is some benefit in adopting the notation from that field.
+Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.
 Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution.
@@
 Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction.
-A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise[] maintaining the same dataset structure).
+A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure).
 A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.
@@
 At current there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.
-The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements})).
+The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).
 GitHub has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).
-Those "releases" can be downloaded directly by users via point-and-click.
+Those releases can be downloaded directly by users via point-and-click, or accessed via the Github API.
 Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.
@@
 
-For many use cases, a point-and-click workflow may create barriers for reproducibility.
+A point-and-click workflow may create barriers for reproducibility \cite{Wilson-2014,Lowndes-2017}.
 Ideally, users should be able to also able access the all versions of the database programmatically.
-To this end, we offer a novel solution.
-We have written a R package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the GitHub API to pull down versions of the dataset; and b) constructs the shell of a second, database specific R package, which can be used to access the dataset.
+To this end, we offer a novel implementation of a VDD system.
+We have written a R package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the GitHub API to pull down versions of the dataset; and b) constructs the shell of a second, database specific R package, which can be used to access versions of the dataset.
 Using this system, a researcher can create and distribute a R package that facilitates access to her data with (very) minimal computational skills.
@@
 %
-%\item Addresses important need, leveraging (use) open source platforms.
+%\item Addresses important need, using open source platforms.
 Shows power of open source: by writing single new package, can provide global distribution and access to data, with first class system for managing contents.
@@
 \centering
-%\includegraphics[width=\linewidth]{figure.pdf}
-\caption{Overview of the technologies involved in distributing Lightweight versioned data.}
+\includegraphics[width=\linewidth]{figures/Figure2.pdf}
+\caption{Overview of the technologies involved in distributing versioned data via datastorr.}
 \label{fig:technology_stack}

### 2017-08-03 ms.tex
merging

@@
 For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
-There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future (REF).
-This future system would be faster computationally, smaller in file size, and thus allow for larger datasets to be versioned efficiently.
-That said, the workflow we describe here currently works well with \texttt{git} and should work more smoothly with its future replacement.
+There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future  \citep{Fli, Dat}.
+This future system would be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently.
+That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.
 
@@
 
-TODO: turn this into a paragraph
-
 \textbf{Pull requests}: Much of the activity on GitHub is on open-source projects with many contributors.
 An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
+
+\subsection{Cloud platforms for storing and distributing version controlled data}
+
+There are many current platforms for storing and distributing versioned controlled data for low or no cost.
+The website GitHub (\smurl{www.github.com}) is a commercial platform for hosting and interacting with git repositories.
+In practice, each dataset should be a separate git repository on github (see Table \ref{tab:examples} for examples).
+Although mainly used for computer code, Github is now also being use to manage scientific data \citep{Perkel-2016}.
+Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
 
@@
 Here we describe the particular integration of github and zenodo.
+It is useful to have the ability to either be specific---reference a parti
 
@@
 A tutorial explaining precisely how to set this up is available at \smurl{github.com/richfitz/datastorr}.
-% is this t
+% is this the right site?
+ropensci?
 
@@
 Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.
-But nonetheless, workable systems are already possible.
+But, nonetheless, workable systems are already possible.
 

### 2017-08-03 ms.tex
more writing

@@
 
+TODO turn this into a paragraph about collaboration
+
 \textbf{Pull requests}: Much of the activity on GitHub is on open-source projects with many contributors.
@@
 
-\subsection{Digital object identifiers}
-
-Tools for creating and archiving digital objects have seen rapid progress in recent years.
-At present, each version of the dataset can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
-Here we describe the particular integration of github and zenodo.
-It is useful to have the ability to either be specific---reference a parti
-
-TODO: FLESH THIS OUT AND ADD REFERENCE TO BLOG POST AND OTHER THINKING ON THIS
-
 \subsection{Distributing data to different user needs}
@@
 Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.
-Releases can also be linked to code-based users via an R package which we describe in the following section.
+Releases can also be linked to digital object identifiers and code-based users via for example an R package.
 
@@
 Ideally, users should be able to also able access the all versions of the database programmatically.
-To this end, we offer a novel implementation of a VDD system.
+Code to access the Github releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.
+To make access via code as easy as possible for users, we offer a novel implementation of a VDD system.
 We have written a R package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the GitHub API to pull down versions of the dataset; and b) constructs the shell of a second, database specific R package, which can be used to access versions of the dataset.
@@
 Indeed, we would love to see implementations in \texttt{Python} or other languages.
+
+\subsection{Digital object identifiers}
+
+Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
+At present, each version of the dataset can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
+Here we describe the particular integration of github and zenodo.
+It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project.
+Zenodo currently automatically supplies both ``concept'' and ``version'' DOI's to the commits associated with Github releases automatically \citep[Figure \ref{fig:semantic} and ][]{Nielsen-2017}.
+This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.
+There is an unresolved issue about properly allocating credit to authors for these citations, and we hope that the large citation mapping services (Web of Science and Google Scholar) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
 

### 2017-08-03 ms.tex
Many small edits

@@
 Increasingly, funding bodies, publishers, and scientific social norms are both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
-This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013}.
+This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
 One thing that is clear is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous.
@@
 
-In this article we i) introduce the concept of Versioned Data Delivery (VDD) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new R package called \texttt{datastorr}, an implementation of a VDD system.
+In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, an implementation of a \textsc{vdd} system.
 While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
@@
 The static model of data distribution widely used in science, where datasets are archived either as supplementary materials for a journal article
-or under its own DOI, is simply not designed to provide streams of versioned
+or under its own Digital Object Identifier (\textsc{doi}), is simply not designed to provide streams of versioned
 data (Table \ref{tab:sql_v_versioneddata}). A particular issue with the static
@@
 % big data: there are a number of high-setup-cost options (stand-alone
-% websites, APIs), which have advantages if you have the resources to
+% websites, \textsc{api}s), which have advantages if you have the resources to
 % develop and maintain them. We wanted a relatively low cost way (both in
@@
 
-In brief, the workflow we present, which we call Versioned Data Delivery (VDD), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
+In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
 Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local  computers.
-With a VDD system, data developers maintain a core set of files, which produce an organised dataset that can be similarly be ``installed'' on a user's local computer.
+With a \textsc{vdd} system, data developers maintain a core set of files, which produce an organised dataset that can be similarly be ``installed'' on a user's local computer.
 In either the development of software or data, successive versions are released over time (called ``releases'').
 The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
-Importantly, VDD uses well-established tools, ensuring high-level performance and stability.
+Importantly, our \textsc{vdd} systems uses well-established tools, ensuring high-level performance and stability.
 
@@
 There are many current platforms for storing and distributing versioned controlled data for low or no cost.
-The website GitHub (\smurl{www.github.com}) is a commercial platform for hosting and interacting with git repositories.
-In practice, each dataset should be a separate git repository on github (see Table \ref{tab:examples} for examples).
-Although mainly used for computer code, Github is now also being use to manage scientific data \citep{Perkel-2016}.
+The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with git repositories.
+In practice, each dataset should be a separate git repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples).
+Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.
 Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
@@
 
-\textbf{Pull requests}: Much of the activity on GitHub is on open-source projects with many contributors.
+\textbf{Pull requests}: Much of the activity on \texttt{GitHub} is on open-source projects with many contributors.
 An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
@@
 
-To fully realize the benefits of a versioned data, the VDD system needs to be communicated to the user.
+To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be communicated to the user.
 Crucially a user should be immediately aware of the relationship among versions.
@@
 The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).
-GitHub has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).
-Those releases can be downloaded directly by users via point-and-click, or accessed via the Github API.
+\texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).
+Those releases can be downloaded directly by users via point-and-click, or accessed via the \texttt{GitHub} \textsc{api}.
 Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.
-Releases can also be linked to digital object identifiers and code-based users via for example an R package.
+Releases can also be linked to digital object identifiers and code-based users via for example an \textsc{R} package.
 
-\subsection{\texttt{datastorr} and dataset-specific R packages}
+\subsection{\texttt{datastorr} and dataset-specific \texttt{R} packages}
 
@@
 Ideally, users should be able to also able access the all versions of the database programmatically.
-Code to access the Github releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.
-To make access via code as easy as possible for users, we offer a novel implementation of a VDD system.
-We have written a R package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the GitHub API to pull down versions of the dataset; and b) constructs the shell of a second, database specific R package, which can be used to access versions of the dataset.
-Using this system, a researcher can create and distribute a R package that facilitates access to her data with (very) minimal computational skills.
+Code to access the \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.
+To make access via code as easy as possible for users, we offer a novel implementation of a \textsc{vdd} system.
+We have written a \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset.
+Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to her data with (very) minimal computational skills.
+We have focussed on the \texttt{R} platform \cite{R-2017} as one of the most prominent platforms for data analysis.
 
-For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/dfalster/baad.data}), which is an interface to the Biomass and Allometry Database (BAAD) stored at \smurl{github.com/dfalster/baad}.
+For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
 \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}.
-And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's harddrive.
+And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
 When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline).
-If a researcher requires an older version of BAAD, perhaps to reproduce an analysis, specifying the version number of the data as an argument in \texttt{baad.data::baad\_data()} will retrieve and store this, without overwriting any other dataset versions they may have previously downloaded.
+If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g.
+\texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded.
+This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.
 
-Using \texttt{datastorr}, researchers can set up their own versioned database (and R package) simply by providing: 1) a github repository name (e.g., ``traitecoevo/taxonlookup''); 2) a single file that will contain the versioned data (e.g., ``taxonlookup.csv''); 3) and the function used to load this into \texttt{R} (e.g., \texttt{read.csv()}).
-A tutorial explaining precisely how to set this up is available at \smurl{github.com/richfitz/datastorr}.
-% is this the right site?
-ropensci?
+Using \texttt{datastorr}, researchers can set up their own versioned database (and R package) simply by providing: 1) a \texttt{GitHub} repository name (e.g., ``traitecoevo/taxonlookup''); 2) a single file that will contain the versioned data (e.g., ``taxonlookup.csv''); 3) and the function used to load this into \texttt{R} (e.g., \texttt{read.csv()}).
+A tutorial explaining precisely how to set this up is available at \smurl{github.com/ropenscilabs/datastorr}.
 
-Then as the data changes (and hopefully improves) with time, update the data
-file and create a github release with a new version number (preferably
+Then as the dataset grows over time, researchers update the git repo and create a \texttt{GitHub} release with a new version number (preferably
 using \href{http://semver.org/}{semantic versioning}). All the versions
@@
 
-We have created \texttt{datastorr} to make VDD easy to implement and manage for \texttt{R} users.
+We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users.
 And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
-However, we can imagine alternative methods for interacting with GitHub and building templates and adopting a VDD system does not critically depend on using \texttt{datastorr}.
-Indeed, we would love to see implementations in \texttt{Python} or other languages.
+However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system does not critically depend on using \texttt{datastorr} or \texttt{GitHub}.
+Indeed, we would love to see implementations in \texttt{Python} or other languages and using other cloud platforms for data distribution.
 
@@
 Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
-At present, each version of the dataset can be automatically archived and assigned a DOI, via several providers (Table \ref{tab:doi_minting}).
-Here we describe the particular integration of github and zenodo.
+At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}).
+Here we describe the particular integration of \texttt{GitHub} and \texttt{Zenodo}.
 It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project.
-Zenodo currently automatically supplies both ``concept'' and ``version'' DOI's to the commits associated with Github releases automatically \citep[Figure \ref{fig:semantic} and ][]{Nielsen-2017}.
+\texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Figure \ref{fig:semantic} and ][]{Nielsen-2017}.
 This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.
@@
 Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.
-As such we argue that there is a niche for a easy, cheap, and flexible VDD system.
+As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.
 
-Many of the key roadblocks preventing a switch from a static to a dynamic data world, were technological: in the past it took great deal of money and expertise to set up a VDD system.
+Many of the key roadblocks preventing a switch from a static to a dynamic data world, were technological: in the past it took great deal of money and expertise to set up a \textsc{vdd} system.
 The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.
@@
 
-The key features of a VDD system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized DOIs.
+The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s.
 Moreover, it is relatively easy to update and add new versions.
@@
 %In our system, the data is downloaded only once per computer, available
-%for offline access, easy to port to other platforms, and can get a doi
+%for offline access, easy to port to other platforms, and can get a \textsc{doi}
 %when desired (i.e.~when it's used for a publication). All old versions
@@
   \hline
-  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web database} & \textbf{Lightweight versioned data}\\
+  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web database} & \textbf{Versioned data delivery}\\
+  \textbf{} & (e.g. \smurl{datadryad.org})& (e.g. \smurl{coraltraits.org}) & (eg. \texttt{datastorr} via \texttt{GitHub})\\
   \hline
-   Webserver        & e.g. dataDryad & custom                          &  Maintained by github.com\\
-   Backend          & none & SQL + Ruby-on-rails 			& git + datastorr \\
-   User access      & Web browser & Web browser 				    & R \\
-   Ease of setup    & Very easy & Hard 							& Easy\\ % server, backups, bandwidth
+   Webserver        & bespoke & bespoke &  \smurl{github.com}\\
+   Backend          & none & SQL + Ruby-on-rails 			& git + \texttt{datastorr} \\
+   User access      & Web browser & Web browser 				    & Web browser or R \\
+   Ease of setup    & Very easy & Hard 							& Easy\\
    Data size        & Up to several Gb & Small-Very large 				& Up to 1Gb\\
    Cost             & Varies & Varies  						& Free \\
-   Bandwidth        &managed by dataDryad & Pro rata 						& Managed by github.com\\
-   Maintainer skills &none & Ruby + php 					& R + git \\
+   Bandwidth        & Managed by provider & Pro rata 						& Managed by \texttt{GitHub}\\
+   Maintainer skills & None & Ruby + php 					& R + git \\
    User skills      &Web browsing& Web browsing  					& basic R \\
    Versioning       &None& Hard 							& Easy \\
-   DOI minting      &Automatic & Many steps 					& Automatic \\
+   \textsc{doi} minting      &Automatic & Manual 					& Automatic \\
   \hline 
@@
 \centering
-\caption{Technologies used to maintain, store and distribute "Lightweight versioned data".}
+\caption{Glossary of technologies used to maintain, store and distribute the Versioned Data Delivery system decsribed in this paper.}
 {\footnotesize
@@
   \textbf{Technology} & \textbf{Description} \\\hline
-   git & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data\\
-   github.com & A commercial webplatform for sharing, visualising and managing `git' repositories. Includes issue tracking\\
-   R     &  Open source statsitcial and data processing language \\
-   CRAN  &  Open source repository of packages for the R language \\
-   datastorr & R package used to fetch versioned releases from github  \\
-   DOI & Digital object identifier, which refers a user to a single stable digital object \\ 
-   DOI ``minting'' & creation of a DOI's \\
+   \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data\\
+   github.com & A commercial webplatform for sharing, visualising and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
+  \textsc{API}   & An Application Programming Interface provides a set of protocols for exchanging information.\\
+   \texttt{R}     &  Open source statistical and data processing language \\
+   \textsc{CRAN}  &  Open source repository of packages for the R language \\
+   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}  \\
+   \textsc{doi} & Digital object identifier, which refers a user to a single stable digital object \\
+   \textsc{doi} ``minting'' & Allocation of a \textsc{doi} to a data product\\
    \hline
@@
 \centering
-\caption{Groups of users and their requirements}
+\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements}
 {\footnotesize
@@
     & & Low on-going maintenance \\
-  Contributor & Contribute to future versions & Add new data \\
+  Contributor & Contribute to future versions of a dataset & Add new data \\
     & & Report errors in existing data  \\
-  Users (all) & Easy access to releases & Introduction \& overview \\
+  Users (all) & Easy access to releases from a versioned dataset & Introduction \& overview \\
     & & Long term stability \\
     & & Clear path for users to become contributors \\
-  Users (computational, machines) & Build reproducible products on these data & Scripted access to releases\\
-    & &Easy installation\\
+  Users (code-based) & Build reproducible products using specific versions of a dataset & Scripted access to releases\\
+    & & Easy installation\\
     & & Long term stability \\
@@
 \centering
-\caption{Alternative platforms for minting DOIs from github repositories}
+\caption{Alternative platforms for minting \textsc{doi}s from \texttt{GitHub} repositories.}
 {\footnotesize
@@
   \textbf{Provider} & \textbf{Features} \\ \hline
-  Zenodo: \smurl{zenodo.org} & Automated DOI for each github release \\
+  \texttt{Zenodo}: \smurl{zenodo.org} & Automated \textsc{doi} for each \texttt{GitHub} release \\
     & Funded and maintained by CERN and OpenAIRE for long term preservation \\
-    & Only includes files embedded in git repo, not built  by code\\
-  Figshare: \smurl{figshare.com} & Builds data package based on github release \\
+    & At present, can only archive files embedded directly in the git repo, not produced by it as outputs\\
+  \texttt{FigShare}: \smurl{figshare.com} & Builds data package based on \texttt{GitHub} release \\
     & Allows you to upload additional files, e.g. datasets built by code \\
-    & Integrated with several journals, e.g. Nature family, PNAS \\
-    & Some limited potential for versioned DOIs\\
+    & Integrated with several journals, e.g. Nature family, \textsc{PNAS} \\
   \hline
@@
 \centering
-\caption{Example datasets using the lightweight versioned data workflow described in this paper.}
+\caption{Example datasets using the Versioned Data Delivery system described in this paper.}
 {\footnotesize
@@
   \hline
-   \textbf{Github repo} & \textbf{R package} & \textbf{Description} \\ \hline
-  \ghsmurl{dfalster/baad/} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} \citep{Falster-2015} provides...\\
-  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & \texttt{taxonlookup} \citep{Pennell-2015a} provides...\\
-  \ghsmurl{github.com/...} & \texttt{fungaltraits} & \texttt{fungaltraits} (in progress) provides... \\
+   \textbf{\texttt{GitHub} repo} & \textbf{R package} & \textbf{Description} \\ \hline
+  \ghsmurl{dfalster/baad/} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \citep{Falster-2015}.\\
+  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & \texttt{taxonlookup} provides a taxonomic lookup table for land plants \citep{Pennell-2015a}.\\
   \hline
@@
 \includegraphics[width=\linewidth]{figures/Figure2.pdf}
-\caption{Overview of the technologies involved in distributing versioned data via datastorr.}
+\caption{Overview of the different users and technologies involved in distributing versioned data via \texttt{datastorr}.}
 \label{fig:technology_stack}

### 2017-08-03 ms.tex
reworked summary

@@
 
-Sharing data has become a key part of reproducible workflows in science.
+The sharing and re-use of data has become a cornerstone of modern science.
+To date, the curation and distribution of data has largely been conducted under one of two frameworks: data is either deposited in large, collaborative databases (often run by institutions and staffed by computational professionals) or else published as static files alongside a specific publication or as a standalone product released by a research group.
+For most medium-sized datasets, the former is clearly overkill and likely unwieldly and unsustainable and the latter, while convenient, is rather limited as it does not allow the dataset to improve nor to grow over time.
+This suggests is a substantial conceptual and technological gap in how we curate and share data.
+In this article, we argue that adopting the paradigm of versioned data, analogous to software versioning, can help fill this gap.
+We introduce a system called Versioned Data Delivery (VDD) and present novel computational tools to make creating, archiving, and distributing versioned data easily, quickly and cheaply.
+These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.
+
+%Sharing data has become a key part of reproducible workflows in science.
 To help this process, numerous linked tools have been developed to allow quick and easy data sharing.

### 2017-08-03 ms.tex
some text cleanup

@@
 
-Data is now a first-class scientific product, as is evidenced by the advent of this journal \cite{Editorial-2014}.
+Data is now a first-class scientific product, as is evidenced by the advent of this journal (\cite{Editorial-2014}).
 Increasingly, funding bodies, publishers, and scientific social norms are both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
 This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
-One thing that is clear is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous.
+One thing that is clear, however, is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous.
 At one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}).
 At the other end are repositories for storing a wide variety of static, stand-alone datasets (eg.
-\texttt{Dryad}, \texttt{Figshare}, and \texttt{Zenodo}; or less optimally, as supplements to research papers)(Table \ref{tab:sql_v_versioneddata}).
+\texttt{Dryad}, \texttt{Figshare}, and \texttt{Zenodo}; or less optimally, as supplements to research papers) (Table \ref{tab:sql_v_versioneddata}).
 We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.
@@
 While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
-While the examples cited here are in ecology and evolution; the issues identified and the solutions put forward are relevant to nearly all data-oriented scientific domains.
+Likewise, the examples cited here come from our own research areas, ecology and evolutionary biology, but we argue that the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.
 
@@
 accessible--the up-to-date data can be distributed, used, and analyses
-based on any of the released versions data can be reproduced.
+based on any of the previously released versions data can be reproduced.
 
@@
 Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}.
-Git is attractive for data management because it tracks all changes in monitored files.
+\texttt{git} is attractive for data management because it tracks all changes in monitored files.
 This history is visible to anyone interacting with the repository.
@@
 An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
-
-\subsection{Cloud platforms for storing and distributing version controlled data}
-
-There are many current platforms for storing and distributing versioned controlled data for low or no cost.
-The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with git repositories.
-In practice, each dataset should be a separate git repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples).
-Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.
-Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
 
@@
 
-\subsection{Conclusion}
+%\subsection{Conclusion}
 

### 2017-08-04 ms.tex
added some keywords. generally i think these are useless but apparently we still have to include them
Para on collaboration, tidying throughout

@@
 \runninghead{Versioned Data Delivery}
-\keywords{}
+\keywords{Data sharing, Version control, Semantic versioning, Meta-analysis}
 
@@
 To date, the curation and distribution of data has largely been conducted under one of two frameworks: data is either deposited in large, collaborative databases (often run by institutions and staffed by computational professionals) or else published as static files alongside a specific publication or as a standalone product released by a research group.
-For most medium-sized datasets, the former is clearly overkill and likely unwieldly and unsustainable and the latter, while convenient, is rather limited as it does not allow the dataset to improve nor to grow over time.
-This suggests is a substantial conceptual and technological gap in how we curate and share data.
-In this article, we argue that adopting the paradigm of versioned data, analogous to software versioning, can help fill this gap.
-We introduce a system called Versioned Data Delivery (VDD) and present novel computational tools to make creating, archiving, and distributing versioned data easily, quickly and cheaply.
-These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.
-
-%Sharing data has become a key part of reproducible workflows in science.
-To help this process, numerous linked tools have been developed to allow quick and easy data sharing.
-Thus far, however, data has been largely considered a fixed and unchanging product---the particular data file for a particular analysis can be quickly unloaded, given a digital identifier, and shared.
-While this is highly useful, this framework does not account for on-going scientific progress: for many important problems, epitomised by problems that lead to meta-analysis, datasets will continue to grow with time---more data points will be added, and new and more useful data structures will be created.
-Essentially this means that data, like science in general, progress with time.
-We suggest that a versioning system for data will facilitate this process.
-Moreover, while the idea of versioned data has been taken up by large data sharing organizations, we present new tools for creating, archiving, and distributing versioned data easily, quickly and cheaply.
+For most medium-sized datasets, the former is clearly overkill and likely unwieldy and unsustainable, and the latter -- while convenient -- is rather limited as it does not allow the dataset to improve nor to grow over time.
+This indicates a substantial conceptual and technological gap in how we curate and share data.
+In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, can help fill this gap.
+We introduce a system called Versioned Data Delivery (VDD) and present novel computational tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
 These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.
@@
 To overcome the limitations of static datasets, many research groups have
-setup dedicated webservers to host and distribute data. In these systems,
+setup dedicated webservers to host and distribute data (Table 1)\citep[e.g.][]{Madin-2016}. In these systems,
 datasets areas maintained in a structure database, such as SQL, and a front
@@
 
-% For small datasets, it is simply not possible to use the same tools developed for
-% big data: there are a number of high-setup-cost options (stand-alone
-% websites, \textsc{api}s), which have advantages if you have the resources to
-% develop and maintain them. We wanted a relatively low cost way (both in
-% money and time).
-
 \section{Results}
 
-\subsection{A lightweight, cheap and scalable workflow for versioned data}
-
-% DF: Bring out general features more.
-Right now git delivers this.
-With the growth of open-source tools and collaborative platforms, we believe all the features needed for delivering the vision outlined above are now present.
-The workflow we describe below builds off existing platforms and combines them in new ways, and offers the potential for widespread uptake at little or no cost.
+\subsection{A lightweight, cheap, and scalable workflow for versioned data}
 
@@
 There are many current platforms for storing and distributing versioned controlled data for low or no cost.
-The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with git repositories.
-In practice, each dataset should be a separate git repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples).
+The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories.
+In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples).
 Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.
@@
 
-\subsection{Potential for collaboration: allowing for one or more data contributors}
+\subsection{Potential for collaboration: allowing for multiple data contributors}
 
-TODO turn this into a paragraph about collaboration
-
-\textbf{Pull requests}: Much of the activity on \texttt{GitHub} is on open-source projects with many contributors.
-An essential feature that of sending a pull request, whereby a would-be contributor submits prospective changes to be incorporated.
+One of the greatest benefits of using \texttt{git} for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
+This is enabled via the distributed and decentralised nature of \texttt{git} repositories.
+Because it tracks changes line-by-line, multiple users can make changes to different parts of the code (data) and the \texttt{git} system will seamlessly integrate these together (if that is possible), or, when needed, flag where there are conflicts that need to be resolved.
+This makes collaboration easy.
+For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project.
+This process is called submitting a ``Pull request''.
+One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
 
@@
 
-Then as the dataset grows over time, researchers update the git repo and create a \texttt{GitHub} release with a new version number (preferably
+Then as the dataset grows over time, researchers update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number (preferably
 using \href{http://semver.org/}{semantic versioning}). All the versions
@@
 This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.
-There is an unresolved issue about properly allocating credit to authors for these citations, and we hope that the large citation mapping services (Web of Science and Google Scholar) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
+There is an unresolved issue about properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
 
@@
 The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.
-We argue that by linking together the tools described above, researchers can now produce, curate and distribute versioned data much more easily and cheaply than in the past.
+We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.
 
@@
 But, nonetheless, workable systems are already possible.
-
-%Topics to consider:
-%
-%\begin{itemize}
-%
-%\item Addresses important need, using open source platforms.
-Shows power of open source: by writing single new package, can provide global distribution and access to data, with first class system for managing contents.
-%\item Versioning and open access essential for reproducibility
-%\item Versioning should enable better benchmarking of models, also because tests can be automated.
-%\item We have focussed on particular set of tools, but could do similar things with other platforms or languages.
-%....
-%\end{itemize}
-
-%(old content unchanged)
-%
-%One of the key problems in switching from a static to a dynamic data
-%world, is that it's currently hard to set up and maintain a system like
-%this. Ideally using a versioned database should be easy for both the
-%user and the data maintainer. This is especially crucial is the data
-%maintainers are busy. Moreover, old versions need to be preserved in a
-%stable and accessible way.
-%
-%To try and achieve to above goals, we built a system to distribute
-%versions of our datasets. We think that with this infrastructure, it is
-%relatively easy to set up and maintain a versioned system with any
-%small-to-medium sized dataset.
-%
-%In our system, the data is downloaded only once per computer, available
-%for offline access, easy to port to other platforms, and can get a \textsc{doi}
-%when desired (i.e.~when it's used for a publication). All old versions
-%are accessible.
-%
-%
-%
-%\subsection{How much expertise is needed to setup and use?}
-%
-%To use a versioned dataset: Very very basic \texttt{R}. With 4 lines of \texttt{R} code, you can
-%have the data downloaded and loaded into R and ready for use.
-%
-%To set a system like this up: Basically there are two general skills required: 1) understanding
-%version control via \texttt{git} and 2) writing a very simple R package.
-%From the prolific Hadley Wickham there is a
-%\href{http://r-pkgs.had.co.nz/git.html}{very clear introduction to git}
-%and \href{http://r-pkgs.had.co.nz/intro.html}{an instruction to the
-%basics of R packages}.
-
-%\subsection{Conclusion}
 
@@
    Webserver        & bespoke & bespoke &  \smurl{github.com}\\
-   Backend          & none & SQL + Ruby-on-rails 			& git + \texttt{datastorr} \\
+   Backend          & none & SQL + Ruby-on-rails 			& \texttt{git} + \texttt{datastorr} \\
    User access      & Web browser & Web browser 				    & Web browser or R \\
@@
    Bandwidth        & Managed by provider & Pro rata 						& Managed by \texttt{GitHub}\\
-   Maintainer skills & None & Ruby + php 					& R + git \\
+   Maintainer skills & None & Ruby + php 					& R + \texttt{git} \\
    User skills      &Web browsing& Web browsing  					& basic R \\
@@
 \centering
-\caption{Glossary of technologies used to maintain, store and distribute the Versioned Data Delivery system decsribed in this paper.}
+\caption{Glossary of technologies used to maintain, store, and distribute the Versioned Data Delivery system described in this paper.}
 {\footnotesize
@@
    \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data\\
-   github.com & A commercial webplatform for sharing, visualising and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
+   github.com & A commercial webplatform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
   \textsc{API}   & An Application Programming Interface provides a set of protocols for exchanging information.\\
@@
     & Funded and maintained by CERN and OpenAIRE for long term preservation \\
-    & At present, can only archive files embedded directly in the git repo, not produced by it as outputs\\
+    & At present, can only archive files embedded directly in the \texttt{git} repo, not produced by it as outputs\\
   \texttt{FigShare}: \smurl{figshare.com} & Builds data package based on \texttt{GitHub} release \\
@@
   \hline
-   \textbf{\texttt{GitHub} repo} & \textbf{R package} & \textbf{Description} \\ \hline
+   \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
   \ghsmurl{dfalster/baad/} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \citep{Falster-2015}.\\
@@
 \includegraphics[width=\linewidth]{figures/Figure2.pdf}
-\caption{Overview of the different users and technologies involved in distributing versioned data via \texttt{datastorr}.}
+\caption{Overview of the different users and technologies involved in distributing a versioned dataset via \texttt{datastorr}.}
 \label{fig:technology_stack}
 \end{figure}
+
 \bibliography{refs}

### 2017-09-21 ms.tex

@@
 
+% I think that the real issue with the first approach is that it's not
+% straightforward for a small group to set up; that is probably packed
+% away in "unsustainable" but could be repgrased to make it clearer
+% perhaps?  Breaking down barriers is always a good aim
+%
+% I really think that something like CKAN (though it doesn't support
+% holding data if I remember correctly) could totally be a complete
+% solution to the sorts of issues that we have here.
+%
+% The "changing over time" is introduced as a problem associated with
+% medium sized data but I think it comes from nowhere in the current
+% draft.  It's more a separate concern than the big/small/medium size
+% angle.
 The sharing and re-use of data has become a cornerstone of modern science.
 To date, the curation and distribution of data has largely been conducted under one of two frameworks: data is either deposited in large, collaborative databases (often run by institutions and staffed by computational professionals) or else published as static files alongside a specific publication or as a standalone product released by a research group.
-For most medium-sized datasets, the former is clearly overkill and likely unwieldy and unsustainable, and the latter -- while convenient -- is rather limited as it does not allow the dataset to improve nor to grow over time.
+For most medium-sized datasets, the former is overkill and likely unwieldy and unsustainable, and the latter -- while convenient -- is rather limited as it does not allow the dataset to improve nor to grow over time.
 This indicates a substantial conceptual and technological gap in how we curate and share data.
@@
 This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
-One thing that is clear, however, is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous.
+We feel that there is currently no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous (indeed a single system may be impossible).
 At one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}).
@@
 We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.
+% Again, I think that the versioning issue arises here out of nowhere.
+% It's a point that I think deserves further elaboration on its own;
+% it cuts across the other two types of data too.  But because genbank
+% and publication-related-archive are immutable data types this is not
+% an issue with much of the existing data.  But people are *always*
+% emailing around the most recent copy of data.  We have data on
+% immunisation rates (WUENIC) where the data are an excel file that is
+% updated periodically upstream with the same name (and you never know
+% when it's been updated or what has changed).  So this is worth
+% drawing out into its own paramgraph I think.
 
-In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, an implementation of a \textsc{vdd} system.
+In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
 While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
@@
 based on any of the previously released versions data can be reproduced.
+% Drawing this out; we might want to highlight that databases like
+% genbank can be seen as versioned (though someone who knows how they
+% work should check the details).  New entries are added but are
+% immutable.  This means the database gets better through time (more
+% entries) but each entry is fixed at creation.  Not all databases can
+% pull that off - for the reasons that you mention above and because
+% many ecological databases try to do the equivalent of getting the
+% "best measurement of X" for each species which is not the same as
+% "an observation of X".
 
-While the benefits of versioned data are fairly clear, determining how to implement it has been contentious .
+% I think these need spelling out in a little more detail as I find
+% versioning a fairly hard sell to most people!
+While the benefits of versioned data are fairly clear, determining how to implement it has been contentious.
 The static model of data distribution widely used in science, where datasets are archived either as supplementary materials for a journal article
 or under its own Digital Object Identifier (\textsc{doi}), is simply not designed to provide streams of versioned
-data (Table \ref{tab:sql_v_versioneddata}). A particular issue with the static
+data (Table \ref{tab:sql_v_versioneddata}).
+% I think that this model is actually ok; for data archiving where the
+% data are a record of what was used in the paper, a single immutable
+% data set makes sense.  It just doesn't extend to things where the
+% research output *is* the dataset.  So this is more a case of
+% pointing out that this model does not apply well to all sorts of
+% data - given how successful the archive-at-publication model has
+% been at getting data out of people's drawers I don't think we should
+% complain that it doesn't do something that it does not need to do.
+A particular issue with the static
 model is that data releases are mostly triggered by publication of academic
@@
 between papers. This is particularly the case with the discovery of small
-errors. For any medium-sized dataset that is put to reuse, errors will
+errors.
+% OK, here again; it's not so much medium-sized as *living*.  The
+% issues in practically doing the versioning are more acute when the
+% data are bigger than small, that's all.
+For any living dataset that is put to reuse, errors will
 inevitably be found, and there needs to be some easy  way to correct them and
@@
 setup dedicated webservers to host and distribute data (Table 1)\citep[e.g.][]{Madin-2016}. In these systems,
-datasets areas maintained in a structure database, such as SQL, and a front
-end to this database is written, often using a platform such as Ruby on rails.
+datasets areas maintained in a structured database, such as SQL, and a custom front
+end to this database must be written.
 When designed well, such systems are very useful for updating or adding to a
 dataset. However, the infrastructure investment (designing maintaining, and
-hosting the database) is clearly overkill for most projects and is a very
+hosting the database) is overkill for most projects and is a very
 substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many research groups. Moreover, the existence of a live webserver,
@@
 
+% This paragraph reads like it used to be elsewhere in the document.
+% I would flip it around and talk about small data first.  You could
+% talk about 'git' (referencing Ram-2013) and say "on the face of it,
+% versioning tools like git appear to provide the ideal interface for
+% versioning data".  Then talk about how git is designed for code and
+% breaks down as the data involved exceed a few 10s to hundreds of MB.
+% After that, we can say that the aim is not to support truely "Big"
+% data because research groups capable of generating such data
+% probably have the wherewithall to look after it.  I know you bring
+% it up below, but I think it would go better here...
 When considering the right tool to store and deliver versioned data, the size
@@
 
-So what is needed is simple tool-chain for
+We feel there is a  need for
 making versions of a dataset and delivering it to potential users, along
@@
 
+% Do we still need results sections in a data journal?
 \section{Results}
@@
 
+% I really like the focus on "release".  It's a happy coincidence that
+% it's what github uses in the interface.  I'm increasingly of the
+% opinion that conflating source control and release (via
+% install_github) is not a good thing in the R community, but it is
+% seductive...
 In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
@@
 
+% With suggested edits, some of this moves up.
 Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.
@@
 
+% So, git doesn't actually work this way, though everyone seems to
+% think that it does.  Git stores *content* and then later tries to
+% de-duplicate it.  It then displays changes as line-by-line diffs
+% (but you can change this to display the changes other ways; see
+% --color-words --ignore-all-space for example.  The problem is that
+% the algorithms that git uses are optimised for typical sizes of code
+% and not for data.  Small data is actually totally fine to store with
+% git.  The other issue that turns up is if the files are _binary_
+% then an arbitrarily small change in the underlying data can result
+% in an arbitrarily large change in the binary representation, which
+% prevents deduplication.
+%
+% Git-LFS tries to step around this by storing large files separately
+% from git's general machinery.  Git-annex did the same.  But they
+% suffer from some of the other issues with git which are that it's a
+% bit of a pain for (especially novice) users to do nontrivial things
+% with.  I _always_ have to look up how to compare two versions across
+% branches.  And there's no simple way of loading in two versions of
+% the data into R without ending up with two working copies or doing
+% some nasty "git show" call.
 Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.
@@
 
+% Though this section I am losing track of the big picture here;
+% there's lots of discussion of the positive benefits of "Something"
+% but without knowing what it is, it's unlikely that these will stick.
+% I wonder if we should move the workflow description higher.
+
 \subsection{Potential for collaboration: allowing for multiple data contributors}
@@
 This is enabled via the distributed and decentralised nature of \texttt{git} repositories.
-Because it tracks changes line-by-line, multiple users can make changes to different parts of the code (data) and the \texttt{git} system will seamlessly integrate these together (if that is possible), or, when needed, flag where there are conflicts that need to be resolved.
+Mltiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
+% We have constant drama at work because the different spreadsheet
+% writers we use across the platforms quote files inconsistently so
+% every save is a complete rewrite of the file.  It's totally
+% impossible to do any collaborative editing....
+ or, when needed, flag where there are conflicts that need to be resolved.
 This makes collaboration easy.
@@
 One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
+ % This section conflates git and github more than is useful.  It
+ % might be worth setting up github as the explicit platform here or
+ % dialing back the language a bit.  GitLab and bitbucket have their
+ % own "merge request" feature (I think).  It's also github that makes
+ % repositories _discoverable_ which is a requirement for getting
+ % outside collaborators in.  Oddly the decentralised nature of git is
+ % not actually that useful in the github model because github is so
+ % strongly centralised.  Will and I did some true decentralised git
+ % collaboration on a plane to nescent once with a bare git repo on a
+ % thumb drive.
 
@@
 A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.
+% I think that major version increments could still be looked at as
+% _breaking_ changes as often as large additions of data.  Changing
+% column names, removing columns, changing the structure of linked
+% tables etc.
 
@@
 Those releases can be downloaded directly by users via point-and-click, or accessed via the \texttt{GitHub} \textsc{api}.
+% I have no idea what this next sentence means.
 Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.
@@
 
+% Something not mentioned; that the versions are cached between runs
+% so that the data are accessed by name/version (rather than by
+% homebrewing a local cache and manually doing a bunch of file name
+% mangling) and that all accesses after the first are very fast
+
 We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users.
@@
 
+% Karthik has started with rdrop2 again I think; I do think that is a reasonable low-entry way of distributing things
+
 \subsection{Digital object identifiers}
+
+% I'm unclear on the precice benefits of dois here (and actually in
+% general to a point); some further clarification might be useful.  So
+% far as I know they are useful because (a) journal urls are terrible
+% and inconsistent (b) because everyone has them.  I don't really know
+% why else you'd need one.  Especially in our case we offer no promise
+% of persistence (deleting the github repo would break the doi and we
+% could update the repo to include different data trivially).  So what
+% is being added?  Our retrival system does not use the doi at all!
 
@@
 
+% The introduction talks at length about medium sized data and little about versioning but this section talks only about versioning
+
 The problem of tracking data through time and versions is likely to be widespread.
@@
 
-Many of the key roadblocks preventing a switch from a static to a dynamic data world, were technological: in the past it took great deal of money and expertise to set up a \textsc{vdd} system.
+Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past it took great deal of money and expertise to set up a \textsc{vdd} system.
 The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.
@@
   \hline
+% Latex being the worst has made the overflow in the second column flow through to affect the third.  I used to know how to deal with this...
   \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
@@
   \end{tabular}
-  }
+  % }
 \label{tab:user_requirements}

### 2017-09-21 ms.tex
Suggestions from Dan Noble
Thanks @daniel1noble!

@@
 
-\title{Versioned data: why it’s needed and how it can be achieved (easily and cheaply)}
+\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
 \author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$,\\ Matthew
@@
 For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
+% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”.
+Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach.
+From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate.
+They just don’t feel like its worth doing.
 There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future  \citep{Fli, Dat}.
@@
 This is enabled via the distributed and decentralised nature of \texttt{git} repositories.
-Mltiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
+Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
 % We have constant drama at work because the different spreadsheet
@@
 A point-and-click workflow may create barriers for reproducibility \cite{Wilson-2014,Lowndes-2017}.
-Ideally, users should be able to also able access the all versions of the database programmatically.
+Ideally, users should be able to access all versions of the database programmatically.
 Code to access the \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.
 To make access via code as easy as possible for users, we offer a novel implementation of a \textsc{vdd} system.
-We have written a \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset.
-Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to her data with (very) minimal computational skills.
+We have written an \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset.
+Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to their data with (very) minimal computational skills.
 We have focussed on the \texttt{R} platform \cite{R-2017} as one of the most prominent platforms for data analysis.
@@
 And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
+% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”.
+You may want to discuss this in greater depth as I don’t see how this is possible.
 When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline).
@@
 And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
+% Reword next senetnce
 However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system does not critically depend on using \texttt{datastorr} or \texttt{GitHub}.
-Indeed, we would love to see implementations in \texttt{Python} or other languages and using other cloud platforms for data distribution.
+We would love to see implementations in \texttt{Python} or other languages and using other cloud platforms for data distribution.
 
@@
 
-The problem of tracking data through time and versions is likely to be widespread.
-There are many alternate solutions, with associated features and drawbacks.
+The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.
+% Reword
 The current structure of scientific research is very distributed, with research groups across the world working on the same question.
@@
 \includegraphics[width=\linewidth]{figures/Figure1.pdf}
-\caption{Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.
+\caption{
+% could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
+Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.
 \textbf{a)} 3 numbers indicate type of change.

### 2017-09-21 ms.tex
an alternative abstract to address #21

@@
 We introduce a system called Versioned Data Delivery (VDD) and present novel computational tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
+These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.
+
+% Alternative abstract, to introduce versioning problem earlier, leaving the small versus large issue until later
+
+The sharing and re-use of data has become a cornerstone of modern science.
+To help this process, numerous linked tools have been developed to allow quick and easy data sharing:    a data file for a particular analysis can be quickly uploaded, given a digital identifier, and shared.
+Thus far, however, this framework does not account for on-going scientific progress: for many important problems, epitomised by problems that lead to one or more meta-analyses, datasets will continue to grow with time---more data points will be added, and new and more useful data structures will be created.
+Essentially this means that datasets, like science in general, progress with time.
+We suggest that many different datasets would be more useful if organized as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.
+In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data curators and users of even small-to-medium sized datasets.
+We introduce a system called Versioned Data Delivery (VDD) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
 These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.

### 2017-09-21 ms.tex
Attempt at focussing intro on living data
- remove standalone section on Why versioned data
- towards #22

@@
 
-% I think that the real issue with the first approach is that it's not
-% straightforward for a small group to set up; that is probably packed
-% away in "unsustainable" but could be repgrased to make it clearer
-% perhaps?  Breaking down barriers is always a good aim
-%
-% I really think that something like CKAN (though it doesn't support
-% holding data if I remember correctly) could totally be a complete
-% solution to the sorts of issues that we have here.
-%
-% The "changing over time" is introduced as a problem associated with
-% medium sized data but I think it comes from nowhere in the current
-% draft.  It's more a separate concern than the big/small/medium size
-% angle.
-The sharing and re-use of data has become a cornerstone of modern science.
-To date, the curation and distribution of data has largely been conducted under one of two frameworks: data is either deposited in large, collaborative databases (often run by institutions and staffed by computational professionals) or else published as static files alongside a specific publication or as a standalone product released by a research group.
-For most medium-sized datasets, the former is overkill and likely unwieldy and unsustainable, and the latter -- while convenient -- is rather limited as it does not allow the dataset to improve nor to grow over time.
-This indicates a substantial conceptual and technological gap in how we curate and share data.
-In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, can help fill this gap.
-We introduce a system called Versioned Data Delivery (VDD) and present novel computational tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
-These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.
-
-% Alternative abstract, to introduce versioning problem earlier, leaving the small versus large issue until later
-
 The sharing and re-use of data has become a cornerstone of modern science.
@@
 
-Data is now a first-class scientific product, as is evidenced by the advent of this journal (\cite{Editorial-2014}).
-Increasingly, funding bodies, publishers, and scientific social norms are both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
-This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
-We feel that there is currently no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous (indeed a single system may be impossible).
-At one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}).
-At the other end are repositories for storing a wide variety of static, stand-alone datasets (eg.
-\texttt{Dryad}, \texttt{Figshare}, and \texttt{Zenodo}; or less optimally, as supplements to research papers) (Table \ref{tab:sql_v_versioneddata}).
-We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.
-% Again, I think that the versioning issue arises here out of nowhere.
-% It's a point that I think deserves further elaboration on its own;
-% it cuts across the other two types of data too.  But because genbank
-% and publication-related-archive are immutable data types this is not
-% an issue with much of the existing data.  But people are *always*
-% emailing around the most recent copy of data.  We have data on
-% immunisation rates (WUENIC) where the data are an excel file that is
-% updated periodically upstream with the same name (and you never know
-% when it's been updated or what has changed).  So this is worth
-% drawing out into its own paramgraph I think.
+As is evidenced in the advent of this journal, publication of quality datasets is now considered a first-class scientific product.
+Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013,Editorial-2014}.
+Datasets are now routinely archived as part of the publishing process.
+Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals, like \emph{Scientific Data}.
+Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
+In particular, the current publishing model has not yet embraced the idea that interesting datasets are constantly evolving, i.e.
+never "finished".
 
-In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
-While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
-Likewise, the examples cited here come from our own research areas, ecology and evolutionary biology, but we argue that the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.
+Most of the datasets published in this journal -- or other similar data descriptors -- are likely to be living entities, meaning the canonical (or master) version of the dataset changes through time, usually for the better.
+Typical changes involve: the addition of new data, improving the quality of existing data, or structuring of the database content.
+For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
+As research around a data product grows, there might be many such additions.
+Ideally, any changes to a dataset would become immediately available to any interested parties, while -- to ensure reproducibility -- past versions of the dataset would remain accessible.
 
-\subsection{Why versioning of data is needed and difficult}
+Unfortunately, current models for publishing datasets do not immediately solve the issue of of how to distribute successive versions of a dataset in an effective and scalable manner.
+The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo} -- where the data is attributed a Digital Object Identifier (\textsc{doi}).
+Once published, the datasets are immutable.
+While future updates may be published under an updated a separate \textsc{doi}, the \textsc{doi} system does not readily connect different versions of a dataset.
+That means that most published datasets exist as a singular snapshot of the dataset, taken at the itme the data descriptor was published.
+While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset.
+Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.
 
-Almost all useful databases are living--they \emph{should} get better
-through time either by the addition of new data or by improving the
-quality of existing data. For example a dataset on plant characteristics
-might become outdated as taxonomic changes occur. Updating the dataset might only
-require a small amount of work, but if there is no
-way to distribute the updated dataset, then that work has to be re-done by
-every user of that dataset. Moreover, those required changes add steps
-between the canonical (or master) dataset and the new analysis (and publication),
-making reproducibility of the new analysis more difficult than it needs
-to be. If the canonical dataset is versioned--and versions are
-accessible--the up-to-date data can be distributed, used, and analyses
-based on any of the previously released versions data can be reproduced.
-% Drawing this out; we might want to highlight that databases like
-% genbank can be seen as versioned (though someone who knows how they
-% work should check the details).  New entries are added but are
-% immutable.  This means the database gets better through time (more
-% entries) but each entry is fixed at creation.  Not all databases can
-% pull that off - for the reasons that you mention above and because
-% many ecological databases try to do the equivalent of getting the
-% "best measurement of X" for each species which is not the same as
-% "an observation of X".
+Many research groups solve the issue of updating data in a variety of suboptimal ways.
+A common solution is to email around the latest versions.
+Another approach is to repeat corrections or updates that have been made elsewhere.
+This practice is both inefficient and a barrier to reproducibility, as the dataset used in a publication may differ to the version previously made available on-line.
+There is therefore a strong need for an easy way to distribute changes to a dataset after it's initial publication to potential users, along with notes on what has changed and why since the previous version.
 
-% I think these need spelling out in a little more detail as I find
-% versioning a fairly hard sell to most people!
-While the benefits of versioned data are fairly clear, determining how to implement it has been contentious.
-The static model of data distribution widely used in science, where datasets are archived either as supplementary materials for a journal article
-or under its own Digital Object Identifier (\textsc{doi}), is simply not designed to provide streams of versioned
-data (Table \ref{tab:sql_v_versioneddata}).
+In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}); ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
+The issue of updating and expanding published data has been partly addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g.
+by adding multiple version of a genetic sequence).
+Yet, such large centralised repositories require a level of infrastructure that is beyond most research groups, or even countries (Table \ref{tab:sql_v_versioneddata}).
+Our focus here is on the wide range of datasets not covered by these repositories, such as those appearing in this journal.
+In most of these cases, the data
+collected will not be ``big'' but rather small-to-medium sized. Such data
+support research projects, but are not in most cases general enough to
+establish bespoke infrastructure for.
+Further, While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
+Likewise, while the examples cited here come from our own research areas, ecology and evolutionary biology, the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.
+
 % I think that this model is actually ok; for data archiving where the
@@
 % complain that it doesn't do something that it does not need to do.
-A particular issue with the static
-model is that data releases are mostly triggered by publication of academic
-papers. But for a living dataset, there may be multiple releases made in
-between papers. This is particularly the case with the discovery of small
-errors.
-% OK, here again; it's not so much medium-sized as *living*.  The
-% issues in practically doing the versioning are more acute when the
-% data are bigger than small, that's all.
-For any living dataset that is put to reuse, errors will
-inevitably be found, and there needs to be some easy  way to correct them and
-distribute the updated dataset.
 
-To overcome the limitations of static datasets, many research groups have
-setup dedicated webservers to host and distribute data (Table 1)\citep[e.g.][]{Madin-2016}. In these systems,
-datasets areas maintained in a structured database, such as SQL, and a custom front
-end to this database must be written.
-When designed well, such systems are very useful for updating or adding to a
-dataset. However, the infrastructure investment (designing maintaining, and
-hosting the database) is overkill for most projects and is a very
-substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many research groups. Moreover, the existence of a live webserver,
-does not in itself guarantee ongoing access to different versions of a
-dataset.
+% A particular issue with the static
+% model is that data releases are mostly triggered by publication of academic
+% papers. But for a living dataset, there may be multiple releases made in
+% between papers. This is particularly the case with the discovery of small
+% errors.
+% % OK, here again; it's not so much medium-sized as *living*.  The
+% % issues in practically doing the versioning are more acute when the
+% % data are bigger than small, that's all.
+% For any living dataset that is put to reuse, errors will
+% inevitably be found, and there needs to be some easy  way to correct them and
+% distribute the updated dataset.
 
-% This paragraph reads like it used to be elsewhere in the document.
-% I would flip it around and talk about small data first.  You could
+% To overcome the limitations of static datasets, many research groups have
+% setup dedicated webservers to host and distribute data (Table 1)\citep[e.g.][]{Madin-2016}. In these systems,
+% datasets areas maintained in a structured database, such as SQL, and a custom front
+% end to this database must be written.
+% When designed well, such systems are very useful for updating or adding to a
+% dataset. However, the infrastructure investment (designing maintaining, and
+% hosting the database) is overkill for most projects and is a very
+% substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many research groups. Moreover, the existence of a live webserver,
+% does not in itself guarantee ongoing access to different versions of a
+% dataset.
+
+\section{A lightweight, cheap, and scalable workflow for delivering versioned data}
+
+% I really like the focus on "release".  It's a happy coincidence that
+% it's what github uses in the interface.
+
+In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
+Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer.
+With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer.
+In either the development of software or data, successive versions are released over time (called ``releases'').
+The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
+Importantly, our \textsc{vdd} systems uses well-established tools, ensuring high-level performance and stability.
+
+The core technologies used are summarised in Table \ref{tab:technologies} and described further below.
+
+\subsection{Version control}
+
+% You could
 % talk about 'git' (referencing Ram-2013) and say "on the face of it,
@@
 % it up below, but I think it would go better here...
-When considering the right tool to store and deliver versioned data, the size
-of the dataset is another essential consideration. ``Big data'' brings with it
-specific needs and so requires a specific set of hardware and software tools.
-Examples of big data include genetic datasets, global datasets of species
-observations, and remote sensing data. This size of this data is bigger than
-that which individual research groups can maintain and often requires a
-specialized governmental or non-governmental institution with full-time staff
-to curate and maintain those research resources. Here we are particularly
-interested in smaller and more specialized problems (e.g., cases where a single 
-research group contributes most of the data). In most of these cases, the data
-collected will not be ``big'' but rather small-to-medium sized. Such data
-support research projects, but are not in most cases general enough to
-establish bespoke infrastructure for. But crucially the data themselves will change
-over time. And currently there is no infrastructure for organizing and
-distributing these types of ``living datasets''.
-
-We feel there is a  need for
-making versions of a dataset and delivering it to potential users, along
-with notes on what has changed and why since the previous version.
-Table \ref{tab:user_requirements} outlines what we consider requirements for
-three groups of people interacting with this system: maintainers,
-contributors, users. The solution is ideally low-cost (both in money and
-time), sophisticated, yet requiring little skill to setup and use.  Ideally
-there would also be a way for users to also become contributors, by flagging
-errors for the curator to correct in the next release, or even for the user to
-fix themselves.
-
-% Do we still need results sections in a data journal?
-\section{Results}
-
-\subsection{A lightweight, cheap, and scalable workflow for versioned data}
-
-% I really like the focus on "release".  It's a happy coincidence that
-% it's what github uses in the interface.  I'm increasingly of the
-% opinion that conflating source control and release (via
-% install_github) is not a good thing in the R community, but it is
-% seductive...
-In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
-Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local  computers.
-With a \textsc{vdd} system, data developers maintain a core set of files, which produce an organised dataset that can be similarly be ``installed'' on a user's local computer.
-In either the development of software or data, successive versions are released over time (called ``releases'').
-The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
-Importantly, our \textsc{vdd} systems uses well-established tools, ensuring high-level performance and stability.
-
-The core technologies used are summarised in Table \ref{tab:technologies} and described further below.
-
-\subsection{Version control}
 
@@
 Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.
-Essentially version control tracks line-by-line changes in text files and creates and maintains a history of those changes.
+In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes.
 Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}.
@@
 % The introduction talks at length about medium sized data and little about versioning but this section talks only about versioning
+
+% I really think that something like CKAN (though it doesn't support
+% holding data if I remember correctly) could totally be a complete
+% solution to the sorts of issues that we have here.
+
+% Discuss Genbank or other large sites?
+% The issues in practically doing the versioning are more acute when the
+% data are bigger than small, that's all.
+
+% From intro:
+% When considering the right tool to store and deliver versioned data, the size
+% of the dataset is another essential consideration. ``Big data'' brings with it
+% specific needs and so requires a specific set of hardware and software tools.
+% Examples of big data include genetic datasets, global datasets of species
+% observations, and remote sensing data. This size of this data is bigger than
+% that which individual research groups can maintain and often requires a
+% specialized governmental or non-governmental institution with full-time staff
+% to curate and maintain those research resources.
 

### 2017-09-21 ms.tex
comment out old abstract. address #21

--- 2017-09-21 ms.tex
@@
 % angle.
-The sharing and re-use of data has become a cornerstone of modern science.
+% The sharing and re-use of data has become a cornerstone of modern science.
 To date, the curation and distribution of data has largely been conducted under one of two frameworks: data is either deposited in large, collaborative databases (often run by institutions and staffed by computational professionals) or else published as static files alongside a specific publication or as a standalone product released by a research group.

### 2017-09-22 ms.tex
Fix small bug preventing compile
wording changes throughout

--- 2017-09-21 ms.tex
@@
 Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
-In particular, the current publishing model has not yet embraced the idea that interesting datasets are constantly evolving, i.e.
-never "finished".
+In particular, the current data publishing model has not yet embraced the idea that many  datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.
+These datasets are constantly evolving, i.e.
+never "finished", and may be the subject of repeated analysis or meta-analysis.
 
-Most of the datasets published in this journal -- or other similar data descriptors -- are likely to be living entities, meaning the canonical (or master) version of the dataset changes through time, usually for the better.
-Typical changes involve: the addition of new data, improving the quality of existing data, or structuring of the database content.
+Many, and perhaps most, of the datasets published in this journal are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time.
+Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions.
 For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
@@
 
-Unfortunately, current models for publishing datasets do not immediately solve the issue of of how to distribute successive versions of a dataset in an effective and scalable manner.
+Unfortunately, current models for publishing datasets do not immediately solve the issue of how to distribute successive versions of a dataset in an effective and scalable manner.
 The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo} -- where the data is attributed a Digital Object Identifier (\textsc{doi}).
 Once published, the datasets are immutable.
-While future updates may be published under an updated a separate \textsc{doi}, the \textsc{doi} system does not readily connect different versions of a dataset.
-That means that most published datasets exist as a singular snapshot of the dataset, taken at the itme the data descriptor was published.
+That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published.
 While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset.
@@
 
-Many research groups solve the issue of updating data in a variety of suboptimal ways.
-A common solution is to email around the latest versions.
+Many research groups internally solve the issue of versioning data in a variety of suboptimal ways.
+A common solution is to email around the latest versions with version numbers appended to the filename.
 Another approach is to repeat corrections or updates that have been made elsewhere.
 This practice is both inefficient and a barrier to reproducibility, as the dataset used in a publication may differ to the version previously made available on-line.
-There is therefore a strong need for an easy way to distribute changes to a dataset after it's initial publication to potential users, along with notes on what has changed and why since the previous version.
+There is therefore a strong need for an easy way to distribute changes to a dataset after its initial publication to potential users, along with notes on what has changed and why since the previous version.
 
@@
 Our focus here is on the wide range of datasets not covered by these repositories, such as those appearing in this journal.
-In most of these cases, the data
-collected will not be ``big'' but rather small-to-medium sized. Such data
-support research projects, but are not in most cases general enough to
-establish bespoke infrastructure for.
-Further, While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
-Likewise, while the examples cited here come from our own research areas, ecology and evolutionary biology, the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.
+In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized.
+Such data support important research projects on particular scientific questions, but are not in most cases general enough to warrant custom infrastructure.
+Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
+Likewise, while the examples cited here come from our own research areas the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.
 
@@
 They just don’t feel like its worth doing.
-There is rapid current work on this problem and prototypes available, and as such, it is likely that \texttt{git} for data will replaced by a more natural system for data in the near future  \citep{Fli, Dat}.
-This future system would be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently.
+There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
+% Add CKAN here?
 That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.
@@
 % I wonder if we should move the workflow description higher.
+
+% WKC: I agree. Not sure what to do with these two sections which are a bit of a tangent relative to our main point
 
@@
 \subsection{`Semantic' versioning}
+
+% WKC: this is back to our main point
 
@@
 
+% WKC: hmmmm not sure we've nailed this yet, but see
+% https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819
+
 Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
@@
 
-The problem of tracking data through time and versions is likely to be widespread. There are many alternate solutions, with associated features and drawbacks.
-% Reword
-The current structure of scientific research is very distributed, with research groups across the world working on the same question.
+%work on this paragraph more
+The problem of tracking data through time and versions is likely to be widespread.
+There are many alternate solutions, with associated features and drawbacks.
+Some large data distribution institutions have begon grabling with this issue in various ways (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), but the current structure of scientific research is very distributed, with research groups across the world working on the same question.
 Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.
@@
   \end{tabular}
-  % }
+}
 \label{tab:user_requirements}

### 2017-09-22 ms.tex
moving random paragraphs to discussion
Minor fixes to some citations

@@
 
-As is evidenced in the advent of this journal, publication of quality datasets is now considered a first-class scientific product.
-Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013,Editorial-2014}.
+As is evidenced in the advent of this journal \cite{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product.
+Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
 Datasets are now routinely archived as part of the publishing process.
-Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals, like \emph{Scientific Data}.
-Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
+Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals.
+Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \citep{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
 In particular, the current data publishing model has not yet embraced the idea that many  datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.
@@
 
-Many, and perhaps most, of the datasets published in this journal are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time.
+Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time.
 Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions.
@@
 
-In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
+In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
 Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer.
@@
 % some nasty "git show" call.
-Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.
+Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.
 For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
@@
 That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.
-
-\subsection{Cloud platforms for storing and distributing version controlled data}
-
-There are many current platforms for storing and distributing versioned controlled data for low or no cost.
-The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories.
-In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples).
-Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.
-Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
-
-% Though this section I am losing track of the big picture here;
-% there's lots of discussion of the positive benefits of "Something"
-% but without knowing what it is, it's unlikely that these will stick.
-% I wonder if we should move the workflow description higher.
-
-% WKC: I agree. Not sure what to do with these two sections which are a bit of a tangent relative to our main point
-
-\subsection{Potential for collaboration: allowing for multiple data contributors}
-
-One of the greatest benefits of using \texttt{git} for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
-This is enabled via the distributed and decentralised nature of \texttt{git} repositories.
-Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
-% We have constant drama at work because the different spreadsheet
-% writers we use across the platforms quote files inconsistently so
-% every save is a complete rewrite of the file.  It's totally
-% impossible to do any collaborative editing....
- or, when needed, flag where there are conflicts that need to be resolved.
-This makes collaboration easy.
-For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project.
-This process is called submitting a ``Pull request''.
-One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
- % This section conflates git and github more than is useful.  It
- % might be worth setting up github as the explicit platform here or
- % dialing back the language a bit.  GitLab and bitbucket have their
- % own "merge request" feature (I think).  It's also github that makes
- % repositories _discoverable_ which is a requirement for getting
- % outside collaborators in.  Oddly the decentralised nature of git is
- % not actually that useful in the github model because github is so
- % strongly centralised.  Will and I did some true decentralised git
- % collaboration on a plane to nescent once with a bare git repo on a
- % thumb drive.
 
@@
 
-A point-and-click workflow may create barriers for reproducibility \cite{Wilson-2014,Lowndes-2017}.
+A point-and-click workflow may create barriers for reproducibility \citep{Wilson-2014,Lowndes-2017}.
 Ideally, users should be able to access all versions of the database programmatically.
@@
 
-For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
+For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
 \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}.
@@
 But, nonetheless, workable systems are already possible.
+
+% potentially move to discussion? 
+\subsection{Cloud platforms for storing and distributing version controlled data}
+
+There are many current platforms for storing and distributing versioned controlled data for low or no cost.
+The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories.
+In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples).
+Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.
+Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
+
+% Though this section I am losing track of the big picture here;
+% there's lots of discussion of the positive benefits of "Something"
+% but without knowing what it is, it's unlikely that these will stick.
+% I wonder if we should move the workflow description higher.
+
+% WKC: I agree. Not sure what to do with these two sections which are a bit of a tangent relative to our main point
+
+% potentially move to discussion?
+\subsection{Potential for collaboration: allowing for multiple data contributors}
+
+One of the greatest benefits of using \texttt{git} for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
+This is enabled via the distributed and decentralised nature of \texttt{git} repositories.
+Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
+% We have constant drama at work because the different spreadsheet
+% writers we use across the platforms quote files inconsistently so
+% every save is a complete rewrite of the file.  It's totally
+% impossible to do any collaborative editing....
+ or, when needed, flag where there are conflicts that need to be resolved.
+This makes collaboration easy.
+For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project.
+This process is called submitting a ``Pull request''.
+One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
+ % This section conflates git and github more than is useful.  It
+ % might be worth setting up github as the explicit platform here or
+ % dialing back the language a bit.  GitLab and bitbucket have their
+ % own "merge request" feature (I think).  It's also github that makes
+ % repositories _discoverable_ which is a requirement for getting
+ % outside collaborators in.  Oddly the decentralised nature of git is
+ % not actually that useful in the github model because github is so
+ % strongly centralised.  Will and I did some true decentralised git
+ % collaboration on a plane to nescent once with a bare git repo on a
+ % thumb drive.
 

### 2017-09-25 ms.tex
organizing a bit

@@
 
-\subsection{Version control}
+\subsection{Version control and `semantic' versioning}
 
@@
 % With suggested edits, some of this moves up.
+
 Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.
@@
 
-% So, git doesn't actually work this way, though everyone seems to
-% think that it does.  Git stores *content* and then later tries to
-% de-duplicate it.  It then displays changes as line-by-line diffs
-% (but you can change this to display the changes other ways; see
-% --color-words --ignore-all-space for example.  The problem is that
-% the algorithms that git uses are optimised for typical sizes of code
-% and not for data.  Small data is actually totally fine to store with
-% git.  The other issue that turns up is if the files are _binary_
-% then an arbitrarily small change in the underlying data can result
-% in an arbitrarily large change in the binary representation, which
-% prevents deduplication.
-%
-% Git-LFS tries to step around this by storing large files separately
-% from git's general machinery.  Git-annex did the same.  But they
-% suffer from some of the other issues with git which are that it's a
-% bit of a pain for (especially novice) users to do nontrivial things
-% with.  I _always_ have to look up how to compare two versions across
-% branches.  And there's no simple way of loading in two versions of
-% the data into R without ending up with two working copies or doing
-% some nasty "git show" call.
-Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.
-For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
-% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”.
-Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach.
-From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate.
-They just don’t feel like its worth doing.
-There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
-% Add CKAN here?
-That said, the workflow we describe here currently works well with \texttt{git} and should work even more quickly with its future replacement.
-
-\subsection{`Semantic' versioning}
-
-% WKC: this is back to our main point
-
 To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be communicated to the user.
@@
 As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions.
-Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction.
-A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure).
+Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results.
+A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it \emph{might} change the results of some users' analyses.
 A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.
+These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.
 % I think that major version increments could still be looked at as
@@
 
-\subsection{Digital object identifiers}
-
-% I'm unclear on the precice benefits of dois here (and actually in
-% general to a point); some further clarification might be useful.  So
-% far as I know they are useful because (a) journal urls are terrible
-% and inconsistent (b) because everyone has them.  I don't really know
-% why else you'd need one.  Especially in our case we offer no promise
-% of persistence (deleting the github repo would break the doi and we
-% could update the repo to include different data trivially).  So what
-% is being added?  Our retrival system does not use the doi at all!
-
-% WKC: hmmmm not sure we've nailed this yet, but see
-% https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819
-
-Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
-At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}).
-Here we describe the particular integration of \texttt{GitHub} and \texttt{Zenodo}.
-It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project.
-\texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Figure \ref{fig:semantic} and ][]{Nielsen-2017}.
-This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.
-There is an unresolved issue about properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
-
 \section{Discussion}
@@
 % The introduction talks at length about medium sized data and little about versioning but this section talks only about versioning
+% WKC: now intro is about versioning and data size issue discussion moved here, which I think is an improvement
 
@@
 There are many alternate solutions, with associated features and drawbacks.
-Some large data distribution institutions have begon grabling with this issue in various ways (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), but the current structure of scientific research is very distributed, with research groups across the world working on the same question.
-Moreover at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.
+Some large data distribution institutions have begon grabling with this issue in various ways (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), but, while there are cases where data curation may be done by a large institution, the current structure of most scientific research is very distributed, with a few research groups across the world working on similar questions.
+Moreover, curation of data on particular questions requires, in many cases, a great deal of expertise in the field, so the experts in the field need to be involved in data curation.
+However, individual research groups have different capabilities compared to institutional data curators: at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.
 As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.
@@
 Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past it took great deal of money and expertise to set up a \textsc{vdd} system.
-The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.
-We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.
-
-The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s.
-Moreover, it is relatively easy to update and add new versions.
-Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.
-But, nonetheless, workable systems are already possible.
-
-% potentially move to discussion? 
-\subsection{Cloud platforms for storing and distributing version controlled data}
-
 There are many current platforms for storing and distributing versioned controlled data for low or no cost.
@@
 Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
+The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.
+We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.
 
-% Though this section I am losing track of the big picture here;
-% there's lots of discussion of the positive benefits of "Something"
-% but without knowing what it is, it's unlikely that these will stick.
-% I wonder if we should move the workflow description higher.
+The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Figure \ref{fig:technology_stack}).
+Moreover, it is relatively easy to update and add new versions.
+Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.
+But, nonetheless, workable systems are already possible.
 
-% WKC: I agree. Not sure what to do with these two sections which are a bit of a tangent relative to our main point
+One of the greatest benefits of using a cloud--based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
+Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
+This makes collaboration easy.
 
-% potentially move to discussion?
-\subsection{Potential for collaboration: allowing for multiple data contributors}
+Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
+At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}).
+Here we describe the particular integration of \texttt{GitHub} and \texttt{Zenodo}.
+It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project.
+\texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Figure \ref{fig:semantic} and ][]{Nielsen-2017}.
+This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.
+There is an unresolved issue about properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
 
-One of the greatest benefits of using \texttt{git} for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
-This is enabled via the distributed and decentralised nature of \texttt{git} repositories.
-Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible),
+% I'm unclear on the precice benefits of dois here (and actually in
+% general to a point); some further clarification might be useful.  So
+% far as I know they are useful because (a) journal urls are terrible
+% and inconsistent (b) because everyone has them.  I don't really know
+% why else you'd need one.  Especially in our case we offer no promise
+% of persistence (deleting the github repo would break the doi and we
+% could update the repo to include different data trivially).  So what
+% is being added?  Our retrival system does not use the doi at all!
+
+% WKC: hmmmm not sure we've nailed this yet, but see
+% https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819
+
+\subsection{Future directions}
+
+Technology for data management and distribution is moving forward rapidly.
+The workflow we describe above works well with current tools and should work even more quickly with its future replacement.
+Here we highlight a few rapidly developing future tools that researchers should be aware of.
+
+Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.
+For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
+% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”.
+Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach.
+From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate.
+They just don’t feel like its worth doing.
+There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
+% Add CKAN here?
+That said...
+% Add CKAN and Git-LFS, and say something useful about the future
+
 % We have constant drama at work because the different spreadsheet
@@
 % impossible to do any collaborative editing....
- or, when needed, flag where there are conflicts that need to be resolved.
-This makes collaboration easy.
-For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project.
+% For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project.
 This process is called submitting a ``Pull request''.
@@
  % thumb drive.
+
+% So, git doesn't actually work this way, though everyone seems to
+% think that it does.  Git stores *content* and then later tries to
+% de-duplicate it.  It then displays changes as line-by-line diffs
+% (but you can change this to display the changes other ways; see
+% --color-words --ignore-all-space for example.  The problem is that
+% the algorithms that git uses are optimised for typical sizes of code
+% and not for data.  Small data is actually totally fine to store with
+% git.  The other issue that turns up is if the files are _binary_
+% then an arbitrarily small change in the underlying data can result
+% in an arbitrarily large change in the binary representation, which
+% prevents deduplication.
+%
+% Git-LFS tries to step around this by storing large files separately
+% from git's general machinery.  Git-annex did the same.  But they
+% suffer from some of the other issues with git which are that it's a
+% bit of a pain for (especially novice) users to do nontrivial things
+% with.  I _always_ have to look up how to compare two versions across
+% branches.  And there's no simple way of loading in two versions of
+% the data into R without ending up with two working copies or doing
+% some nasty "git show" call.
 

### 2017-09-25 ms.tex
more details

@@
 However, individual research groups have different capabilities compared to institutional data curators: at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.
-As such we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.
+As such, we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.
 
-Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past it took great deal of money and expertise to set up a \textsc{vdd} system.
-There are many current platforms for storing and distributing versioned controlled data for low or no cost.
+Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many current platforms for storing and distributing versioned controlled data for low or no cost.
 The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories.
@@
 This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.
-There is an unresolved issue about properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
 
@@
 
+% subsection only temporary as not allowed by journal guidelines.
 \subsection{Future directions}
@@
 % Add CKAN and Git-LFS, and say something useful about the future
+
+Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
 

### 2017-09-25 ms.tex
more clean-up

@@
 \texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).
-Those releases can be downloaded directly by users via point-and-click, or accessed via the \texttt{GitHub} \textsc{api}.
-% I have no idea what this next sentence means.
-Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.
+Those releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.
+This system also allows the curator to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.
+
+% I have no idea what this next sentence means.  WKC: cut
+%Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.
 Releases can also be linked to digital object identifiers and code-based users via for example an \textsc{R} package.
@@
 This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.
+Users may also have several versions of the same database on their computer and unambiguously access the different versions with one simple function call.
+The dataset versions are cached between runs such that the data are accessed by name and version, so that after the first download of each version access to the data should be very fast.
+
+% Something not mentioned; that the versions are cached between runs
+% so that the data are accessed by name/version (rather than by
+% homebrewing a local cache and manually doing a bunch of file name
+% mangling) and that all accesses after the first are very fast
+% WKC: I had a go at this--see above
 
@@
 Then as the dataset grows over time, researchers update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number (preferably
-using \href{http://semver.org/}{semantic versioning}). All the versions
-will be available to the user.
-
-% Something not mentioned; that the versions are cached between runs
-% so that the data are accessed by name/version (rather than by
-% homebrewing a local cache and manually doing a bunch of file name
-% mangling) and that all accesses after the first are very fast
+using \href{http://semver.org/}{semantic versioning}). All the releases are simultaneously
+available to the user by both point-and-click and programmatically.
 
@@
 % Reword next senetnce
-However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system does not critically depend on using \texttt{datastorr} or \texttt{GitHub}.
-We would love to see implementations in \texttt{Python} or other languages and using other cloud platforms for data distribution.
-
-% Karthik has started with rdrop2 again I think; I do think that is a reasonable low-entry way of distributing things
+However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system.
+The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
 
@@
 % WKC: now intro is about versioning and data size issue discussion moved here, which I think is an improvement
-
-% I really think that something like CKAN (though it doesn't support
-% holding data if I remember correctly) could totally be a complete
-% solution to the sorts of issues that we have here.
 
@@
 
-% WKC: hmmmm not sure we've nailed this yet, but see
+% WKC: hmmmm not sure we've nailed this yet, see
 % https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819
@@
 % subsection only temporary as not allowed by journal guidelines.
+% trick is to say something about how things are getting better rapidly, but not scare people away from doing this now.  
 \subsection{Future directions}
@@
 The workflow we describe above works well with current tools and should work even more quickly with its future replacement.
-Here we highlight a few rapidly developing future tools that researchers should be aware of.
+Here, we highlight a few rapidly developing future tools that researchers should be aware of.
 
@@
 For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
-% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”.
-Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach.
-From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate.
-They just don’t feel like its worth doing.
 There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
-% Add CKAN here?
-That said...
-% Add CKAN and Git-LFS, and say something useful about the future
+% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future
 
@@
 % some nasty "git show" call.
+
+% I really think that something like CKAN (though it doesn't support
+% holding data if I remember correctly) could totally be a complete
+% solution to the sorts of issues that we have here.
+
+% Karthik has started with rdrop2 again I think; I do think that is a reasonable low-entry way of distributing things
+
+% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”.
+Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach.
+From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate.
+They just don’t feel like its worth doing.
 

### 2017-09-25 ms.tex
Reorder authors
Fixes #26

@@
 \title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
-\author{Richard G. FitzJohn$^1$, Daniel S. Falster$^2$,\\ Matthew
+\author{Daniel S. Falster$^1$, Richard G. FitzJohn$^2$,\\ Matthew
   W. Pennell$^3$, and William K. Cornwell$^{2,*}$}
 \affiliation{
-$^1$ Imperial College, London, United Kingdom\\
-$^2$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
+$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
 University of New South Wales, Sydney, NSW 2052, Australia\\
+$^2$ Imperial College, London, United Kingdom\\
 $^3$ Department of Zoology and Biodiversity Research Centre,\\

### 2017-09-25 ms.tex
mwp edits

@@
 \author{Daniel S. Falster$^1$, Richard G. FitzJohn$^2$,\\ Matthew
-  W. Pennell$^3$, and William K. Cornwell$^{2,*}$}
+  W. Pennell$^3$, and William K. Cornwell$^{1,*}$}
 \affiliation{
 $^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
+$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom\\
 University of New South Wales, Sydney, NSW 2052, Australia\\
-$^2$ Imperial College, London, United Kingdom\\
 $^3$ Department of Zoology and Biodiversity Research Centre,\\
@@
 To help this process, numerous linked tools have been developed to allow quick and easy data sharing:    a data file for a particular analysis can be quickly uploaded, given a digital identifier, and shared.
-Thus far, however, this framework does not account for on-going scientific progress: for many important problems, epitomised by problems that lead to one or more meta-analyses, datasets will continue to grow with time---more data points will be added, and new and more useful data structures will be created.
-Essentially this means that datasets, like science in general, progress with time.
+Thus far, however, this framework does not accommodate on-going scientific progress: for many important problems, particularly those suitable for meta-analyses, datasets will continue to grow with time --- more data points will be added, and new and more useful data structures will be created.
+Essentially this means that datasets, like science in general, improve with time.
 We suggest that many different datasets would be more useful if organized as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.
@@
 We introduce a system called Versioned Data Delivery (VDD) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
-These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that naturally matches the scientific process.
+These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.
 
@@
 
-As is evidenced in the advent of this journal \cite{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product.
-Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
+As is evidenced in the advent of this journal \citep{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product.
+Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \citep[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
 Datasets are now routinely archived as part of the publishing process.
@@
 These datasets are constantly evolving, i.e.
-never "finished", and may be the subject of repeated analysis or meta-analysis.
+they are never "finished", and may be the subject of repeated analysis or meta-analysis.
 
@@
 As research around a data product grows, there might be many such additions.
-Ideally, any changes to a dataset would become immediately available to any interested parties, while -- to ensure reproducibility -- past versions of the dataset would remain accessible.
+Ideally, any changes to a dataset would become immediately available to any interested parties, while --- to ensure reproducibility --- past versions of the dataset would remain accessible.
 
 Unfortunately, current models for publishing datasets do not immediately solve the issue of how to distribute successive versions of a dataset in an effective and scalable manner.
-The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo} -- where the data is attributed a Digital Object Identifier (\textsc{doi}).
+The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories, such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}, where the data is attributed a Digital Object Identifier (\textsc{doi}).
 Once published, the datasets are immutable.
@@
 Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results.
-A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it \emph{might} change the results of some users' analyses.
+A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses.
 A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.
@@
 
-At current there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.
+Currently there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.
 The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).
@@
 \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}.
-And importantly \texttt{baad.data} contains no actual data so is very quick to install and takes up virtually no space on the user's hard-drive.
+And importantly \texttt{baad.data} contains no actual data so it is very quick to install and takes up virtually no space on the user's hard-drive.
 % Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”.
@@
 %work on this paragraph more
-The problem of tracking data through time and versions is likely to be widespread.
-There are many alternate solutions, with associated features and drawbacks.
-Some large data distribution institutions have begon grabling with this issue in various ways (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), but, while there are cases where data curation may be done by a large institution, the current structure of most scientific research is very distributed, with a few research groups across the world working on similar questions.
-Moreover, curation of data on particular questions requires, in many cases, a great deal of expertise in the field, so the experts in the field need to be involved in data curation.
-However, individual research groups have different capabilities compared to institutional data curators: at a time of funding scarcity, many world-class research groups do not have excess funds to hire and support technology specialists.
-As such, we argue that there is a niche for a easy, cheap, and flexible \textsc{vdd} system.
+We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact.
+In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought a great deal about data versioning and developed use-case specific solution.
+But as we mentioned above, these datasets tend to be those managed by large, well-funded institutions and involve standard data types (e.g., sequences).
+But for an individual research group with expertise on a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter.
+As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.
 
@@
 Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
-The proliferation of cloud--based tools and open--source software have now reduced these roadblocks considerably.
+The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.
 We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.
@@
 
-One of the greatest benefits of using a cloud--based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
+One of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
 Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
@@
 Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
+Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.
 

### 2017-09-25 ms.tex
minor changes to discussion

@@
 
-The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Figure \ref{fig:technology_stack}).
+%The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Figure \ref{fig:technology_stack}).
 Moreover, it is relatively easy to update and add new versions.
@@
 
-One of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
-Multiple users can make changes to different parts of the code (or in our case data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
-This makes collaboration easy.
+Additionally, one of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
+Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
+As a result adopting a \textsc{vdd} system has the added benefit of facilitating collaboration among research groups.
 
-Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
+%Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
 At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}).

### 2017-09-25 ms.tex
copyediting

@@
 
-Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will changes through time.
-Typical changes may involve: the addition of new data, improving the quality of existing data, better linked to external databases, or re-structuring of the database content to be able to address new questions.
-For example a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
+Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will change through time.
+Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions.
+For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
 As research around a data product grows, there might be many such additions.
-Ideally, any changes to a dataset would become immediately available to any interested parties, while --- to ensure reproducibility --- past versions of the dataset would remain accessible.
+Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure reproducibility.
 
-Unfortunately, current models for publishing datasets do not immediately solve the issue of how to distribute successive versions of a dataset in an effective and scalable manner.
+However, current models for publishing datasets do not facilitate distributing successive versions of a dataset in an effective and scalable manner.
 The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories, such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}, where the data is attributed a Digital Object Identifier (\textsc{doi}).
@@
 Yet, such large centralised repositories require a level of infrastructure that is beyond most research groups, or even countries (Table \ref{tab:sql_v_versioneddata}).
-Our focus here is on the wide range of datasets not covered by these repositories, such as those appearing in this journal.
+Our focus here is on the wide range of datasets, such as those appearing in this journal, that are not covered by these repositories.
 In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized.
@@
 Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
-Likewise, while the examples cited here come from our own research areas the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.
+Likewise, while the examples cited here come from our own research areas, the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.
 
@@
 
-To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be communicated to the user.
+To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be transparent and understandable by both data sharers and users.
 Crucially a user should be immediately aware of the relationship among versions.
@@
 
-Currently there are two classes of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.
+Currently there are two types of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.
 The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).
@@
 We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact.
-In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought a great deal about data versioning and developed use-case specific solution.
-But as we mentioned above, these datasets tend to be those managed by large, well-funded institutions and involve standard data types (e.g., sequences).
-But for an individual research group with expertise on a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter.
+In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought a great deal about data versioning and developed use-case specific solutions.
+But as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences).
+But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter.
 As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

### 2017-09-25 ms.tex
added acknowledgements section. please add yours.

@@
 \title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
-\author{Daniel S. Falster$^1$, Richard G. FitzJohn$^2$,\\ Matthew
+\author{Daniel S. Falster$^1,*$, Richard G. FitzJohn$^2$,\\ Matthew
   W. Pennell$^3$, and William K. Cornwell$^{1,*}$}
@@
 University of New South Wales, Sydney, NSW 2052, Australia\\
+$^2$ Imperial College, London, United Kingdom\\
 $^3$ Department of Zoology and Biodiversity Research Centre,\\
 University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
-$^*$ Corresponding author: w.cornwell@unsw.edu.au\\
+$^*$ Corresponding author: daniel.falster@unsw.edu.au\\
 }
@@
 
+\section{Acknowledgements}
+
+M.W.P. was funded by a NSERC Discovery Grant.
+
 \newpage
@@
 
+\newpage
+
 \bibliography{refs}

### 2017-10-16 ms.tex
Edits up to discussion
Closes #29, closes #30, closes #24

@@
 \title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
-\author{Daniel S. Falster$^1,*$, Richard G. FitzJohn$^2$,\\ Matthew
-  W. Pennell$^3$, and William K. Cornwell$^{1,*}$}
+\author{Daniel S. Falster$^{1,*}$, Richard G. FitzJohn$^2$,\\ Matthew
+  W. Pennell$^3$, and William K. Cornwell$^{1}$}
 \affiliation{
-$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,\\
+$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
+University of New South Wales, Sydney, NSW 2052, Australia\\
 $^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom\\
-University of New South Wales, Sydney, NSW 2052, Australia\\
-$^2$ Imperial College, London, United Kingdom\\
 $^3$ Department of Zoology and Biodiversity Research Centre,\\
@@
 
-% For Analyses and Articles, the main text (excluding abstract, Methods, references and figure legends) is approximately 3,000 words.
-The abstract is typically 100-170 words, unreferenced.
-An Introduction is followed by sections headed Results, Discussion, Methods.
-The Methods and Results should be divided by topical subheadings; the Discussion does not contain subheadings.
-The Methods should be followed by References, Data Citations (optional), Acknowledgements and a Competing interests statement.
-A Data Citation section should be included when externally hosted datasets are referred to in the work and each can be cited formally with a unique, stable identifier and repository name.
-
 \section{Summary}
@@
 In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data curators and users of even small-to-medium sized datasets.
-We introduce a system called Versioned Data Delivery (VDD) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
+We introduce a system called Versioned Data Delivery (\textsc{vdd}) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
 These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.
@@
 
-Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art version of the dataset will change through time.
+Many, perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time.
 Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions.
@@
 However, current models for publishing datasets do not facilitate distributing successive versions of a dataset in an effective and scalable manner.
-The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories, such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}, where the data is attributed a Digital Object Identifier (\textsc{doi}).
-Once published, the datasets are immutable.
+The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories where the data is attributed a Digital Object Identifier (\textsc{doi}) -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}.
+Publishing involves a single data deposition.
+And once published, datasets are immutable.
 That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published.
@@
 Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.
+We refer to this type of data delivery as being ``Static'' (Table \ref{tab:publishing_models}).
 
-Many research groups internally solve the issue of versioning data in a variety of suboptimal ways.
+Many research groups solve the issue of versioning data internally in a variety of ad hoc and suboptimal ways.
 A common solution is to email around the latest versions with version numbers appended to the filename.
 Another approach is to repeat corrections or updates that have been made elsewhere.
-This practice is both inefficient and a barrier to reproducibility, as the dataset used in a publication may differ to the version previously made available on-line.
+Alas, these practices are both inefficient and a barrier to reproducibility, as the dataset used in a future publications may differ to the version previously made available on-line.
 There is therefore a strong need for an easy way to distribute changes to a dataset after its initial publication to potential users, along with notes on what has changed and why since the previous version.
 
-In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}); ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
-The issue of updating and expanding published data has been partly addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g.
+In this article we
+\begin{enumerate}
+  \item Introduce the concept of Versioned Data Delivery (\textsc{vdd}).
+  \item Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets.
+  \item Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
+\end{enumerate}
+The issue of updating and expanding published data has already been addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g.
 by adding multiple version of a genetic sequence).
-Yet, such large centralised repositories require a level of infrastructure that is beyond most research groups, or even countries (Table \ref{tab:sql_v_versioneddata}).
+Yet these ``Dynamic'' web databases (Table \ref{tab:publishing_models}), require a level of infrastructure that is beyond most research groups.
 Our focus here is on the wide range of datasets, such as those appearing in this journal, that are not covered by these repositories.
 In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized.
-Such data support important research projects on particular scientific questions, but are not in most cases general enough to warrant custom infrastructure.
+Such data support important research projects on particular scientific questions, but are not, in most cases, general enough to warrant custom infrastructure.
 Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
-Likewise, while the examples cited here come from our own research areas, the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.
-
-% I think that this model is actually ok; for data archiving where the
-% data are a record of what was used in the paper, a single immutable
-% data set makes sense.  It just doesn't extend to things where the
-% research output *is* the dataset.  So this is more a case of
-% pointing out that this model does not apply well to all sorts of
-% data - given how successful the archive-at-publication model has
-% been at getting data out of people's drawers I don't think we should
-% complain that it doesn't do something that it does not need to do.
-
-% A particular issue with the static
-% model is that data releases are mostly triggered by publication of academic
-% papers. But for a living dataset, there may be multiple releases made in
-% between papers. This is particularly the case with the discovery of small
-% errors.
-% % OK, here again; it's not so much medium-sized as *living*.  The
-% % issues in practically doing the versioning are more acute when the
-% % data are bigger than small, that's all.
-% For any living dataset that is put to reuse, errors will
-% inevitably be found, and there needs to be some easy  way to correct them and
-% distribute the updated dataset.
-
-% To overcome the limitations of static datasets, many research groups have
-% setup dedicated webservers to host and distribute data (Table 1)\citep[e.g.][]{Madin-2016}. In these systems,
-% datasets areas maintained in a structured database, such as SQL, and a custom front
-% end to this database must be written.
-% When designed well, such systems are very useful for updating or adding to a
-% dataset. However, the infrastructure investment (designing maintaining, and
-% hosting the database) is overkill for most projects and is a very
-% substantial barrier to entry. Furthermore, the system requires ongoing funds to maintain it, making it often unsustainable for many research groups. Moreover, the existence of a live webserver,
-% does not in itself guarantee ongoing access to different versions of a
-% dataset.
 
@@
 
-% I really like the focus on "release".  It's a happy coincidence that
-% it's what github uses in the interface.
-
-In brief, the workflow we present, which we call Versioned Data Delivery (\textsc{vdd}), borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
+In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing Versioned Data (\textsc{vd}).
 Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer.
@@
 The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
-Importantly, our \textsc{vdd} systems uses well-established tools, ensuring high-level performance and stability.
+Importantly, our \textsc{vdd} systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.
 
 The core technologies used are summarised in Table \ref{tab:technologies} and described further below.
+Several groups will interact with the \textsc{vdd} system, including dataset \emph{maintainers}, \emph{contributors}, and \emph{users}.
+The requirements of these different groups are outlined in Table \ref{tab:user_requirements}.
 
-\subsection{Version control and `semantic' versioning}
-
-% You could
-% talk about 'git' (referencing Ram-2013) and say "on the face of it,
-% versioning tools like git appear to provide the ideal interface for
-% versioning data".  Then talk about how git is designed for code and
-% breaks down as the data involved exceed a few 10s to hundreds of MB.
-% After that, we can say that the aim is not to support truely "Big"
-% data because research groups capable of generating such data
-% probably have the wherewithall to look after it.  I know you bring
-% it up below, but I think it would go better here...
-
-% With suggested edits, some of this moves up.
+\subsection{Version control}
 
@@
 In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes.
-Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}.
-\texttt{git} is attractive for data management because it tracks all changes in monitored files.
-This history is visible to anyone interacting with the repository.
-It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.
+Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016, Lowndes-2017}.
+\texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g.
+``.csv'', ``.tsv'', ``.txt'').
+The history is visible to anyone interacting with the repository.
+It also allows users to annotate changes (``commits'') with informative messages detailing the rationale for those changes.
+In it's present form, \texttt{git} can handle individual data files at least up to 100MB, which includes a large fraction of scientific cases.
 
-To fully realize the benefits of a versioned data, the \textsc{vdd} system needs to be transparent and understandable by both data sharers and users.
-Crucially a user should be immediately aware of the relationship among versions.
-Software development has dealt with a very similar problem, and we suggest there is benefit in adopting the notation from that field.
-Specifically, we suggest applying the theory of ``semantic versioning'' developed for software distribution.
-This system uses (the potentially familiar) notation of form ``1.0.0'' for successive versions (Fig \ref{fig:semantic}).
-As has been noted (REFS), labelling of data versions presents a similar problem to the labelling of software releases: signalling to users the types of change that has occurred between successive versions.
-Applying semantic versioning to data, a change from 1.0.0 to 1.0.1 implies tiny changes, for example a minor error correction which is unlikely to ``break'', or change in a substancial way, users' analyses, although there may be minor changes in the results.
-A change from 1.0.0 to 1.1.0 implies a substantial enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses.
-A change from 1.0.0 to 2.0.0 implies a very major revision, for example improving the entire structure of the dataset and adding new columns.
+As a general strategy for tracking a dataset under version control with git, we recommend:
+\begin{enumerate}
+  \item Dataset \emph{maintainers} establish a separate \texttt{git} repository for any dataset to be distributed.
+  \item Saving all files as plain text, so that \texttt{git} can identify line-by-line changes. For example, you might maintain tabular data as a ``csv''.
+  \item Saving data in their rawest form. In some datasets you might only have a single file. Others may have may files that get manipulated or combined in some way to produce a unified product.
+  \item Including in the \texttt{git} repository any code needed to manipulate or compile the raw data files into the final dataset.
+For example, you might combine many independent datasets into one unified dataset.
+  \item Documenting any change in the dataset as a ``commit'' into the \texttt{git} repository logs, with informative commit messages outlining why a particular change was made.
+\end{enumerate}
+
+\subsection{Hosting and distributing versioned data}
+
+Datasets stored under version control via \texttt{git} reach their real potential when hosted at a suitable internet hosting service \citep{Ram-2013,Perkel-2016}.
+Here we focus on the platform \texttt{GitHub} (Table \ref{tab:technologies}), but similar functions could be achieved via other providers such as \href{http://bitbucket.org}{bitbucket.org} and \href{http://gitlab.com}{gitlab.com}.
+Hosting of a \texttt{git} repository enables dataset \emph{maintainers} to connect with other potential \emph{contributors} and also \emph{users} (Fig.
+\ref{fig:technology_stack}).
+These platforms are designed to work with \texttt{git} repositories, and thus offer many helpful features, such as ability to record issues, host documentation, or review edits over time.
+
+Of particular interest for current purposes, is \texttt{GitHub}'s ability to host a stream of ``releases'' from the dataset, alongside the \texttt{git} repository containing all the rawfiles.
+Each ``release'' is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset \emph{Maintainer} decided to generate a new  version of the data for distribution.
+While \emph{users} could in principle download the entire \texttt{git} repository, actually all they want are these 'releases'.
+
+Deciding when to make a new release is at the discretion of the dataset \emph{Maintainer}.
+In practice, one makes fewer releases than one does commits into the \texttt{git} repository, though there is nothing stopping \emph{maintainers} from releasing a new version for every commit.
+The flexibility here allows \emph{maintainers} to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.
+
+Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data users to access the data: those that interact with the data via point and click downloading and those that use programmatic interaction (Fig.
+\ref{fig:technology_stack} and Table \ref{tab:user_requirements}).
+Specifically, \texttt{GitHub} releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.
+
+\subsection{`Semantic' versioning}
+
+To realize the full benefits of a versioned controlled dataset, users should be able to easily intuit the types of changes that have occurred among versions.
+Since software development has effectively already dealt with almost identical; problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.
+
+Specifically, we suggest applying the theory of ``semantic versioning'', developed for software distribution (see \href{http://semver.org/}{semver.org}), to successive releases of a dataset.
+Semantic versioning uses tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers.
+For example, version ``1.0.0''.
+Changes in the version number then signal the type of changes that occurred in the dataset (Fig \ref{fig:semantic}).
+Applying semantic versioning to data, a change from:
+\begin{itemize}
+  \item  {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a {\bf correction} or ``patch'', for example a small error correction which is unlikely to ``break'', or change in a substantial way, a users' analyses, although there may be minor changes in the results.
+  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a {\bf minor} enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses.
+  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a very {\bf major} revision, for example improving the entire structure of the dataset and adding new columns.
 These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.
+\end{itemize}
+Any user of a dataset where releases were tagged using semantic versioning, would immediately know the types of change that might have occurred when requesting different versions and set their there expectations accordingly.
+
 % I think that major version increments could still be looked at as
@@
 
-\subsection{Distributing data to different user needs}
-
-Currently there are two types of data users: those that  interact with the data via point and click downloading and those that use code-based interaction.
-The system we describe allows for both types (Figure \ref{fig:technology_stack} and Table \ref{tab:user_requirements}).
-\texttt{GitHub} has a feature called ``releases'' in which specific commits can be turned into a version for distribution (with associated version number).
-Those releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.
-This system also allows the curator to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.
-
-% I have no idea what this next sentence means.  WKC: cut
-%Releases can also be embedded into any kind of website that allows the data curator to facilitate discovery by more users and to provide context.
-Releases can also be linked to digital object identifiers and code-based users via for example an \textsc{R} package.
-
 \subsection{\texttt{datastorr} and dataset-specific \texttt{R} packages}
 
-A point-and-click workflow may create barriers for reproducibility \citep{Wilson-2014,Lowndes-2017}.
-Ideally, users should be able to access all versions of the database programmatically.
-Code to access the \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle for reproducibility.
-To make access via code as easy as possible for users, we offer a novel implementation of a \textsc{vdd} system.
-We have written an \textsc{R} package, \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that a) interacts with the \texttt{GitHub} \textsc{api} to pull down versions of the dataset; and b) constructs the shell of a second, database specific \textsc{R} package, which can be used to access versions of the dataset.
-Using this system, a researcher can create and distribute a \textsc{R} package that facilitates access to their data with (very) minimal computational skills.
-We have focussed on the \texttt{R} platform \cite{R-2017} as one of the most prominent platforms for data analysis.
+To aid reproducibility and efficient usage, many users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}).
+Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle.
+To make it easier for users to access \textsc{vd} via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.
+
+Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that facilitates access to releases of any \textsc{vd} hosted on \texttt{GitHub} (Fig.
+\ref{fig:technology_stack}).
+Specifically, the  \texttt{datastorr} package:
+\begin{enumerate}
+  \item Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}.
+  \item Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
+\end{enumerate}
+Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.
 
 For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
-\texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}.
-And importantly \texttt{baad.data} contains no actual data so it is very quick to install and takes up virtually no space on the user's hard-drive.
-% Dan Noble - Next 5 lines: I found this part to be a bit difficult to grasp as you discuss how updated versions can "download the latest version”, but then in the same sentence discuss how this works seamlessly “offline”.
-You may want to discuss this in greater depth as I don’t see how this is possible.
-When the function \texttt{baad.data::baad\_data()} is called, the system will download the latest version; if there have not been any changes since the last time the function was called, it will simply retrieve the data from memory, making the system very efficient (and also allowing the system to work seamlessly if the computer is offline).
-If a researcher requires an older version of \textsc{BAAD}, specifying the version number of the data as an argument, e.g.
-\texttt{baad.data::baad\_data("1.0.1")} will retrieve it, without overwriting any other dataset versions they may have previously downloaded.
-This enables users to reproduce an analysis using a specific version of the data, even while new data is posted.
-Users may also have several versions of the same database on their computer and unambiguously access the different versions with one simple function call.
-The dataset versions are cached between runs such that the data are accessed by name and version, so that after the first download of each version access to the data should be very fast.
+The R package \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}.
+For a user, accessing a version of the data is a simple as typing a single line of code (Fig.
+\ref{fig:technology_stack}).
+Accessing a different version of the data involves changing only the version number.
+From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \citep[e.g.][]{Duursma-2016,Falster-2016}.
 
-% Something not mentioned; that the versions are cached between runs
-% so that the data are accessed by name/version (rather than by
-% homebrewing a local cache and manually doing a bunch of file name
-% mangling) and that all accesses after the first are very fast
-% WKC: I had a go at this--see above
+Using \texttt{datastorr}, dataset \emph{maintainers} can set up their own R package to deliver there versioned dataset simply by providing:
+\begin{enumerate}
+  \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored
+  \item the filename in the release that will contain the \textsc{vd}
+  \item the function used to load this into \texttt{R}.
+\end{enumerate}
+A full tutorial explaining precisely how to set this up is available at \smurl{github.com/ropenscilabs/datastorr}.
 
-Using \texttt{datastorr}, researchers can set up their own versioned database (and R package) simply by providing: 1) a \texttt{GitHub} repository name (e.g., ``traitecoevo/taxonlookup''); 2) a single file that will contain the versioned data (e.g., ``taxonlookup.csv''); 3) and the function used to load this into \texttt{R} (e.g., \texttt{read.csv()}).
-A tutorial explaining precisely how to set this up is available at \smurl{github.com/ropenscilabs/datastorr}.
+Then as the dataset grows over time, the \emph{maintainers} update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number.
+All the releases are simultaneously available to any user, both point-and-click and programmatically.
 
-Then as the dataset grows over time, researchers update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number (preferably
-using \href{http://semver.org/}{semantic versioning}). All the releases are simultaneously
-available to the user by both point-and-click and programmatically.
-
-We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users.
-And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
-% Reword next senetnce
-However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system.
-The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
+The dataset-specific packages created by \texttt{datastorr} are designed to be computationally efficient and also work offline.
+Packages created by \texttt{datastorr} contain no actual data, only the rules for fetching the data.
+As such, the basic package structure is quick to install and takes up virtually no space on the user's hard-drive.
+The package functions by fetching each data version once (the first time it is requested), and then caching these files locally for future reuse.
+Moreover, \emph{users} can have several versions of the same dataset on their computer and unambiguously access the different versions with one simple function call.
 
 \section{Discussion}
+
 % General problem. An example of solution. Key features
@@
 We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact.
-In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought a great deal about data versioning and developed use-case specific solutions.
-But as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences).
+In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions.
+But so far these as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences).
 But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter.
 As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.
+
+% Moved from above, need to integrate better
+We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users.
+And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
+% Reword next senetnce
+However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system.
+The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
 
@@
 The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.
-We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.
+We argue that by linking together the tools described above, researchers can now produce, curate, and distribute \textsc{vd} much more easily and cheaply than in the past.
 
-%The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and code-based access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Figure \ref{fig:technology_stack}).
+%The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and programmatic access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Fig.
+\ref{fig:technology_stack}).
 Moreover, it is relatively easy to update and add new versions.
@@
 It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project.
-\texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Figure \ref{fig:semantic} and ][]{Nielsen-2017}.
+\texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Fig.
+\ref{fig:semantic} and ][]{Nielsen-2017}.
 This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.
@@
 
-Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \cite{Perkel-2016}.
+Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.
 For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
@@
 % impossible to do any collaborative editing....
-% For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project maintainers to be reviewed and potentially included in the project.
+% For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project \emph{maintainers} to be reviewed and potentially included in the project.
 This process is called submitting a ``Pull request''.
@@
 \section{Acknowledgements}
-
-M.W.P. was funded by a NSERC Discovery Grant.
+We thank D Noble for comments on an earlier draft and C Boettiger  for helpful discussions. DSF was funded by the Australian Research Council. MWP was funded by a NSERC Discovery Grant.
 
@@
   } 
-\label{tab:sql_v_versioneddata}
+\label{tab:publishing_models}
 \end{table}
@@
 \centering
-\caption{Glossary of technologies used to maintain, store, and distribute the Versioned Data Delivery system described in this paper.}
+\caption{Overview of technologies used to maintain, store, and distribute the Versioned Data as described in this paper.}
 {\footnotesize
@@
   \textbf{Technology} & \textbf{Description} \\\hline
+  \textsc{API}   & An Application Programming Interface provides a set of protocols for exchanging information.\\
+%   \textsc{CRAN}  &  Open source repository of packages for the R language \\
+   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}  \\
+   \textsc{doi} & Digital object identifier, which refers a user to a single digital object \\
+%   \textsc{doi} ``minting'' & Allocation of a \textsc{doi} to a data product\\
    \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data\\
-   github.com & A commercial webplatform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
-  \textsc{API}   & An Application Programming Interface provides a set of protocols for exchanging information.\\
+   github.com & A commercial web platform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
    \texttt{R}     &  Open source statistical and data processing language \\
-   \textsc{CRAN}  &  Open source repository of packages for the R language \\
-   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}  \\
-   \textsc{doi} & Digital object identifier, which refers a user to a single stable digital object \\
-   \textsc{doi} ``minting'' & Allocation of a \textsc{doi} to a data product\\
    \hline
@@
 \centering
-\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements}
+\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements.}
 {\footnotesize
@@
     & & Clear path for users to become contributors \\
-  Users (code-based) & Build reproducible products using specific versions of a dataset & Scripted access to releases\\
+  Users (programmatic) & Build reproducible products using specific versions of a dataset & Programmatic access to releases\\
     & & Easy installation\\
@@
 \end{table}
+
+\newpage
+
+\begin{table}[h!]
+\centering
+\caption{Example datasets using the Versioned Data Delivery system described in this paper.}
+{\footnotesize
+\vspace{1cm}
+
+  \begin{tabular}{p{3.5cm}p{3cm}p{7cm}}
+  \hline
+   \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
+  \ghsmurl{dfalster/baad} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \citep{Falster-2015}.\\
+  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & Provides a taxonomic lookup table for land plants \citep{Pennell-2015a}.\\
+  \hline
+  \end{tabular}
+  }
+\label{tab:examples}
+\end{table}
+
+\newpage
 
@@
 
-\begin{table}[h!]
-\centering
-\caption{Example datasets using the Versioned Data Delivery system described in this paper.}
-{\footnotesize
-\vspace{1cm}
-
-  \begin{tabular}{p{3.5cm}p{3cm}p{7cm}}
-  \hline
-   \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
-  \ghsmurl{dfalster/baad/} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \citep{Falster-2015}.\\
-  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & \texttt{taxonlookup} provides a taxonomic lookup table for land plants \citep{Pennell-2015a}.\\
-  \hline
-  \end{tabular}
-  }
-\label{tab:examples}
-\end{table}
-
-\newpage
 \section{Figures}
@@
 
-\newpage
+\clearpage
 

### 2017-10-17 ms.tex
some edits throughout the text (also, i think we should avoid using an acronym for versioned data)

@@
 
-Many, perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time.
+Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time.
 Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions.
@@
 
-In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing Versioned Data (\textsc{vd}).
+In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing Versioned Data.
 Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer.
@@
 Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle.
-To make it easier for users to access \textsc{vd} via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.
+To make it easier for users to access Versioned Data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.
 
-Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that facilitates access to releases of any \textsc{vd} hosted on \texttt{GitHub} (Fig.
+Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that facilitates access to releases of any Versioned Data hosted on \texttt{GitHub} (Fig.
 \ref{fig:technology_stack}).
@@
 \begin{enumerate}
-  \item Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}.
+  \item Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub};
   \item Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
@@
 For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
-The R package \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}.
+The R package \texttt{baad.data} consists of only a few simple functions and associated help files, that were automatically generated with \texttt{datastorr}.
 For a user, accessing a version of the data is a simple as typing a single line of code (Fig.
@@
 \begin{enumerate}
-  \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored
-  \item the filename in the release that will contain the \textsc{vd}
+  \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored;
+  \item the filename in the release that will contain the Versioned Data;
   \item the function used to load this into \texttt{R}.
@@
 %work on this paragraph more
-We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact.
+We acknowledge that the issues we have identified in this article will not come as news to most readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact.
 In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions.
@@
 % Reword next senetnce
-However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system.
+But the design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} (or an entirely different service and technology) and accessing versioned data programmatically.
 The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
 
-Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many current platforms for storing and distributing versioned controlled data for low or no cost.
+Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.
 The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories.
@@
 The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.
-We argue that by linking together the tools described above, researchers can now produce, curate, and distribute \textsc{vd} much more easily and cheaply than in the past.
+We argue that by linking together the tools described above, researchers can now produce, curate, and distribute Versioned Data much more easily and cheaply than in the past.
 
@@
 For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
-There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
+There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
 % Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

### 2017-10-17 ms.tex
reworking the discussion a bit

@@
 %work on this paragraph more
-We acknowledge that the issues we have identified in this article will not come as news to most readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact.
+We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact.
 In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions.
@@
 % Moved from above, need to integrate better
-We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users.
-And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
+We have created \texttt{datastorr} to make such a \textsc{vdd} system easy to implement and manage for \texttt{R} users.
+And as we state above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
 % Reword next senetnce
-But the design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} (or an entirely different service and technology) and accessing versioned data programmatically.
+But the design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} --- or an entirely different service and technology --- and accessing versioned data programmatically.
 The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
 
-Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.
+%Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.
 The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories.
@@
 Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.
-Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
+
+ A central feature of this concept is that the versioned data is maintained in the cloud.
+This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.
 The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.
 We argue that by linking together the tools described above, researchers can now produce, curate, and distribute Versioned Data much more easily and cheaply than in the past.
+Additionally, one of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
+Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
+As a result adopting a \textsc{vdd} system has the added benefit of facilitating collaboration among research groups.
+
+ While our first attempt \textsc{vdd} system, \texttt{datastorr} is currently being used, there remain several challenges that we anticipate will be overcome as both technology and scientific norms change.
+For example, because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.
+For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
+There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
+Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
+Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.
+
+% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future
 
@@
 But, nonetheless, workable systems are already possible.
-
-Additionally, one of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
-Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
-As a result adopting a \textsc{vdd} system has the added benefit of facilitating collaboration among research groups.
 
@@
 % trick is to say something about how things are getting better rapidly, but not scare people away from doing this now.  
-\subsection{Future directions}
+%\subsection{Future directions}
 
-Technology for data management and distribution is moving forward rapidly.
+%Technology for data management and distribution is moving forward rapidly.
 The workflow we describe above works well with current tools and should work even more quickly with its future replacement.
@@
 
-Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.
+%Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.
 For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
-There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
+%There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
 % Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future
 
-Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
+%Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
 Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.

### 2017-10-17 ms.tex
commenting out table 5 as it is not referenced in text - we already have plenty of tables as it is

@@
 
-\begin{table}[h!]
-\centering
-\caption{Alternative platforms for minting \textsc{doi}s from \texttt{GitHub} repositories.}
-{\footnotesize
-\vspace{1cm}
-  \begin{tabular}{p{4cm}p{8cm}}
-  \hline
-  \textbf{Provider} & \textbf{Features} \\ \hline
-  \texttt{Zenodo}: \smurl{zenodo.org} & Automated \textsc{doi} for each \texttt{GitHub} release \\
-    & Funded and maintained by CERN and OpenAIRE for long term preservation \\
-    & At present, can only archive files embedded directly in the \texttt{git} repo, not produced by it as outputs\\
-  \texttt{FigShare}: \smurl{figshare.com} & Builds data package based on \texttt{GitHub} release \\
-    & Allows you to upload additional files, e.g. datasets built by code \\
-    & Integrated with several journals, e.g. Nature family, \textsc{PNAS} \\
-  \hline
-  \end{tabular}
-  }
-\label{tab:doi_minting}
-\end{table}
+%\begin{table}[h!]
+%\centering
+%\caption{Alternative platforms for minting \textsc{doi}s from \texttt{GitHub} repositories.}
+%{\footnotesize
+%\vspace{1cm}
+%  \begin{tabular}{p{4cm}p{8cm}}
+%  \hline
+%  \textbf{Provider} & \textbf{Features} \\ \hline
+%  \texttt{Zenodo}: \smurl{zenodo.org} & Automated \textsc{doi} for each \texttt{GitHub} release \\
+%    & Funded and maintained by CERN and OpenAIRE for long term preservation \\
+%    & At present, can only archive files embedded directly in the \texttt{git} repo, not produced by it as outputs\\
+%  \texttt{FigShare}: \smurl{figshare.com} & Builds data package based on \texttt{GitHub} release \\
+%    & Allows you to upload additional files, e.g. datasets built by code \\
+%    & Integrated with several journals, e.g. Nature family, \textsc{PNAS} \\
+%  \hline
+%  \end{tabular}
+%  }
+%\label{tab:doi_minting}
+%\end{table}
 
-\newpage
+%\newpage
 

### 2017-11-09 ms.tex
Edits throughout, messup of discussion

@@
 \newcommand{\datastorr}{\texttt{datastorr}}
-\newcommand{\smurl}[1]{{\footnotesize\url{#1}}}
+\newcommand{\smurl}[1]{{\footnotesize\href{https://#1}{#1}}}
 \newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{#1}}}
@@
 $^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
-University of New South Wales, Sydney, NSW 2052, Australia\\
+University of New South Wales, Sydney NSW 2052, Australia\\
 $^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom\\
 $^3$ Department of Zoology and Biodiversity Research Centre,\\
-University of British Columbia, Vancouver, B.C. V6T 1Z4 Canada\\
+University of British Columbia, Vancouver B.C. V6T 1Z4, Canada\\
 $^*$ Corresponding author: daniel.falster@unsw.edu.au\\
@@
 The sharing and re-use of data has become a cornerstone of modern science.
-To help this process, numerous linked tools have been developed to allow quick and easy data sharing:    a data file for a particular analysis can be quickly uploaded, given a digital identifier, and shared.
-Thus far, however, this framework does not accommodate on-going scientific progress: for many important problems, particularly those suitable for meta-analyses, datasets will continue to grow with time --- more data points will be added, and new and more useful data structures will be created.
-Essentially this means that datasets, like science in general, improve with time.
-We suggest that many different datasets would be more useful if organized as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.
-In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data curators and users of even small-to-medium sized datasets.
-We introduce a system called Versioned Data Delivery (\textsc{vdd}) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
+To help this process, multiple tools allow quick and easy data sharing: a data file can be quickly uploaded, given a digital identifier, and shared.
+So far, however, data publishing models have not accommodated on-going scientific improvements in data: for many important problems, datasets continue to grow with time -- more data points will be added, and new and more useful data structures will be created.
+In other words, datasets, like scientific knowledge in general, advance with time.
+We therefore suggest that many datasets would be more usefully published as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.
+In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data maintainers and users of even small-to-medium sized datasets.
+We also introduce a system called Versioned Data Delivery (\textsc{vdd}) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
 These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.
@@
 Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \citep{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
-In particular, the current data publishing model has not yet embraced the idea that many  datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.
+In particular, the current model for publishing data has not yet embraced the idea that many datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.
 These datasets are constantly evolving, i.e.
@@
 
-Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time.
-Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions.
+Many of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time.
+Typical changes include: the adding new data, improving the quality of existing data, integrating with other datasets, or re-structuring the dataset content to be able to address new questions.
 For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
 As research around a data product grows, there might be many such additions.
-Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure reproducibility.
+Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure previous works remain reproducible.
 
@@
 Many research groups solve the issue of versioning data internally in a variety of ad hoc and suboptimal ways.
-A common solution is to email around the latest versions with version numbers appended to the filename.
+A common solution is to email around the latest versions with version numbers or dates appended to the filename.
 Another approach is to repeat corrections or updates that have been made elsewhere.
@@
 
-In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing Versioned Data.
+In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
 Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer.
 With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer.
-In either the development of software or data, successive versions are released over time (called ``releases'').
+In either the development of software or data, successive versions (called ``releases'') are distributed over time.
 The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
@@
 Of particular interest for current purposes, is \texttt{GitHub}'s ability to host a stream of ``releases'' from the dataset, alongside the \texttt{git} repository containing all the rawfiles.
-Each ``release'' is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset \emph{Maintainer} decided to generate a new  version of the data for distribution.
-While \emph{users} could in principle download the entire \texttt{git} repository, actually all they want are these 'releases'.
+Each release is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset \emph{maintainer} decided to generate a new version of the data for distribution.
+While \emph{users} could in principle download the entire \texttt{git} repository, actually all they want are the releases.
 
-Deciding when to make a new release is at the discretion of the dataset \emph{Maintainer}.
+Deciding when to make a new release is at the discretion of the dataset \emph{maintainer}.
 In practice, one makes fewer releases than one does commits into the \texttt{git} repository, though there is nothing stopping \emph{maintainers} from releasing a new version for every commit.
-The flexibility here allows \emph{maintainers} to do internal work between releases and only \emph{release} the data to users when the revision represents a clear improvement on the previous release.
+The flexibility here allows \emph{maintainers} to do internal work between releases and only release the data to users when the revision represents a clear improvement on the previous release.
 
 Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data users to access the data: those that interact with the data via point and click downloading and those that use programmatic interaction (Fig.
-\ref{fig:technology_stack} and Table \ref{tab:user_requirements}).
+\ref{fig:technology_stack}, Table \ref{tab:user_requirements}).
 Specifically, \texttt{GitHub} releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.
 
-\subsection{`Semantic' versioning}
+\subsection{Semantic versioning}
 
@@
 
-Specifically, we suggest applying the theory of ``semantic versioning'', developed for software distribution (see \href{http://semver.org/}{semver.org}), to successive releases of a dataset.
+Specifically, we suggest applying the theory of ``semantic versioning'', developed for software distribution (\href{http://semver.org/}{semver.org}), to successive releases of a dataset.
 Semantic versioning uses tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers.
@@
 \begin{itemize}
-  \item  {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a {\bf correction} or ``patch'', for example a small error correction which is unlikely to ``break'', or change in a substantial way, a users' analyses, although there may be minor changes in the results.
-  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a {\bf minor} enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the database curator it might change the results of some users' analyses.
-  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a very {\bf major} revision, for example improving the entire structure of the dataset and adding new columns.
-These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the database or both.
+  \item {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a ``correction'' (or ``patch''), for example a small error correction which is unlikely to break, or change in a substantial way, a users' analyses, although there may be minor changes in the results.
+  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor'' enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset curator it might change the results of an analysis done with the prior version and so warrants closer inspection.
+  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a very ``major'' revision, for example improving the entire structure of the dataset and adding new columns.
+These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the dataset or both.
 \end{itemize}
 Any user of a dataset where releases were tagged using semantic versioning, would immediately know the types of change that might have occurred when requesting different versions and set their there expectations accordingly.
-
-% I think that major version increments could still be looked at as
-% _breaking_ changes as often as large additions of data.  Changing
-% column names, removing columns, changing the structure of linked
-% tables etc.
 
@@
 Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle.
-To make it easier for users to access Versioned Data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.
+To make it easier for users to access versioned data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.
 
-Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that facilitates access to releases of any Versioned Data hosted on \texttt{GitHub} (Fig.
+Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), that facilitates access to releases of any versioned data hosted on \texttt{GitHub} (Fig.
 \ref{fig:technology_stack}).
@@
   \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored;
-  \item the filename in the release that will contain the Versioned Data;
+  \item the filename in the release that will contain the versioned data;
   \item the function used to load this into \texttt{R}.
@@
 \section{Discussion}
-
 % General problem. An example of solution. Key features
 
-% The introduction talks at length about medium sized data and little about versioning but this section talks only about versioning
-% WKC: now intro is about versioning and data size issue discussion moved here, which I think is an improvement
+The main issue we have identified in this article will not be new for many readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about how to handle dataset updates.
+Some specific solutions for versioning do already exist, such as with genetic sequence data on \texttt{GenBank} (see \smurl{www.ncbi.nlm.nih.gov/genbank/sequenceids/}).
+But so far, versioning solutions have only been developed for large dynamic databases managed by well-funded institutions or consortia and involving simple standardised data types.
+While individual research groups working with a specialized type of data could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required (Table \ref{tab:publishing_models}) will be discouraging for most groups.
+This suggests there is a substantial need for an easy, cheap, and scalable solution for serving versioned data.
+By adopting best practices from software engineering, we believe a workable system now exists.
+Moreover, we created the \texttt{datastorr} package to make such a \textsc{vdd} system easy to implement for \texttt{R} users.
+As it builds off established and open source software and data science platforms, the proposed system is already easy to deploy on relatively large scale.
 
-% Discuss Genbank or other large sites?
-% The issues in practically doing the versioning are more acute when the
-% data are bigger than small, that's all.
+The proposed \textsc{vdd} system breaks from current norms for scientific data publishing in that it builds directly off hosts supporting \texttt{git} repositories, specifically that run by the commercial company \texttt{GitHub}.
 
-% From intro:
-% When considering the right tool to store and deliver versioned data, the size
-% of the dataset is another essential consideration. ``Big data'' brings with it
-% specific needs and so requires a specific set of hardware and software tools.
-% Examples of big data include genetic datasets, global datasets of species
-% observations, and remote sensing data. This size of this data is bigger than
-% that which individual research groups can maintain and often requires a
-% specialized governmental or non-governmental institution with full-time staff
-% to curate and maintain those research resources.
-
-%work on this paragraph more
-We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact.
-In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions.
-But so far these as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences).
-But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter.
-As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.
-
-% Moved from above, need to integrate better
-We have created \texttt{datastorr} to make such a \textsc{vdd} system easy to implement and manage for \texttt{R} users.
-And as we state above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
-% Reword next senetnce
-But the design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} --- or an entirely different service and technology --- and accessing versioned data programmatically.
+The design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} --- or an entirely different service and technology --- and accessing versioned data programmatically.
 The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
 
-%Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.
+And as we state above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
+
+GitHub - commercial
+
+Integrate with DOIs
+
+Future technology changes
+
+(ideas) Also the OKFN data package idea http://data.okfn.org might be worth merging in here, especially as we can fill most of the required data fields from the DESCRIPTION file.
+
+(Matt says drop) Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.
 The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories.
@@
 The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.
-We argue that by linking together the tools described above, researchers can now produce, curate, and distribute Versioned Data much more easily and cheaply than in the past.
+We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.
 Additionally, one of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
@@
 
-% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future
+(idea) Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future
 
-%The key features of a \textsc{vdd} system that are already possible are a versioning system to communicate small and large changes to users, both point-and-click and programmatic access to all versions, reliable distribution to many users, and stable, organized \textsc{doi}s (see Fig.
-\ref{fig:technology_stack}).
-Moreover, it is relatively easy to update and add new versions.
-Technology is moving forward rapidly and so it is very likely that the particular tools we describe above will add features and be even easier to use.
-But, nonetheless, workable systems are already possible.
+Something about DOI: easy to integrate.
 
-%Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
+(old) Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
 At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}).
@@
 
-% subsection only temporary as not allowed by journal guidelines.
-% trick is to say something about how things are getting better rapidly, but not scare people away from doing this now.  
-%\subsection{Future directions}
-
-%Technology for data management and distribution is moving forward rapidly.
-The workflow we describe above works well with current tools and should work even more quickly with its future replacement.
-Here, we highlight a few rapidly developing future tools that researchers should be aware of.
-
-%Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.
-For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
-%There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
-% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future
-
-%Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
-Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.
-
 % We have constant drama at work because the different spreadsheet
@@
 One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
+
  % This section conflates git and github more than is useful. It
  % might be worth setting up github as the explicit platform here or
- % dialing back the language a bit.  GitLab and bitbucket have their
- % own "merge request" feature (I think).  It's also github that makes
- % repositories _discoverable_ which is a requirement for getting
- % outside collaborators in.  Oddly the decentralised nature of git is
- % not actually that useful in the github model because github is so
- % strongly centralised.  Will and I did some true decentralised git
- % collaboration on a plane to nescent once with a bare git repo on a
- % thumb drive.
+ % dialing back the language a bit.
 
-% So, git doesn't actually work this way, though everyone seems to
-% think that it does.  Git stores *content* and then later tries to
-% de-duplicate it.  It then displays changes as line-by-line diffs
-% (but you can change this to display the changes other ways; see
-% --color-words --ignore-all-space for example.  The problem is that
-% the algorithms that git uses are optimised for typical sizes of code
-% and not for data.  Small data is actually totally fine to store with
-% git.  The other issue that turns up is if the files are _binary_
-% then an arbitrarily small change in the underlying data can result
-% in an arbitrarily large change in the binary representation, which
-% prevents deduplication.
-%
-% Git-LFS tries to step around this by storing large files separately
+%% Git-LFS tries to step around this by storing large files separately
 % from git's general machinery. Git-annex did the same. But they
@@
 
-% Karthik has started with rdrop2 again I think; I do think that is a reasonable low-entry way of distributing things
-
-% Dan Noble: I think this is a pragmatic view, but I would have a tendency not to discuss how the tools you can use for a VDD approach are likely to be "replaced in the near future”.
-Of course, things change, but this will not entice the people you want the paper to appeal to to adopt this sort of approach.
-From my experience, some of the resistance to adopting this stuff comes from programs or repositories being “replaced” at such a high rate.
-They just don’t feel like its worth doing.
-
 \section{Acknowledgements}
@@
 \centering
-\caption{Alternative frameworks for setting up and managing a living database}
+\caption{Alternative frameworks for setting up and managing a living dataset}
 {\footnotesize
@@
   \hline
-  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web database} & \textbf{Versioned data delivery}\\
+  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web database} & \textbf{Versioned Data Delivery}\\
   \textbf{} & (e.g. \smurl{datadryad.org})& (e.g. \smurl{coraltraits.org}) & (eg. \texttt{datastorr} via \texttt{GitHub})\\
@@
 \centering
-\caption{Overview of technologies used to maintain, store, and distribute the Versioned Data as described in this paper.}
+\caption{Overview of technologies used to maintain, store, and distribute the versioned data as described in this paper.}
 {\footnotesize
 \vspace{1cm}
-  \begin{tabular}{p{5cm}p{10cm}}
+  \begin{tabular}{p{3cm}p{12cm}}
   \hline
@@
   \textsc{API} & An Application Programming Interface provides a set of protocols for exchanging information.\\
-%   \textsc{CRAN}  &  Open source repository of packages for the R language \\
-   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}  \\
-   \textsc{doi} & Digital object identifier, which refers a user to a single digital object \\
-%   \textsc{doi} ``minting'' & Allocation of a \textsc{doi} to a data product\\
-   \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data\\
+   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}. \\
+   \textsc{doi} & Digital object identifier, which refers a user to a single published object. \\
+   \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data.\\
    github.com & A commercial web platform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
-   \texttt{R}     &  Open source statistical and data processing language \\
+   \texttt{R} & Open source statistical and data processing language.\\
    \hline
@@
     & & Clear path for users to become contributors \\
-  Users (programmatic) & Build reproducible products using specific versions of a dataset & Programmatic access to releases\\
-    & & Easy installation\\
+  Users (programmatic) & Build reproducible products using specific versions of a dataset & Programmatic access to releases \\
+    & & Easy installation \\
     & & Long term stability \\
@@
 \end{table}
-
-\newpage
 

### 2017-11-09 ms.tex
final two paragraphs

@@
 While our first attempt \textsc{vdd} system, \texttt{datastorr} is currently being used, there remain several challenges that we anticipate will be overcome as both technology and scientific norms change.
-For example, because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to data \citep{Perkel-2016}.
-For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
-There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
 Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
@@
 % solution to the sorts of issues that we have here.
+
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}
+
+The design choices we made in datastorr represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with versioned data stored in the cloud.
+The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
+Indeed, as with every technology the best available approach is certain to change: there are currently a number of promising technologies---CKAN, rdrop2 and Git-LFS---that may facilitate even better \textsc{vdd} systems in the future.
+There is also rapid current work on new implementations which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
+
+In short, many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data that are cheap and relatively easy to set up.
 

### 2017-11-09 ms.tex
Refine discussion

@@
 \section{Discussion}
-% General problem. An example of solution. Key features
 
@@
 
-The proposed \textsc{vdd} system breaks from current norms for scientific data publishing in that it builds directly off hosts supporting \texttt{git} repositories, specifically that run by the commercial company \texttt{GitHub}.
+A central feature of the proposed  \textsc{vdd} system is that data are maintained in the cloud.
+This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig.
+\ref{fig:technology_stack}).
+Cloud systems thus act as a central point for the collection, curation, and distribution of the data.
+Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013, Perkel-2016}.
+Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
+Adopting a \textsc{vdd} system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.
 
-The design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} --- or an entirely different service and technology --- and accessing versioned data programmatically.
-The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
-
-And as we state above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
-
-GitHub - commercial
-
-Integrate with DOIs
-
-Future technology changes
-
-(ideas) Also the OKFN data package idea http://data.okfn.org might be worth merging in here, especially as we can fill most of the required data fields from the DESCRIPTION file.
-
-(Matt says drop) Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.
-The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories.
-In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples).
-Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}.
-
-A central feature of this concept is that the versioned data is maintained in the cloud.
-This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for the collection, curation, and distribution of the data.
-The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.
-We argue that by linking together the tools described above, researchers can now produce, curate, and distribute versioned data much more easily and cheaply than in the past.
-Additionally, one of the greatest benefits of using a cloud-based tool for development of open-source software has been the way it encourages contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013}.
-Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
-As a result adopting a \textsc{vdd} system has the added benefit of facilitating collaboration among research groups.
-
-While our first attempt \textsc{vdd} system, \texttt{datastorr} is currently being used, there remain several challenges that we anticipate will be overcome as both technology and scientific norms change.
-Another area requiring attention is properly allocating credit to authors for these citations, and we hope that the large citation mapping services (\texttt{Web of Science} and \texttt{Google Scholar}) will adapt to the community standard once the research community settles on best archiving practices for versioned data.
-Likewise, we are optimistic that our funding bodies, research institutions, and scientific colleagues will further recognize the value in adding to, improving, and curating existing datasets in addition to collecting new ones.
-
-(idea) Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future
-
-Something about DOI: easy to integrate.
-
-(old) Systematically archiving data is a key part of good workflow \citep{Wilkinson-2016, Piwowar-2011, Whitlock-2011}, and tools for creating and archiving digital objects have seen rapid progress in recent years.
-At present, each version of the dataset can be automatically archived and assigned a \textsc{doi}, via several providers (Table \ref{tab:doi_minting}).
-Here we describe the particular integration of \texttt{GitHub} and \texttt{Zenodo}.
-It is useful to have the ability to either be specific---reference a particular version for the purposes of reproducibility or general--reference the entire project.
-\texttt{Zenodo} currently automatically supplies both ``concept'' and ``version'' \textsc{doi}'s to the commits associated with \texttt{GitHub} releases automatically \citep[Fig.
-\ref{fig:semantic} and ][]{Nielsen-2017}.
-This allows users to very precisely reference either generally or specifically, both of which may be appropriate in different situations.
-
-% I'm unclear on the precice benefits of dois here (and actually in
-% general to a point); some further clarification might be useful. So
-% far as I know they are useful because (a) journal urls are terrible
-% and inconsistent (b) because everyone has them. I don't really know
-% why else you'd need one. Especially in our case we offer no promise
-% of persistence (deleting the github repo would break the doi and we
-% could update the repo to include different data trivially). So what
-% is being added? Our retrival system does not use the doi at all!
-
-% WKC: hmmmm not sure we've nailed this yet, see
-% https://github.com/traitecoevo/data_versioning/issues/10#issuecomment-317059819
-
-% We have constant drama at work because the different spreadsheet
-% writers we use across the platforms quote files inconsistently so
-% every save is a complete rewrite of the file. It's totally
-% impossible to do any collaborative editing....
-% For open-source projects, external parties are also welcome to copy (in \texttt{git} speak ``clone'' or ``fork'') a repository, make their own modifications and send these back to the project \emph{maintainers} to be reviewed and potentially included in the project.
-This process is called submitting a ``Pull request''.
-One of the potential advantages of moving the workflow for managing data onto \texttt{GitHub} is the potential for seamless collaboration and for receiving pull requests from contributors \citep{Perkel-2016}.
-
- % This section conflates git and github more than is useful. It
- % might be worth setting up github as the explicit platform here or
- % dialing back the language a bit.
-
-%% Git-LFS tries to step around this by storing large files separately
-% from git's general machinery. Git-annex did the same. But they
-% suffer from some of the other issues with git which are that it's a
-% bit of a pain for (especially novice) users to do nontrivial things
-% with. I _always_ have to look up how to compare two versions across
-% branches. And there's no simple way of loading in two versions of
-% the data into R without ending up with two working copies or doing
-% some nasty "git show" call.
-
-% I really think that something like CKAN (though it doesn't support
-% holding data if I remember correctly) could totally be a complete
-% solution to the sorts of issues that we have here.
-
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{0.4pt}}
+An important concern for any data delivery system is the stability and reliability of the system.
+In the short term, users want minimal downtime, high speed, and seamless operation.
+As one of the largest companies hosting computer source code, \texttt{GitHub} provides exceptional performance in this regard -- certainly as good or better than nearly any system scientists might build themselves.
+In the long term, scientists desire long-term preservation of datasets, software, and published papers.
+To enhance long-term stability of data-versions released on \texttt{GitHub}, users can also choose to automatically archive every version in one of several traditional data archives, with a \textsc{doi} (Digital Object Identifier) minted for each release.
+Currently, both \texttt{Zenodo} (\smurl{zenodo.org}) and  \texttt{FigShare} \smurl{figshare.com} each integrate with \texttt{GitHub} for archiving of material hosted there.
 
@@
 The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
-Indeed, as with every technology the best available approach is certain to change: there are currently a number of promising technologies---CKAN, rdrop2 and Git-LFS---that may facilitate even better \textsc{vdd} systems in the future.
+Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies---such as CKAN, OKFN or Git-LFS---facilitate even better \textsc{vdd} systems in the future.
 There is also rapid current work on new implementations which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
 
-In short, many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data that are cheap and relatively easy to set up.
+In short, many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system.
+By adopting best practices from software engineering, we believe a relatively easy, cheap, and scalable solution for serving versioned data now exists.
 
@@
 
-%\begin{table}[h!]
-%\centering
-%\caption{Alternative platforms for minting \textsc{doi}s from \texttt{GitHub} repositories.}
-%{\footnotesize
-%\vspace{1cm}
-%  \begin{tabular}{p{4cm}p{8cm}}
-%  \hline
-%  \textbf{Provider} & \textbf{Features} \\ \hline
-%  \texttt{Zenodo}: \smurl{zenodo.org} & Automated \textsc{doi} for each \texttt{GitHub} release \\
-%    & Funded and maintained by CERN and OpenAIRE for long term preservation \\
-%    & At present, can only archive files embedded directly in the \texttt{git} repo, not produced by it as outputs\\
-%  \texttt{FigShare}: \smurl{figshare.com} & Builds data package based on \texttt{GitHub} release \\
-%    & Allows you to upload additional files, e.g. datasets built by code \\
-%    & Integrated with several journals, e.g. Nature family, \textsc{PNAS} \\
-%  \hline
-%  \end{tabular}
-%  }
-%\label{tab:doi_minting}
-%\end{table}
-
-%\newpage
-
 \section{Figures}

### 2017-11-09 ms.tex
edits to the discussion

@@
 
-The main issue we have identified in this article will not be new for many readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about how to handle dataset updates.
-Some specific solutions for versioning do already exist, such as with genetic sequence data on \texttt{GenBank} (see \smurl{www.ncbi.nlm.nih.gov/genbank/sequenceids/}).
+The main issue we have identified in this article will not come as news to many readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about how to handle dataset updates.
+Some specific solutions for versioning already exist, such as those designed for genetic sequence data (e.g., \texttt{GenBank}; see \smurl{www.ncbi.nlm.nih.gov/genbank/sequenceids/}).
 But so far, versioning solutions have only been developed for large dynamic databases managed by well-funded institutions or consortia and involving simple standardised data types.
-While individual research groups working with a specialized type of data could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required (Table \ref{tab:publishing_models}) will be discouraging for most groups.
+While individual research groups working with a specialized type of data could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required (Table \ref{tab:publishing_models}) will be discouraging for most.
 This suggests there is a substantial need for an easy, cheap, and scalable solution for serving versioned data.
@@
 Cloud systems thus act as a central point for the collection, curation, and distribution of the data.
-Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously, and even individuals from outside the initial group of project participants \citep{Rogers-2013, Perkel-2016}.
+Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \citep{Rogers-2013, Perkel-2016}.
 Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
@@
 As one of the largest companies hosting computer source code, \texttt{GitHub} provides exceptional performance in this regard -- certainly as good or better than nearly any system scientists might build themselves.
-In the long term, scientists desire long-term preservation of datasets, software, and published papers.
+In the long term, scientists want their datasets, software, and papers to preserved and remain accessible.
 To enhance long-term stability of data-versions released on \texttt{GitHub}, users can also choose to automatically archive every version in one of several traditional data archives, with a \textsc{doi} (Digital Object Identifier) minted for each release.
@@
 The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
-Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies---such as CKAN, OKFN or Git-LFS---facilitate even better \textsc{vdd} systems in the future.
-There is also rapid current work on new implementations which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
+Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies --- such as CKAN, OKFN or Git-LFS --- facilitate even better \textsc{vdd} systems in the future.
+There are some really exciting ideas that are currently under development which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep[e.g.,][]{Fli, Dat}.
 

### 2017-11-09 ms.tex
Format for Scientific Data

@@
-\documentclass[a4paper,11pt]{article}
-\usepackage[osf]{mathpazo}
-\usepackage{ms}
-\usepackage[]{natbib}
-\raggedright
+\documentclass[english]{article}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage{babel}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{fancyhdr}
+\usepackage[colorlinks=false, allcolors=grey, breaklinks=true]{hyperref}
+
+\RequirePackage[superscript,biblabel,nomove]{cite}
+\bibliographystyle{assets/naturemag-doi}
+\RequirePackage{assets/jabbrv}
+
+%\bibliographystyle{naturemag}
+\newcommand{\scidatalogo}{\includegraphics[height=36pt]{assets/SciData_logo.jpg}}
+\pagestyle{fancy}
+\fancyhf{}
+\renewcommand{\headrulewidth}{0pt}
+\setlength{\headheight}{40pt} 
+\lhead{\textsc{\scidatalogo}}
+\rhead{Falster \emph{et al.} -- Versioned Data Delivery}
 
 \newcommand{\datastorr}{\texttt{datastorr}}
-\newcommand{\smurl}[1]{{\footnotesize\href{https://#1}{#1}}}
-\newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{#1}}}
+\newcommand{\smurl}[1]{{\footnotesize\href{https://#1}{\nolinkurl{#1}}}}
+\newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{\nolinkurl{#1}}}}
 
-\usepackage{graphicx}
+\begin{document}
 
 \title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
-\author{Daniel S. Falster$^{1,*}$, Richard G. FitzJohn$^2$,\\ Matthew
-  W. Pennell$^3$, and William K. Cornwell$^{1}$}
-\affiliation{
+
+\author{Daniel S. Falster\textsuperscript{1,*}, Richard G. FitzJohn\textsuperscript{2},\\ Matthew
+  W. Pennell\textsuperscript{3}, and William K. Cornwell\textsuperscript{1}}
+
+\maketitle
+\thispagestyle{fancy}
 $^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
-University of New South Wales, Sydney NSW 2052, Australia\\
-$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom\\
-$^3$ Department of Zoology and Biodiversity Research Centre,\\
-University of British Columbia, Vancouver B.C. V6T 1Z4, Canada\\
-$^*$ Corresponding author: daniel.falster@unsw.edu.au\\
-}
-\date{}
+University of New South Wales, Sydney NSW 2052, Australia; 
+$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom; 
+$^3$ Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver B.C. V6T 1Z4, Canada;
+$^*$ Corresponding author: Daniel Falster (daniel.falster@unsw.edu.au)
 
-\bibliographystyle{mee}
+\begin{abstract}
+The sharing and re-use of data has become a cornerstone of modern science.
+Multiple platforms now allow quick and easy data sharing.
+So far, however, data publishing models have not accommodated on-going scientific improvements in data: for many problems, datasets continue to grow with time -- more records are added, errors fixed, and new data structures are created.
+In other words, datasets, like scientific knowledge, advance with time.
+We therefore suggest that many datasets would be usefully published as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.
+In this article, we argue for adopting the paradigm and processes for versioned data, analogous to software versioning.
+We also introduce a system called Versioned Data Delivery and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
+These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.
+\end{abstract}
 
-\usepackage[title,titletoc,toc]{appendix}
+%\keywords{Data sharing, Version control, Semantic versioning, Meta-analysis}
 
-\mstype{Article Note}
-\runninghead{Versioned Data Delivery}
-\keywords{Data sharing, Version control, Semantic versioning, Meta-analysis}
+\section*{Background}
 
-\begin{document}
-\mstitlepage
-\noindent
-\parindent=1.5em
-\addtolength{\parskip}{.3em}
-\doublespacing
-\linenumbers
-
-\section{Summary}
-
-The sharing and re-use of data has become a cornerstone of modern science.
-To help this process, multiple tools allow quick and easy data sharing: a data file can be quickly uploaded, given a digital identifier, and shared.
-So far, however, data publishing models have not accommodated on-going scientific improvements in data: for many important problems, datasets continue to grow with time -- more data points will be added, and new and more useful data structures will be created.
-In other words, datasets, like scientific knowledge in general, advance with time.
-We therefore suggest that many datasets would be more usefully published as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.
-In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data maintainers and users of even small-to-medium sized datasets.
-We also introduce a system called Versioned Data Delivery (\textsc{vdd}) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
-These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.
-
-\section{Introduction}
-
-As is evidenced in the advent of this journal \citep{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product.
-Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \citep[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
+As is evidenced in the advent of this journal \cite{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product.
+Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
 Datasets are now routinely archived as part of the publishing process.
 Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals.
-Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \citep{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
+Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
 In particular, the current model for publishing data has not yet embraced the idea that many datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.
 These datasets are constantly evolving, i.e.
-they are never "finished", and may be the subject of repeated analysis or meta-analysis.
+they are never ``finished'', and may be the subject of repeated analysis or meta-analysis.
 
@@
 
-In this article we
-\begin{enumerate}
-  \item Introduce the concept of Versioned Data Delivery (\textsc{vdd}).
-  \item Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets.
-  \item Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
-\end{enumerate}
-The issue of updating and expanding published data has already been addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g.
+In this article we 1) Introduce the concept of Versioned Data Delivery (\textsc{vdd}); 2) Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and 3) Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
+The issue of updating and expanding published data has already been addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data, where new data can be added and there exist abilities to correct errors in existing records (e.g.
 by adding multiple version of a genetic sequence).
@@
 
-\section{A lightweight, cheap, and scalable workflow for delivering versioned data}
+\section*{A lightweight, cheap, and scalable workflow for delivering versioned data}
 
-In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
+In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
 Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer.
@@
 
-\subsection{Version control}
+\subsection*{Version control}
 
@@
 In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes.
-Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016, Lowndes-2017}.
+Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \cite{Ram-2013, Perkel-2016, Lowndes-2017}.
 \texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g.
@@
 
-\subsection{Hosting and distributing versioned data}
+\subsection*{Hosting and distributing versioned data}
 
-Datasets stored under version control via \texttt{git} reach their real potential when hosted at a suitable internet hosting service \citep{Ram-2013,Perkel-2016}.
+Datasets stored under version control via \texttt{git} reach their real potential when hosted at a suitable internet hosting service \cite{Ram-2013,Perkel-2016}.
 Here we focus on the platform \texttt{GitHub} (Table \ref{tab:technologies}), but similar functions could be achieved via other providers such as \href{http://bitbucket.org}{bitbucket.org} and \href{http://gitlab.com}{gitlab.com}.
@@
 
-\subsection{Semantic versioning}
+\subsection*{Semantic versioning}
 
@@
 
-\subsection{\texttt{datastorr} and dataset-specific \texttt{R} packages}
+\subsection*{\texttt{datastorr} and dataset-specific \texttt{R} packages}
 
@@
 Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle.
-To make it easier for users to access versioned data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.
+To make it easier for users to access versioned data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \cite{R-2017}.
 
@@
 \ref{fig:technology_stack}).
-Specifically, the \texttt{datastorr} package:
-\begin{enumerate}
-  \item Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub};
-  \item Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
-\end{enumerate}
+Specifically, the \texttt{datastorr} package: 1)Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}; and 2) Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
 Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.
 
-For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
+For example, \texttt{datastorr} has been used to build several packages (Table \ref{tab:examples}), including \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
 The R package \texttt{baad.data} consists of only a few simple functions and associated help files, that were automatically generated with \texttt{datastorr}.
@@
 Accessing a different version of the data involves changing only the version number.
-From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \citep[e.g.][]{Duursma-2016,Falster-2016}.
+From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \cite{Duursma-2016,Falster-2016}.
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 Cloud systems thus act as a central point for the collection, curation, and distribution of the data.
-Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \citep{Rogers-2013, Perkel-2016}.
+Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}.
 Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
@@
 Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies --- such as CKAN, OKFN or Git-LFS --- facilitate even better \textsc{vdd} systems in the future.
-There are some really exciting ideas that are currently under development which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep[e.g.,][]{Fli, Dat}.
+There are some really exciting ideas that are currently under development which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \cite{Fli, Dat}.
 
@@
 
-\section{Acknowledgements}
+\section*{Acknowledgements}
 We thank D Noble for comments on an earlier draft and C Boettiger for helpful discussions. DSF was funded by the Australian Research Council. MWP was funded by a NSERC Discovery Grant.
+
+RF developed the datastorr package. All authors discussed the concepts and wrote the paper.
 
@@
 
-\section{Tables}
+\section*{Tables}
 
@@
 \centering
-\caption{Alternative frameworks for setting up and managing a living dataset}
+\caption{Alternative frameworks for distributing data.}
 {\footnotesize
 \vspace{1cm}
-  \begin{tabular}{p{2.5cm}p{3.5cm}p{3.5cm}p{4cm}}
+  \begin{tabular}{p{2.5cm}p{3cm}p{3cm}p{4cm}}
   \hline
-  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web database} & \textbf{Versioned Data Delivery}\\
-  \textbf{} & (e.g. \smurl{datadryad.org})& (e.g. \smurl{coraltraits.org}) & (eg. \texttt{datastorr} via \texttt{GitHub})\\
-  \hline
-   Webserver        & bespoke & bespoke &  \smurl{github.com}\\
-   Backend          & none & SQL + Ruby-on-rails 			& \texttt{git} + \texttt{datastorr} \\
+  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web databases} & \textbf{Versioned Data Delivery}\\
+   Platform (e.g.) 	& datadryad.org 			& coraltraits.org	  &  github.com\\
    User access      & Web browser 				& Web browser 			& Web browser or R \\
@@
    Bandwidth        & Managed by provider & Pro rata 					& Managed by \texttt{GitHub}\\
-   Maintainer skills & None & Ruby + php 					& R + \texttt{git} \\
-   User skills      &Web browsing& Web browsing  					& basic R \\
-   Versioning       &None& Hard 							& Easy \\
+   Maintainer skills & None 							& High							& R + \texttt{git} \\
+   User skills      & Web browsing 	& Web browsing  					& Web browsing or R \\
+   Versioning       & None 					& Hard 										& Easy \\
    \textsc{doi} minting  &Automatic & Manual 									& Automatic \\
@@
 \vspace{1cm}
-  \begin{tabular}{p{3cm}p{12cm}}
+  \begin{tabular}{p{3cm}p{9cm}}
   \hline
@@
 \vspace{1cm}
-  \begin{tabular}{p{2cm}p{5cm}p{7cm}}
+  \begin{tabular}{p{2cm}p{4cm}p{6cm}}
   \hline
@@
 
-  \begin{tabular}{p{3.5cm}p{3cm}p{7cm}}
+  \begin{tabular}{p{3.5cm}p{2cm}p{6.5cm}}
   \hline
    \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
-  \ghsmurl{dfalster/baad} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \citep{Falster-2015}.\\
-  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & Provides a taxonomic lookup table for land plants \citep{Pennell-2015a}.\\
+  \ghsmurl{dfalster/baad} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \cite{Falster-2015}.\\
+  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & Provides a taxonomic lookup table for land plants \cite{Pennell-2015a}.\\
   \hline
@@
 
-\section{Figures}
+\section*{Figures}
 

### 2017-11-21 ms.tex
Seperate into Intro, Methods, Results

@@
 
-\section*{Background}
+\section*{Introduction}
 
@@
 
-\section*{A lightweight, cheap, and scalable workflow for delivering versioned data}
+\section*{Methods}
 
@@
 
-Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), that facilitates access to releases of any versioned data hosted on \texttt{GitHub} (Fig.
+To this end, we have developed a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), that facilitates access to releases of any versioned data hosted on \texttt{GitHub} (Fig.
 \ref{fig:technology_stack}).
-Specifically, the \texttt{datastorr} package: 1)Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}; and 2) Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
+Specifically, the \texttt{datastorr} package: 1) Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}; and 2) Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
+
+\section*{Results}
+
+Using the tools outlined above, researchers have immediate access to a lightweight, cheap, and scalable workflow for delivering versioned data.
 Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.
-
 For example, \texttt{datastorr} has been used to build several packages (Table \ref{tab:examples}), including \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
@@
 
-Then as the dataset grows over time, the \emph{maintainers} update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number.
+Then, as the dataset grows over time, the \emph{maintainers} update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number.
 All the releases are simultaneously available to any user, both point-and-click and programmatically.

### 2017-12-15 ms.tex
format for gigascience

--- 2017-11-09 ms.tex
@@
-\documentclass[english]{article}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{babel}
-\usepackage{amsmath}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Please note that whilst this template provides a 
+% preview of the typeset manuscript for submission, it 
+% will not necessarily be the final publication layout.
+%
+% letterpaper/a4paper: US/UK paper size toggle
+% num-refs/alpha-refs: numeric/author-year citation and bibliography toggle
+
+%\documentclass[letterpaper]{oup-contemporary}
+\documentclass[a4paper,num-refs]{assets/oup-contemporary}
+
+%%% Journal toggle; only specific options recognised.
+%%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
+\journal{gigascience}
+
 \usepackage{graphicx}
-\usepackage{fancyhdr}
-\usepackage[colorlinks=false, allcolors=grey, breaklinks=true]{hyperref}
+\usepackage{siunitx}
 
-\RequirePackage[superscript,biblabel,nomove]{cite}
-\bibliographystyle{assets/naturemag-doi}
-\RequirePackage{assets/jabbrv}
+%%% Flushend: You can add this package to automatically balance the final page, but if things go awry (e.g.
+section contents appearing out-of-order or entire blocks or paragraphs are coloured), remove it!
+% \usepackage{flushend}
 
-%\bibliographystyle{naturemag}
-\newcommand{\scidatalogo}{\includegraphics[height=36pt]{assets/SciData_logo.jpg}}
-\pagestyle{fancy}
-\fancyhf{}
-\renewcommand{\headrulewidth}{0pt}
-\setlength{\headheight}{40pt} 
-\lhead{\textsc{\scidatalogo}}
-\rhead{Falster \emph{et al.} -- Versioned Data Delivery}
+\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
 
-\newcommand{\datastorr}{\texttt{datastorr}}
-\newcommand{\smurl}[1]{{\footnotesize\href{https://#1}{\nolinkurl{#1}}}}
-\newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{\nolinkurl{#1}}}}
+%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
+\author[1,\authfn{1}]{Daniel S. Falster}
+\author[2]{Richard G. FitzJohn}
+\author[3]{Matthew  W. Pennell}
+\author[1]{William K. Cornwell}
+
+\affil[1]{Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
+University of New South Wales, Sydney NSW 2052, Australia}
+\affil[2]{School of Public Health, Imperial College, London SW7 2AZ United Kingdom}
+\affil[3]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver B.C. V6T 1Z4, Canada}
+
+%%% Author Notes
+\authnote{\authfn{1}daniel.falster@unsw.edu.au}
+
+%%% Paper category
+\papercat{Technical Note}
+
+%%% "Short" author for running page header
+\runningauthor{Falster et al.}
+
+%%% Should only be set by an editor
+\jvolume{00}
+\jnumber{0}
+\jyear{2018}
 
@@
 
-\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
-
-\author{Daniel S. Falster\textsuperscript{1,*}, Richard G. FitzJohn\textsuperscript{2},\\ Matthew
-  W. Pennell\textsuperscript{3}, and William K. Cornwell\textsuperscript{1}}
-
+\begin{frontmatter}
 \maketitle
-\thispagestyle{fancy}
-$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
-University of New South Wales, Sydney NSW 2052, Australia; 
-$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom; 
-$^3$ Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver B.C. V6T 1Z4, Canada;
-$^*$ Corresponding author: Daniel Falster (daniel.falster@unsw.edu.au)
-
 \begin{abstract}
@@
 
-%\keywords{Data sharing, Version control, Semantic versioning, Meta-analysis}
+\begin{keywords}
+Data sharing; Version control; Semantic versioning
+\end{keywords}
+\end{frontmatter}
 
-\section*{Background}
+%%% Key points will be printed at top of second page
+\begin{keypoints*}
+\begin{itemize}
+\item Datasets are often being expanded and improved; yet current publishing does not accommodate multiple dataset versions.
+\item We present a Version Data Delivery system, for updating and distributing versioned data.
+\item Using semantic versioning to label versions of dataset conveys helps users identify the type of change that has occurred.
+\end{itemize}
+\end{keypoints*}
 
-As is evidenced in the advent of this journal \cite{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product.
+\newcommand{\datastorr}{\texttt{datastorr}}
+\newcommand{\smurl}[1]{{\footnotesize\href{https://#1}{\nolinkurl{#1}}}}
+\newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{\nolinkurl{#1}}}}
+
+\section{Introduction}
+
+Publications of quality datasets are now considered a first-class scientific products.
 Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
@@
 
-Many of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time.
-Typical changes include: the adding new data, improving the quality of existing data, integrating with other datasets, or re-structuring the dataset content to be able to address new questions.
+Many of the high-quality datasets published in scientific journals are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time.
+Typical changes include: the adding new data, improving the quality of existing data, integrating with other datasets, or re-structuring the dataset content.
 For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
 As research around a data product grows, there might be many such additions.
-Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure previous works remain reproducible.
+Ideally, any changes to a dataset would become immediately available to others, while past versions of the dataset would remain accessible to ensure previous works can be reproduced.
 
-However, current models for publishing datasets do not facilitate distributing successive versions of a dataset in an effective and scalable manner.
-The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories where the data is attributed a Digital Object Identifier (\textsc{doi}) -- such as \texttt{Dryad}, \texttt{Figshare}, or \texttt{Zenodo}.
-Publishing involves a single data deposition.
-And once published, datasets are immutable.
-That means that most published datasets exist as a singular snapshot of the dataset, taken at the time the data descriptor was published.
-While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, then that work has to be re-done by every user of that dataset.
+Yet, current models for publishing datasets do not facilitate distributing successive versions of a dataset.
+The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories where the data is attributed a Digital Object Identifier (DOI) -- such as \texttt{DataDryad}, \texttt{Figshare}, or \texttt{Zenodo}.
+Publishing involves a single deposition.
+Once published, datasets are immutable.
+We refer to this type of data delivery as being ``Static'' (Table \ref{tab:publishing_models}).
+In the Static model, published datasets exist as a singular snapshot taken at the time the data descriptor was published.
+While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, that work has to be re-done by every user of that dataset.
 Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.
-We refer to this type of data delivery as being ``Static'' (Table \ref{tab:publishing_models}).
 
-Many research groups solve the issue of versioning data internally in a variety of ad hoc and suboptimal ways.
-A common solution is to email around the latest versions with version numbers or dates appended to the filename.
+Many research groups solve the issue of versioning data internally in a variety of ad-hoc and suboptimal ways.
+A common solution is to email around the latest version with version numbers or dates appended to the filename.
 Another approach is to repeat corrections or updates that have been made elsewhere.
 Alas, these practices are both inefficient and a barrier to reproducibility, as the dataset used in a future publications may differ to the version previously made available on-line.
-There is therefore a strong need for an easy way to distribute changes to a dataset after its initial publication to potential users, along with notes on what has changed and why since the previous version.
+There is therefore a strong need for an easy way to distribute updates to a dataset after its initial publication, along with notes on what has changed and why since the previous version.
 
-In this article we 1) Introduce the concept of Versioned Data Delivery (\textsc{vdd}); 2) Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and 3) Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
+In this article we 1) Introduce the concept of Versioned Data Delivery (VDD); 2) Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and 3) Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a VDD system.
 The issue of updating and expanding published data has already been addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data, where new data can be added and there exist abilities to correct errors in existing records (e.g.
@@
 
-\section*{A lightweight, cheap, and scalable workflow for delivering versioned data}
+\begin{table*}[h!]
+\centering
+\caption{Alternative frameworks for distributing data.}
+{
+\vspace{0.2cm}
+  \begin{tabular}{p{2.5cm}|p{3cm}p{3cm}p{4cm}}
+  \hline
+  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web databases} & \textbf{Versioned Data Delivery}\\
+  \hline
+   Platform (e.g.) 	& datadryad.org 			& coraltraits.org	  &  github.com\\
+   User access      & Web browser 				& Web browser 			& Web browser or R \\
+   Ease of setup    & Very easy 					& Hard 							& Easy\\
+   Data size        & Up to several Gb 		& Small-Very large 	& Up to 1Gb\\
+   Cost             & Varies 							& Varies  					& Free \\
+   Bandwidth        & Managed by provider & Pro rata 					& Managed by \texttt{GitHub}\\
+   Maintainer skills & None 							& High							& R + \texttt{git} \\
+   User skills      & Web browsing 	& Web browsing  					& Web browsing or R \\
+   Versioning       & None 					& Hard 										& Easy \\
+   DOI minting  &Automatic & Manual 									& Automatic \\
+  \hline 
+  \\
 
-In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
+  \end{tabular}
+  } 
+\label{tab:publishing_models}
+\end{table*}
+
+\section{A lightweight, cheap, and scalable workflow for delivering versioned data}
+
+In brief, the VDD workflow we present borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
 Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer.
-With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer.
+With a VDD system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer.
 In either the development of software or data, successive versions (called ``releases'') are distributed over time.
 The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
-Importantly, our \textsc{vdd} systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.
+Importantly, our VDD systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.
 
 The core technologies used are summarised in Table \ref{tab:technologies} and described further below.
-Several groups will interact with the \textsc{vdd} system, including dataset \emph{maintainers}, \emph{contributors}, and \emph{users}.
-The requirements of these different groups are outlined in Table \ref{tab:user_requirements}.
+Several groups will interact with the VDD system, including dataset \emph{maintainers}, \emph{contributors}, and \emph{users}.
+The requirements of each are outlined in Table \ref{tab:user_requirements}.
 
-\subsection*{Version control}
+\subsection{Version control}
 
@@
 
-\subsection*{Hosting and distributing versioned data}
+\subsection{Hosting and distributing versioned data}
 
@@
 
-Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data users to access the data: those that interact with the data via point and click downloading and those that use programmatic interaction (Fig.
+Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data users accessing the data: those that interact with the data via point and click downloading, and those that use programmatic interaction (Fig.
 \ref{fig:technology_stack}, Table \ref{tab:user_requirements}).
-Specifically, \texttt{GitHub} releases can be downloaded directly by users via point-and-click, or accessed programmatically via the \texttt{GitHub} \textsc{api}.
+Specifically, \texttt{GitHub} releases can be downloaded directly by users or accessed programmatically via the \texttt{GitHub} API.
 
-\subsection*{Semantic versioning}
+\begin{figure}[!hb]
+\centering
+\includegraphics[width=\linewidth]{figures/Figure1.pdf}
+\caption{
+% could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
+Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.
+\textbf{a)} 3 numbers indicate type of change.
+\textbf{b)} A typical release stream.}
+\label{fig:semantic}
+\end{figure}
+
+\subsection{Semantic versioning}
 
@@
   \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor'' enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset curator it might change the results of an analysis done with the prior version and so warrants closer inspection.
-  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a very ``major'' revision, for example improving the entire structure of the dataset and adding new columns.
-These changes are very likely to change the results of most users' analyses or to break code that was written to work on the previous version of the dataset or both.
+  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a ``major'' revision, for example improving the entire structure of the dataset or adding new columns.
+These changes are likely to change the results of most users' analyses or to break code that was written to work on the previous version of the dataset.
 \end{itemize}
@@
 
-\subsection*{\texttt{datastorr} and dataset-specific \texttt{R} packages}
+\begin{table*}[t!]
+\centering
+\caption{Overview of technologies used to maintain, store, and distribute the versioned data as described in this paper.}
+\vspace{0.2cm}
+  \begin{tabular}{p{3cm}p{13cm}}
+  \hline
+  \textbf{Technology} & \textbf{Description} \\\hline
+  API & An Application Programming Interface provides a set of protocols for exchanging information.\\
+   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}. \\
+   DOI & Digital object identifier, which refers a user to a single published object. \\
+   \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data.\\
+   github.com & A commercial web platform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
+   \texttt{R} & Open source statistical and data processing language.\\
+   \hline
+  \end{tabular}
+\label{tab:technologies}
+\end{table*}
+
+\begin{figure*}[!hb]
+\centering
+\includegraphics[width=\linewidth]{figures/Figure2.pdf}
+\caption{Overview of the different users and technologies involved in distributing a versioned dataset via \texttt{datastorr}.}
+\label{fig:technology_stack}
+\end{figure*}
+
+\subsection{The \texttt{datastorr} package and dataset-specific \texttt{R} packages}
 
@@
 Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle.
-To make it easier for users to access versioned data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \cite{R-2017}.
+To make it easier for users to access versioned data via code, we offer a novel implementation of VDD focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \cite{R-2017}.
 
@@
 \ref{fig:technology_stack}).
-Specifically, the \texttt{datastorr} package: 1)Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}; and 2) Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
+Specifically, the \texttt{datastorr} package: 1)Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}; and 2) Contains the main code needed to interact with the \texttt{GitHub} API to retrieve versions of the dataset.
 Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.
@@
   \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored;
-  \item the filename in the release that will contain the versioned data;
-  \item the function used to load this into \texttt{R}.
+  \item the filename in the release that contains data;
+  \item the function used to load the data file into \texttt{R}.
 \end{enumerate}
@@
 The package functions by fetching each data version once (the first time it is requested), and then caching these files locally for future reuse.
-Moreover, \emph{users} can have several versions of the same dataset on their computer and unambiguously access the different versions with one simple function call.
+Moreover, \emph{users} can store several versions of a dataset on their computer and unambiguously access different versions with single function.
 
-\section*{Discussion}
+\begin{table*}[ht!]
+\centering
+\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements.}
+\vspace{0.2cm}
+  \begin{tabular}{p{3cm}p{7cm}p{6cm}}
+  \hline
+% Latex being the worst has made the overflow in the second column flow through to affect the third. I used to know how to deal with this...
+  \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
+  Maintainer & Create and distribute versioned datasets & Low technical overhead \\
+    & & Easy workflow for releasing new versions \\
+    & & Long term preservation \\
+    & & Crowd-source error checking and contributions \\
+    & & Low initial and ongoing cost and maintenance \\
+  Contributor & Contribute to future versions of a dataset & Add new data \\
+    & & Report errors in existing data \\
+  Users (all) & Easy access to releases from a versioned dataset & Introduction \& overview \\
+    & & Long term stability \\
+    & & Clear path for users to become contributors \\
+  Users (programmatic) & Reproduce products using specific versions of a dataset & Programmatic access to releases \\
+    & & Easy installation \\
+    & & Long term stability \\
+  \hline
+  \end{tabular}
+\label{tab:user_requirements}
+\end{table*}
+
+\begin{table*}[b!]
+\centering
+\caption{Example datasets using the Versioned Data Delivery system described in this paper.}
+\vspace{0.2cm}
+  \begin{tabular}{p{4cm}p{2cm}p{10cm}}
+  \hline
+   \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
+  \ghsmurl{dfalster/baad} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \cite{Falster-2015}.\\
+  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & Provides a taxonomic lookup table for land plants \cite{Pennell-2015a}.\\
+  \hline
+  \end{tabular}
+\label{tab:examples}
+\end{table*}
+
+\section{Discussion}
 
@@
 By adopting best practices from software engineering, we believe a workable system now exists.
-Moreover, we created the \texttt{datastorr} package to make such a \textsc{vdd} system easy to implement for \texttt{R} users.
+Moreover, we created the \texttt{datastorr} package to make such a VDD system easy to implement for \texttt{R} users.
 As it builds off established and open source software and data science platforms, the proposed system is already easy to deploy on relatively large scale.
 
-A central feature of the proposed  \textsc{vdd} system is that data are maintained in the cloud.
+A central feature of the proposed  VDD system is that data are maintained in the cloud.
 This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig.
@@
 Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
-Adopting a \textsc{vdd} system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.
+Adopting a VDD system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.
 
@@
 In the long term, scientists want their datasets, software, and papers to preserved and remain accessible.
-To enhance long-term stability of data-versions released on \texttt{GitHub}, users can also choose to automatically archive every version in one of several traditional data archives, with a \textsc{doi} (Digital Object Identifier) minted for each release.
+To enhance long-term stability of data-versions released on \texttt{GitHub}, users can also choose to automatically archive every version in one of several traditional data archives, with a DOI (Digital Object Identifier) minted for each release.
 Currently, both \texttt{Zenodo} (\smurl{zenodo.org}) and  \texttt{FigShare} \smurl{figshare.com} each integrate with \texttt{GitHub} for archiving of material hosted there.
 
-The design choices we made in datastorr represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with versioned data stored in the cloud.
+The design choices we made in datastorr represent only one of many possible ways to adopt a VDD system; we can imagine many alternative models for interacting with versioned data stored in the cloud.
 The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
-Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies --- such as CKAN, OKFN or Git-LFS --- facilitate even better \textsc{vdd} systems in the future.
+Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies --- such as CKAN, OKFN or Git-LFS --- facilitate even better VDD systems in the future.
 There are some really exciting ideas that are currently under development which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \cite{Fli, Dat}.
 
-In short, many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system.
+In short, many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a VDD system.
 By adopting best practices from software engineering, we believe a relatively easy, cheap, and scalable solution for serving versioned data now exists.
 
-\section*{Acknowledgements}
-We thank D Noble for comments on an earlier draft and C Boettiger for helpful discussions. DSF was funded by the Australian Research Council. MWP was funded by a NSERC Discovery Grant.
+\section{Availability of source code and requirements}
 
+\begin{itemize}
+\item Project name:  \texttt{datastorr} 
+\item Project home page: \smurl{github.com/ropenscilabs/datastorr}.
+\item Operating system(s): Platform independent
+\item Programming language: R
+\item License: MIT
+\end{itemize}
+
+\section{Declarations}
+
+\subsection{List of abbreviations}
+API = An Application Programming Interface provides a set of protocols for exchanging information. DOI = Digital Object Identifier. VDD = Versioned Data Delivery
+
+\subsection{Ethical Approval}
+Not applicable.
+
+\subsection{Consent for publication}
+Not applicable.
+\subsection{Competing Interests}
+The author(s) declare that they have no competing interests.
+
+\subsection{Funding}
+
+DSF was funded by the Australian Research Council. MWP was funded by a NSERC Discovery Grant.
+
+\subsection{Author's Contributions}
 RF developed the datastorr package. All authors discussed the concepts and wrote the paper.
 
-\newpage
+\section{Acknowledgements}
 
-\section*{Tables}
+We thank D Noble for comments on an earlier draft and C Boettiger for helpful discussions.
 
-\begin{table}[h!]
-\centering
-\caption{Alternative frameworks for distributing data.}
-{\footnotesize
-\vspace{1cm}
-  \begin{tabular}{p{2.5cm}p{3cm}p{3cm}p{4cm}}
-  \hline
-  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web databases} & \textbf{Versioned Data Delivery}\\
-   Platform (e.g.) 	& datadryad.org 			& coraltraits.org	  &  github.com\\
-   User access      & Web browser 				& Web browser 			& Web browser or R \\
-   Ease of setup    & Very easy 					& Hard 							& Easy\\
-   Data size        & Up to several Gb 		& Small-Very large 	& Up to 1Gb\\
-   Cost             & Varies 							& Varies  					& Free \\
-   Bandwidth        & Managed by provider & Pro rata 					& Managed by \texttt{GitHub}\\
-   Maintainer skills & None 							& High							& R + \texttt{git} \\
-   User skills      & Web browsing 	& Web browsing  					& Web browsing or R \\
-   Versioning       & None 					& Hard 										& Easy \\
-   \textsc{doi} minting  &Automatic & Manual 									& Automatic \\
-  \hline 
-  \\
+% \section{Authors' information}
 
-  \end{tabular}
-  } 
-\label{tab:publishing_models}
-\end{table}
-
-\newpage
-
-\begin{table}[h!]
-\centering
-\caption{Overview of technologies used to maintain, store, and distribute the versioned data as described in this paper.}
-{\footnotesize
-\vspace{1cm}
-  \begin{tabular}{p{3cm}p{9cm}}
-  \hline
-  \textbf{Technology} & \textbf{Description} \\\hline
-  \textsc{API} & An Application Programming Interface provides a set of protocols for exchanging information.\\
-   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}. \\
-   \textsc{doi} & Digital object identifier, which refers a user to a single published object. \\
-   \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data.\\
-   github.com & A commercial web platform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
-   \texttt{R} & Open source statistical and data processing language.\\
-   \hline
-  \end{tabular}
-  }
-\label{tab:technologies}
-\end{table}
-
-\newpage
-
-\begin{table}[h!]
-\centering
-\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements.}
-{\footnotesize
-\vspace{1cm}
-  \begin{tabular}{p{2cm}p{4cm}p{6cm}}
-  \hline
-% Latex being the worst has made the overflow in the second column flow through to affect the third. I used to know how to deal with this...
-  \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
-  Maintainer & Create and distribute versioned datasets & Low technical overhead \\
-    & & Easy workflow for releasing new versions \\
-    & & Long term preservation \\
-    & & Easy to crowd-source error checking and contributions \\
-    & & Low initial cost \\
-    & & Low on-going maintenance \\
-  Contributor & Contribute to future versions of a dataset & Add new data \\
-    & & Report errors in existing data \\
-  Users (all) & Easy access to releases from a versioned dataset & Introduction \& overview \\
-    & & Long term stability \\
-    & & Clear path for users to become contributors \\
-  Users (programmatic) & Build reproducible products using specific versions of a dataset & Programmatic access to releases \\
-    & & Easy installation \\
-    & & Long term stability \\
-  \hline
-  \end{tabular}
-}
-\label{tab:user_requirements}
-\end{table}
-
-\newpage
-
-\begin{table}[h!]
-\centering
-\caption{Example datasets using the Versioned Data Delivery system described in this paper.}
-{\footnotesize
-\vspace{1cm}
-
-  \begin{tabular}{p{3.5cm}p{2cm}p{6.5cm}}
-  \hline
-   \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
-  \ghsmurl{dfalster/baad} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \cite{Falster-2015}.\\
-  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & Provides a taxonomic lookup table for land plants \cite{Pennell-2015a}.\\
-  \hline
-  \end{tabular}
-  }
-\label{tab:examples}
-\end{table}
-
-\newpage
-
-\section*{Figures}
-
-\begin{figure}[!hb]
-\centering
-\includegraphics[width=\linewidth]{figures/Figure1.pdf}
-\caption{
-% could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
-Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.
-\textbf{a)} 3 numbers indicate type of change.
-\textbf{b)} A typical release stream.}
-\label{fig:semantic}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!hb]
-\centering
-\includegraphics[width=\linewidth]{figures/Figure2.pdf}
-\caption{Overview of the different users and technologies involved in distributing a versioned dataset via \texttt{datastorr}.}
-\label{fig:technology_stack}
-\end{figure}
-
-\clearpage
+% You may choose to use this section to include any relevant information about the author(s) that may aid the reader's interpretation of the article, and understand the standpoint of the author(s).
+This may include details about the authors' qualifications, current positions they hold at institutions or societies, or any other relevant background information.
+Please refer to authors using their initials.
+Note this section should not be used to describe any competing interests.
 

### 2018-10-03 ms.tex
Rework paper to address reviewer's comments

@@
 
-\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
+\title{Datastorr: a workflow and package for delivering successive versions of ``living data'' directly into R}
 
@@
 
-\affil[1]{Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
+\affil[1]{Evolution \& Ecology Research Centre, and School of Biological, Earth and Environmental Sciences,
 University of New South Wales, Sydney NSW 2052, Australia}
-\affil[2]{School of Public Health, Imperial College, London SW7 2AZ United Kingdom}
+\affil[2]{Department of Infectious Disease Epidemiology, Imperial College London, Faculty of Medicine, Norfolk Place, London W2 1PG, United Kingdom}
 \affil[3]{Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver B.C. V6T 1Z4, Canada}
@@
 Multiple platforms now allow quick and easy data sharing.
-So far, however, data publishing models have not accommodated on-going scientific improvements in data: for many problems, datasets continue to grow with time -- more records are added, errors fixed, and new data structures are created.
-In other words, datasets, like scientific knowledge, advance with time.
-We therefore suggest that many datasets would be usefully published as a series of versions, with a simple naming system to allow users to perceive the type of change between versions.
-In this article, we argue for adopting the paradigm and processes for versioned data, analogous to software versioning.
-We also introduce a system called Versioned Data Delivery and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply.
-These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.
+So far, however, data publishing offers limited functions for interacting with evolving datasets - those that continue to grow with time as more records are added, errors fixed, and new data structures are created.
+In this article, we describe a workflow for maintaining and distributing successive versions of an evolving dataset, allowing Users to retrieve and load different versions of a dataset directly into the R platform.
+Our workflow utilises tools and platforms used for development and distribution of successive versions of a open source software, including version control, GitHuB and semantic versioning, and applies these to the analogous process of developing successive versions of an open source datasets.
+Moreover, we argue that this model allows for individual research groups to achieve a dynamic and versioned model of data delivery at no cost and with maximal performance.
 \end{abstract}
@@
 
-\newcommand{\datastorr}{\texttt{datastorr}}
-\newcommand{\smurl}[1]{{\footnotesize\href{https://#1}{\nolinkurl{#1}}}}
-\newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{\nolinkurl{#1}}}}
+\newcommand{\smurl}[1]{\href{https://#1}{#1}}
+\newcommand{\ghsmurl}[1]{\href{https://github.com/#1}{#1}}
 
@@
 
-Publications of quality datasets are now considered a first-class scientific products.
-Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
-Datasets are now routinely archived as part of the publishing process.
-Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals.
-Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}.
-In particular, the current model for publishing data has not yet embraced the idea that many datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper.
-These datasets are constantly evolving, i.e.
-they are never ``finished'', and may be the subject of repeated analysis or meta-analysis.
+Publication of a quality dataset -- any unique collection of measurements, stored in one or several files -- is now considered a first-class scientific product.
+Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing datasets, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
+Evidence of this trend is seen in the increasing numbers of standalone ``Data papers'' appearing in both standard domain-level journals and specialised data journals.
+Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Force11-2014, Lowndes-2017, Perkel-2016, VanNoorden-2013, Kratz-2015,Wilkinson-2016, Yenni-2018}.
+In particular, in some areas, such as our own area of ecology and evolution, we are only starting to support the fact that many high-quality datasets used in research are evolving entities \cite{Yenni-2018}.
 
-Many of the high-quality datasets published in scientific journals are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time.
-Typical changes include: the adding new data, improving the quality of existing data, integrating with other datasets, or re-structuring the dataset content.
-For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
-As research around a data product grows, there might be many such additions.
-Ideally, any changes to a dataset would become immediately available to others, while past versions of the dataset would remain accessible to ensure previous works can be reproduced.
+An evolving (or `living') dataset is one that is subject to recurrent or occasional change.
+Typical changes may include improving the quality of existing data, adding new data, re-structuring the dataset content, or integrating with other datasets.
+For example, a dataset on biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
+Moreover, change may not be a temporary state, rather the dataset continue to evolve over extended periods \cite[e.g.][]{Ernest-2018}.
+Evolving datasets are never ``finished'', and as such there is no ``master'' or ``canonical'' version.
+Rather, as research around a data product grows, there might be many valid versions produced.
+Even datasets that are not initially envisioned as evolving may become so as minor errors are identified and corrected during use.
 
-Yet, current models for publishing datasets do not facilitate distributing successive versions of a dataset.
-The most common current working model for publishing scientific datasets is to archive them in a variety of open-source data repositories where the data is attributed a Digital Object Identifier (DOI) -- such as \texttt{DataDryad}, \texttt{Figshare}, or \texttt{Zenodo}.
-Publishing involves a single deposition.
-Once published, datasets are immutable.
-We refer to this type of data delivery as being ``Static'' (Table \ref{tab:publishing_models}).
-In the Static model, published datasets exist as a singular snapshot taken at the time the data descriptor was published.
-While it might only require a small amount of work to update the dataset, if there is no way to distribute the updated dataset, that work has to be re-done by every user of that dataset.
-Moreover, those required changes add steps between the canonical dataset and any analysis using it, making reproducibility of the new analysis more difficult than it needs to be.
+A common approach taken by those maintaining an evolving dataset is to release sequential versions of the dataset, each containing a snapshot of the dataset at the time of release \cite[e.g.][]{Falster-2015, Pennell-2015a, Yenni-2018, Abolfathi-2018}.
+Ideally, the latest versions of a dataset would be immediately available to global users, along with notes describing the changes when compared to previous versions.
+For the sake of reproducibility, previous versions of the dataset should also remain available.
+Historically, small research groups might have solved the issue of versioning data internally, for example by emailing around the latest version.
+However, as science moves towards more systematic publishing of data, scalable solutions are needed to distribute dataset versions to a wider variety of users.
 
-Many research groups solve the issue of versioning data internally in a variety of ad-hoc and suboptimal ways.
-A common solution is to email around the latest version with version numbers or dates appended to the filename.
-Another approach is to repeat corrections or updates that have been made elsewhere.
-Alas, these practices are both inefficient and a barrier to reproducibility, as the dataset used in a future publications may differ to the version previously made available on-line.
-There is therefore a strong need for an easy way to distribute updates to a dataset after its initial publication, along with notes on what has changed and why since the previous version.
+One approach taken by large research consortia has been to create dedicated web servers for archiving and delivering of data.
+Projects such as the Sloan Digital Sky Survey (\smurl{https://www.sdss.org}) have sophisticated infrastructure and processes for storing and distributing successive versions of very large datasets \cite{Abolfathi-2018}.
+The issue of updating and expanding published data has also been addressed in some centralised repositories, like genetic sequences (via \texttt{GenBank}), where new data can be added and there exist abilities to correct errors in existing records.
+Yet these web databases require a level of funding and technological infrastructure that is beyond most research groups.
 
-In this article we 1) Introduce the concept of Versioned Data Delivery (VDD); 2) Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and 3) Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a VDD system.
-The issue of updating and expanding published data has already been addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data, where new data can be added and there exist abilities to correct errors in existing records (e.g.
-by adding multiple version of a genetic sequence).
-Yet these ``Dynamic'' web databases (Table \ref{tab:publishing_models}), require a level of infrastructure that is beyond most research groups.
-Our focus here is on the wide range of datasets, such as those appearing in this journal, that are not covered by these repositories.
-In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized.
-Such data support important research projects on particular scientific questions, but are not, in most cases, general enough to warrant custom infrastructure.
-Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.
+Smaller research projects, however, rely on more generic data repositories for publishing of data.
+A common approach for publishing a dataset is to release it under a Digital Object Identifier (DOI) in a standalone data repository, such as DataDryad, Figshare, and Zenodo.
+While these platforms did not initially support versioning of datasets, all have subsequently added the ability to support multiple versions of a dataset, either under a single or different DOIs.
+Yet, while these new features in principle allow users to access multiple versions of a dataset, the release, discovery, and access to multiple versions of a dataset is not always straightforward.
 
-\begin{table*}[h!]
-\centering
-\caption{Alternative frameworks for distributing data.}
-{
-\vspace{0.2cm}
-  \begin{tabular}{p{2.5cm}|p{3cm}p{3cm}p{4cm}}
-  \hline
-  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web databases} & \textbf{Versioned Data Delivery}\\
-  \hline
-   Platform (e.g.) 	& datadryad.org 			& coraltraits.org	  &  github.com\\
-   User access      & Web browser 				& Web browser 			& Web browser or R \\
-   Ease of setup    & Very easy 					& Hard 							& Easy\\
-   Data size        & Up to several Gb 		& Small-Very large 	& Up to 1Gb\\
-   Cost             & Varies 							& Varies  					& Free \\
-   Bandwidth        & Managed by provider & Pro rata 					& Managed by \texttt{GitHub}\\
-   Maintainer skills & None 							& High							& R + \texttt{git} \\
-   User skills      & Web browsing 	& Web browsing  					& Web browsing or R \\
-   Versioning       & None 					& Hard 										& Easy \\
-   DOI minting  &Automatic & Manual 									& Automatic \\
-  \hline 
-  \\
+We believe more can be done to streamline the delivery of a potentially large number of dataset versions to users, especially for small research teams with limited budgets.
+There are at least three challenges.
+First, dataset developers need a cheap -- ideally free -- and reliable system  to create and distribute versioned datasets with low technical overhead.
+Second, users need an easy mechanism to discover the existence of new (or all) versions of a dataset.
+Third, users need a mechanism to retrieve different versions.
+For those using a computational language such as \texttt{R} \cite{R-2017}, all versions of a dataset would ideally be both accessible and discoverable directly from within \texttt{R}.
 
-  \end{tabular}
-  } 
-\label{tab:publishing_models}
-\end{table*}
+In this article we outline how emerging technologies from software development  (Table \ref{tab:technologies}) can be used to address these challenges, enabling small research groups to create and maintain a stream of versions for small-to-medium sized datasets (up to 2Gb), and distribute these directly into the \texttt{R} computational environment for a potentially unlimited number of users at zero financial cost.
+To achieve this we developed a new \texttt{R} package called \texttt{datastorr}, which together with other the technologies allows for easy and scalable delivery of streams of versioned data directly into \texttt{R}.
+At time of publishing this article, this workflow was being used to distribute versions of datasets across a wide range of topics (Table \ref{tab:examples}), suggesting a potentially wide domain of application.
 
-\section{A lightweight, cheap, and scalable workflow for delivering versioned data}
-
-In brief, the VDD workflow we present borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data.
-Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer.
-With a VDD system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer.
-In either the development of software or data, successive versions (called ``releases'') are distributed over time.
-The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data.
-Importantly, our VDD systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.
-
-The core technologies used are summarised in Table \ref{tab:technologies} and described further below.
-Several groups will interact with the VDD system, including dataset \emph{maintainers}, \emph{contributors}, and \emph{users}.
-The requirements of each are outlined in Table \ref{tab:user_requirements}.
-
-\subsection{Version control}
-
-Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.
-In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes.
-Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \cite{Ram-2013, Perkel-2016, Lowndes-2017}.
-\texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g.
-``.csv'', ``.tsv'', ``.txt'').
-The history is visible to anyone interacting with the repository.
-It also allows users to annotate changes (``commits'') with informative messages detailing the rationale for those changes.
-In it's present form, \texttt{git} can handle individual data files at least up to 100MB, which includes a large fraction of scientific cases.
-
-As a general strategy for tracking a dataset under version control with git, we recommend:
-\begin{enumerate}
-  \item Dataset \emph{maintainers} establish a separate \texttt{git} repository for any dataset to be distributed.
-  \item Saving all files as plain text, so that \texttt{git} can identify line-by-line changes. For example, you might maintain tabular data as a ``csv''.
-  \item Saving data in their rawest form. In some datasets you might only have a single file. Others may have may files that get manipulated or combined in some way to produce a unified product.
-  \item Including in the \texttt{git} repository any code needed to manipulate or compile the raw data files into the final dataset.
-For example, you might combine many independent datasets into one unified dataset.
-  \item Documenting any change in the dataset as a ``commit'' into the \texttt{git} repository logs, with informative commit messages outlining why a particular change was made.
-\end{enumerate}
-
-\subsection{Hosting and distributing versioned data}
-
-Datasets stored under version control via \texttt{git} reach their real potential when hosted at a suitable internet hosting service \cite{Ram-2013,Perkel-2016}.
-Here we focus on the platform \texttt{GitHub} (Table \ref{tab:technologies}), but similar functions could be achieved via other providers such as \href{http://bitbucket.org}{bitbucket.org} and \href{http://gitlab.com}{gitlab.com}.
-Hosting of a \texttt{git} repository enables dataset \emph{maintainers} to connect with other potential \emph{contributors} and also \emph{users} (Fig.
-\ref{fig:technology_stack}).
-These platforms are designed to work with \texttt{git} repositories, and thus offer many helpful features, such as ability to record issues, host documentation, or review edits over time.
-
-Of particular interest for current purposes, is \texttt{GitHub}'s ability to host a stream of ``releases'' from the dataset, alongside the \texttt{git} repository containing all the rawfiles.
-Each release is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset \emph{maintainer} decided to generate a new version of the data for distribution.
-While \emph{users} could in principle download the entire \texttt{git} repository, actually all they want are the releases.
-
-Deciding when to make a new release is at the discretion of the dataset \emph{maintainer}.
-In practice, one makes fewer releases than one does commits into the \texttt{git} repository, though there is nothing stopping \emph{maintainers} from releasing a new version for every commit.
-The flexibility here allows \emph{maintainers} to do internal work between releases and only release the data to users when the revision represents a clear improvement on the previous release.
-
-Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data users accessing the data: those that interact with the data via point and click downloading, and those that use programmatic interaction (Fig.
-\ref{fig:technology_stack}, Table \ref{tab:user_requirements}).
-Specifically, \texttt{GitHub} releases can be downloaded directly by users or accessed programmatically via the \texttt{GitHub} API.
-
-\begin{figure}[!hb]
-\centering
-\includegraphics[width=\linewidth]{figures/Figure1.pdf}
-\caption{
-% could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
-Semantic versioning allows users to anticipate the types of changes that have occurred between successive versions of a dataset.
-\textbf{a)} 3 numbers indicate type of change.
-\textbf{b)} A typical release stream.}
-\label{fig:semantic}
-\end{figure}
-
-\subsection{Semantic versioning}
-
-To realize the full benefits of a versioned controlled dataset, users should be able to easily intuit the types of changes that have occurred among versions.
-Since software development has effectively already dealt with almost identical; problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.
-
-Specifically, we suggest applying the theory of ``semantic versioning'', developed for software distribution (\href{http://semver.org/}{semver.org}), to successive releases of a dataset.
-Semantic versioning uses tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers.
-For example, version ``1.0.0''.
-Changes in the version number then signal the type of changes that occurred in the dataset (Fig \ref{fig:semantic}).
-Applying semantic versioning to data, a change from:
-\begin{itemize}
-  \item {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a ``correction'' (or ``patch''), for example a small error correction which is unlikely to break, or change in a substantial way, a users' analyses, although there may be minor changes in the results.
-  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor'' enhancement, for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset curator it might change the results of an analysis done with the prior version and so warrants closer inspection.
-  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a ``major'' revision, for example improving the entire structure of the dataset or adding new columns.
-These changes are likely to change the results of most users' analyses or to break code that was written to work on the previous version of the dataset.
-\end{itemize}
-Any user of a dataset where releases were tagged using semantic versioning, would immediately know the types of change that might have occurred when requesting different versions and set their there expectations accordingly.
-
-\begin{table*}[t!]
+\begin{table*}[ht!]
 \centering
@@
   \textbf{Technology} & \textbf{Description} \\\hline
-  API & An Application Programming Interface provides a set of protocols for exchanging information.\\
-   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}. \\
-   DOI & Digital object identifier, which refers a user to a single published object. \\
+  % \textbf{Technologies} &\\
   \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data.\\
-   github.com & A commercial web platform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
-   \texttt{R} & Open source statistical and data processing language.\\
+  \texttt{GitHub} & A commercial web platform available at \smurl{github.com} for sharing, visualising, and managing \texttt{git} repositories.
+Includes ability to browse the `history', `issue' tracking, and ability to host `releases'.
+Also has a well developed Application Programming Interface (API) enabling programmatic access to dataset releases.\\
+  \texttt{R} & Widely-used and open-source language for data processing and statistical analysis.\\
+  \texttt{datastorr} & A package in \texttt{R} used to fetch versioned releases of a dataset hosted on \texttt{GitHub}. \\
+  \texttt{semantic versioning} & ???\\
   \hline
@@
 \centering
-\includegraphics[width=\linewidth]{figures/Figure2.pdf}
-\caption{Overview of the different users and technologies involved in distributing a versioned dataset via \texttt{datastorr}.}
+\includegraphics[width=\linewidth]{figures/Figure-stack.pdf}
+\caption{Overview of the workflow, including Users and technologies, involved in maintaining an distributing a versioned dataset via \texttt{datastorr}.
+Core features of our approach are shown with black boxes and arrows.
+Optional extensions are shown in grey (see Discussion for details).}
 \label{fig:technology_stack}
@@
 
-\subsection{The \texttt{datastorr} package and dataset-specific \texttt{R} packages}
+\section{A lightweight, cheap, and scalable workflow for delivering versioned data into \texttt{R}}
 
-To aid reproducibility and efficient usage, many users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}).
-Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle.
-To make it easier for users to access versioned data via code, we offer a novel implementation of VDD focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \cite{R-2017}.
+In brief, the workflow we present here borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versions of a dataset.
+Our approach envisions multiple parties involved in the creation and/or use of a versioned dataset, including \emph{Developers}, \emph{Contributors} and \emph{Users} (Fig.
+\ref{fig:technology_stack}).
+Each of these will likely have different goals and requirements (see Table \ref{tab:user_requirements}).
+When building a piece of software, Developers maintain a core set of code which produces the binary executable file that is eventually installed on a User's local computer.
+Analogously, Developers of a dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e.
+loaded) on a User's local computer.
+In either the development of software or data, successive versions -- called ``releases'' -- are distributed as snapshots of the generated product at a particular point in time.
 
-Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), that facilitates access to releases of any versioned data hosted on \texttt{GitHub} (Fig.
+The similarity in workflow between software and data allows us to deploy the re-purpose some of the same technological platforms that are used to maintain and distribute versions of a software product to maintain and distribute versions of a dataset (Table \ref{tab:technologies}).
+Importantly, these tools are available free of charge for open source projects and already well developed -- ensuring high-level performance and stability.
+Moreover, the combination of technologies allow us to address the goals and requirements of the different parties involved in the creation and use of a versioned dataset (see Table \ref{tab:user_requirements}).
+
+An overview of the proposed system is as follows.
+\begin{itemize}
+  \item Raw data files are stored under version control in a \texttt{git} repository --  a free and leading version control system used in software development -- by the dataset Developers.
+All the files that go together to build a single dataset are stored in the repository, together any code used to manipulate these files to create the dataset that is ultimately distributed.
+  \item Changes to the raw data files and code are tracked by the Developers using  \texttt{git}'s ability to make ``commits'' -- granular and annotated snapshots of the source files over time.
+  \item The \texttt{git} repository is hosted on \texttt{GitHub} -- a leading platform for hosting, enabling multiple Developers or other Contributors to work collaboratively on improving a dataset.
+  \item Developers use the files in the repository to make a release of the dataset -- a snapshot of the generated data product at a particular commit -- and upload these to \texttt{GitHub}, where they are hosted alongside the raw files and labelled using ``semantic versioning''.
+The version labels indicate both the ordering of versions and the magnitude of change expected between different versions.
+  \item Using the \texttt{datastorr} package, Users can both retrieve a list of all available versions of the dataset, and retrieve particular versions of the dataset on demand, and load them directly into \texttt{R}.
+\end{itemize}
+
+Below we elaborate on each of the different technologies.
+
+\subsection{Version control}
+
+Version control, primarily an open-source variety called \texttt{git}, has become widespread in software development.
+In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes.
+Increasingly version control has been applied to scientific code and also data management, especially for small-to-medium sized datasets \cite{Ram-2013, Perkel-2016, Lowndes-2017}.
+\texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g.
+``.csv'', ``.tsv'', ``.txt''; with some tricks git can also indicate changes in some other file types such as ``.xlsx'').
+It  allows users to annotate commits with informative messages detailing the rationale for those changes.
+The ``history'' of commits is also visible to anyone interacting with the repository.
+In its present form, \texttt{git} can handle individual data files at least up to 100MB, which includes a large fraction of scientific cases.
+
+As a general strategy for tracking a dataset under version control with \texttt{git}, we recommend:
+\begin{itemize}
+  \item Developers establish a separate \texttt{git} repository for each dataset to be distributed.
+  \item Saving all files as plain text, so that \texttt{git} can identify line-by-line changes. For example, save tabular data as a ``csv''.
+  \item Saving data in their rawest form. In some datasets you might only have a single file. Others may have may files that get manipulated or combined in some way to produce a unified product.
+  \item Including in the \texttt{git} repository any code needed to manipulate or compile the raw data files into the final dataset.
+For example, you might combine many independent datasets into one unified dataset.
+  \item Documenting any changes in the dataset by making a commit in the \texttt{git} repository, with informative message outlining why the change was made.
+\end{itemize}
+
+\subsection{Hosting and distributing versioned data}
+
+Datasets stored under version control via \texttt{git} reach their real potential when hosted at a suitable internet hosting service \cite{Ram-2013,Perkel-2016}.
+Here we focus on the platform \texttt{GitHub} (Table \ref{tab:technologies}).
+Hosting of a \texttt{git} repository enables dataset Developers to connect with other potential Contributors and also Users (Fig.
 \ref{fig:technology_stack}).
-Specifically, the \texttt{datastorr} package: 1)Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}; and 2) Contains the main code needed to interact with the \texttt{GitHub} API to retrieve versions of the dataset.
-Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.
+These platforms are designed to work with \texttt{git} repositories, and thus offer many helpful features, such as ability to record issues, host documentation, or review edits over time.
+
+Of particular interest for current purposes, is \texttt{GitHub}'s ability to host a stream of releases from the dataset, alongside the \texttt{git} repository containing all the raw files.
+Each release is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset Developer decided to generate a new version of the data for distribution.
+While Users could in principle download the entire \texttt{git} repository, actually all they want (most of the time) are the releases.
+
+Deciding when to make a new release is at the discretion of the dataset Developer.
+In practice, one makes fewer releases than one does commits into the \texttt{git} repository, though there is nothing stopping Developers from releasing a new version for every commit.
+The flexibility here allows Developers to do internal work between releases and only release the data to Users when the revision represents a clear improvement on the previous release.
+
+Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data Users accessing the data: those that interact with the data via point and click downloading, and those that use programmatic interaction (Fig.
+\ref{fig:technology_stack}, Table \ref{tab:user_requirements}).
+Specifically, \texttt{GitHub} releases can be downloaded directly by Users or accessed programmatically via the \texttt{GitHub} API.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=\linewidth]{figures/Figure-versions.pdf}
+\caption{
+% could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
+Semantic versioning allows dataset Developers to communicate to Users the types of changes that have occurred between successive versions of a dataset.}
+\label{fig:semantic}
+\end{figure}
+
+\begin{table*}[b!]
+\centering
+\caption{Example datasets currently delivered using the \texttt{datastorr} package for \texttt{R}}
+\vspace{0.2cm}
+  \begin{tabular}{p{4cm}p{11cm}}
+  \hline
+   \textbf{\texttt{GitHub} repository} & \textbf{Dataset description} \\ \hline
+  \ghsmurl{dfalster/baad.data} & Size dimensions of plants for many species from across the world \cite{Falster-2015}.\\
+  \ghsmurl{traitecoevo/taxonlookup} & Taxonomy of world's land plants \cite{Pennell-2015a}.\\
+  \ghsmurl{ecohealthalliance/cites} & Trade details from Convention on International Trade in Endangered Species (CITES)\\
+  \ghsmurl{madams1/nbadata} & Statistics from the National Basketball Association (NBA) seasons 1996-97 to 2016-17 \\
+  \ghsmurl{madams1/floridainmates} & Dataset from the Florida Department of Corrections on the state's inmate population\\
+  \ghsmurl{traitecoevo/fungaltraits} & Dataset on the traits of fungi \cite{Cornwell-2018}.\\
+
+  \hline
+  \end{tabular}
+\label{tab:examples}
+\end{table*}
+
+\subsection{Semantic versioning}
+
+To realize the full benefits of a versioned controlled dataset, Users should be able to easily intuit the types of changes that have occurred among versions.
+Since software development has effectively already dealt with almost identical; problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.
+
+Specifically, we suggest applying the theory of semantic versioning, developed for software distribution (\href{http://semver.org/}{semver.org}), to successive releases of a dataset.
+Semantic versioning uses a tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers.
+For example, version ``1.0.0''.
+Changes in the version number then signal the type of changes that occurred in the dataset (Fig \ref{fig:semantic}).
+Applying semantic versioning to data, a change from:
+\begin{itemize}
+  \item {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a ``correction'' (or ``patch''), for example a small error correction which is unlikely to break, or change in a substantial way, a user's analyses, although there may be minor changes in the results.
+  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor enhancement'', for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset curator it might change the results of an analysis done with the prior version and so warrants closer inspection.
+  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a ``major update'', for example improving the entire structure of the dataset, or adding of new columns.
+These changes are likely to change the results of most user's analyses or to break code that was written to work on the previous version of the dataset.
+\end{itemize}
+Any User of a dataset where releases were tagged using semantic versioning, would immediately know the types of change that might have occurred when requesting different versions and set their there expectations accordingly.
+
+\subsection{Loading data versions directly into \texttt{R} using the \texttt{datastorr} package}
+
+For efficient usage and to aid reproducibility, many Users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}).
+Code to access a stream of \texttt{GitHub} releases could be written individually by each User, but this creates an unnecessary technological hurdle.
+To make it easier for Users to access versioned data via code, we developed a new package for the \texttt{R} platform, as one of the most prominent platforms for data science \cite{R-2017}.
+
+Our package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), facilitates access to releases of any versioned data hosted on \texttt{GitHub} (Fig.
+\ref{fig:technology_stack}).
+Specifically, the \texttt{datastorr} package: 1) Contains the main code needed to interact with the \texttt{GitHub} API to retrieve versions of the dataset; and 2) Enables users to construct the shell of a second, dataset-specific \texttt{R} package, which can be distributed and used to access releases for a specific repository stored on \texttt{GitHub}.
+Using \texttt{datastorr}, a researcher can create and distribute a custom \texttt{R} package that facilitates access to their data with (very) minimal computational skills.
 
@@
 Accessing a different version of the data involves changing only the version number.
-From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \cite{Duursma-2016,Falster-2016}.
+From the user's perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \cite{Duursma-2016,Falster-2016}.
 
-Using \texttt{datastorr}, dataset \emph{maintainers} can set up their own R package to deliver there versioned dataset simply by providing:
+Using \texttt{datastorr}, dataset Developers can set up their own \texttt{R} package to deliver the versioned dataset simply by providing:
 \begin{enumerate}
@@
 \end{enumerate}
-A full tutorial explaining precisely how to set this up is available at \smurl{github.com/ropenscilabs/datastorr}.
 
-Then as the dataset grows over time, the \emph{maintainers} update the \texttt{git} repo and create a \texttt{GitHub} release with a new version number.
+Then as the dataset grows over time, the Developers update the \texttt{git} repository and create a \texttt{GitHub} release with a new version number.
 All the releases are simultaneously available to any user, both point-and-click and programmatically.
@@
 The package functions by fetching each data version once (the first time it is requested), and then caching these files locally for future reuse.
-Moreover, \emph{users} can store several versions of a dataset on their computer and unambiguously access different versions with single function.
+Moreover, Users can store several versions of a dataset on their computer and unambiguously access different versions with single function.
 
-\begin{table*}[ht!]
+\begin{table*}[t!]
 \centering
-\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements.}
+\caption{Goals and requirements of different parties involved in the creation and use of a versioned dataset.}
 \vspace{0.2cm}
@@
 % Latex being the worst has made the overflow in the second column flow through to affect the third. I used to know how to deal with this...
-  \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
-  Maintainer & Create and distribute versioned datasets & Low technical overhead \\
+  \textbf{Group} & \textbf{Primary goal} & \textbf{Requirements} \\ \hline
+  Developer & Create and distribute versions of a dataset & Low technical overhead \\
+    & & Low initial and ongoing cost and maintenance \\
     & & Easy workflow for releasing new versions \\
+    & & Enable User feedback in error checking and contributions \\
     & & Long term preservation \\
-    & & Crowd-source error checking and contributions \\
-    & & Low initial and ongoing cost and maintenance \\
   Contributor & Contribute to future versions of a dataset & Add new data \\
     & & Report errors in existing data \\
-  Users (all) & Easy access to releases from a versioned dataset & Introduction \& overview \\
+  Users (all) & Easy access to all versions of a dataset & Background information\\
     & & Long term stability \\
-    & & Clear path for users to become contributors \\
-  Users (programmatic) & Reproduce products using specific versions of a dataset & Programmatic access to releases \\
+    & & Ability to give feedback and contribute \\
+  Users (programmatic) & Reproduce products using specific versions of a dataset & Programmatic access to all versions of a dataset \\
     & & Easy installation \\
@@
 
-\begin{table*}[b!]
-\centering
-\caption{Example datasets using the Versioned Data Delivery system described in this paper.}
-\vspace{0.2cm}
-  \begin{tabular}{p{4cm}p{2cm}p{10cm}}
-  \hline
-   \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
-  \ghsmurl{dfalster/baad} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \cite{Falster-2015}.\\
-  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & Provides a taxonomic lookup table for land plants \cite{Pennell-2015a}.\\
-  \hline
-  \end{tabular}
-\label{tab:examples}
-\end{table*}
-
 \section{Discussion}
 
-The main issue we have identified in this article will not come as news to many readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about how to handle dataset updates.
-Some specific solutions for versioning already exist, such as those designed for genetic sequence data (e.g., \texttt{GenBank}; see \smurl{www.ncbi.nlm.nih.gov/genbank/sequenceids/}).
-But so far, versioning solutions have only been developed for large dynamic databases managed by well-funded institutions or consortia and involving simple standardised data types.
-While individual research groups working with a specialized type of data could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required (Table \ref{tab:publishing_models}) will be discouraging for most.
-This suggests there is a substantial need for an easy, cheap, and scalable solution for serving versioned data.
-By adopting best practices from software engineering, we believe a workable system now exists.
-Moreover, we created the \texttt{datastorr} package to make such a VDD system easy to implement for \texttt{R} users.
-As it builds off established and open source software and data science platforms, the proposed system is already easy to deploy on relatively large scale.
+The key issue we are dealing with in this article may be familiar to many readers: many datasets are constantly evolving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is as yet little consensus about how to maintain and distribute multiple versions of a dataset, especially for small research teams.
+While such teams could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required are discouraging.
+Moreover, existing data sharing platforms offer a limited set of features for the delivery of successive versions of a dataset.
+This suggests there is a substantial need for an easy, cheap, and scalable solution for maintaining and distributing successive versions of a dataset.
+By adopting open-source and scalable practices from software development, we believe a workable system already largely exists.
+To aid this process, we created the \texttt{datastorr} package to deliver dataset versions directly into the \texttt{R} environment.
+The approach and package is already being used to deliver several versioned datasets spanning a wide range of topics (Table \ref{tab:examples}).
+Moreover, as it builds off established and open source software and data science platforms (Table \ref{tab:technologies}), the proposed system is already easy to deploy on relatively large scale.
 
-A central feature of the proposed  VDD system is that data are maintained in the cloud.
+\subsection{Related approaches}
+
+Our contribution here connects with a growing number of recommendations and technologies supporting the publishing and reuse of data, including multiple versions of datasets.
+Such contributions include community guidance on good practice in data citation \cite{Force11-2014} and the FAIR principles for making datasets Findable, Accessible, Interoperable, and Reusable by both machines and humans \cite{Wilkinson-2016}.
+In our proposed system, information about appropriate attribution for any dataset  (whatever that is determined to be) should be made readily available, either on the landing page within GitHub, or even better distributed as part of the versioned dataset itself.
+Similarly, datasets can be structured to make them follow the FAIR principles, to the extent possible.
+Notably, our workflow with the \texttt{datastorr} package demands machine access to datasets - a core focus for the FAIR principles.
+While our proposed workflow does not currently enhance discoverability of new datasets, this is a broad challenge faced by all data platforms and researchers.
+
+Within the \texttt{R}  ecosystem, the \texttt{datastorr} package complements other approaches for creating and delivering datasets.
+One common approach used within \texttt{R} is to embed data directly within an \texttt{R} package, which can then be distributed via the Comprehensive R Archive Network (\smurl{cran.r-project.org/}).
+Moreover, dedicated packages are being developed to assist dataset Developers in creating data packages \cite{Finak-2018}.
+An advantage of this approach, compared to ours, is that the data are immediately available in the package (whereas our packages only contain instructions for fetching the data).
+This advantage however also brings limitations.
+Notably, datasets must be under 5MB, and only one version of a dataset package can be installed on any given machine at any one time.
+\texttt{datastorr} may therefore offer a viable approach for overcoming these limitations.
+
+There are also many emerging or alternative technologies that offer other possible ways to implement a system for storing and distributing versions of a dataset.
+Our solution currently emphasises the platform GitHub, but similar functions could be achieved via other \texttt{git} hosts such as \href{http://bitbucket.org}{bitbucket.org} and \href{http://gitlab.com}{gitlab.com}.
+Git repositories can also be extended to accommodate larger files using  features like \href{https://git-lfs.github.com/}{Git-Large File Storage} or \href{http://git-annex.branchable.com/}{git-annex}.
+More fundamentally, there are  emerging alternatives for version control specifically designed for data, such as the \smurl{datproject.org}, and other new platforms for distributing data, such as the \href{https://en.wikipedia.org/wiki/CKAN}{Comprehensive Knowledge Archive Network, (CKAN)} and \href{https://okfn.org/}{Open Knowledge International, (OKFN)}.
+
+The key here is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data, which may be achieved with all of these approaches.
+Indeed, as with every technology the best available approach is certain to evolve, especially as emerging technologies facilitate even better delivery of data in the future.
+
+\subsection{Further advantages and extensions}
+
+A central feature of the proposed system is that data are maintained in the cloud.
 This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig.
@@
 Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}.
-Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
-Adopting a VDD system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.
+Multiple Developers can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
+The proposed system of data delivery thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.
 
@@
 As one of the largest companies hosting computer source code, \texttt{GitHub} provides exceptional performance in this regard -- certainly as good or better than nearly any system scientists might build themselves.
+Thus short-term concerns of reliable and fast performance are almost guaranteed.
+
 In the long term, scientists want their datasets, software, and papers to preserved and remain accessible.
-To enhance long-term stability of data-versions released on \texttt{GitHub}, users can also choose to automatically archive every version in one of several traditional data archives, with a DOI (Digital Object Identifier) minted for each release.
-Currently, both \texttt{Zenodo} (\smurl{zenodo.org}) and  \texttt{FigShare} \smurl{figshare.com} each integrate with \texttt{GitHub} for archiving of material hosted there.
-
-The design choices we made in datastorr represent only one of many possible ways to adopt a VDD system; we can imagine many alternative models for interacting with versioned data stored in the cloud.
-The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.
-Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies --- such as CKAN, OKFN or Git-LFS --- facilitate even better VDD systems in the future.
-There are some really exciting ideas that are currently under development which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \cite{Fli, Dat}.
-
-In short, many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a VDD system.
-By adopting best practices from software engineering, we believe a relatively easy, cheap, and scalable solution for serving versioned data now exists.
+While our proposed system for data delivery does not guarantee long-term preservation, users can also choose to automatically archive data-versions released on \texttt{GitHub} version in one of several traditional data archives, with a DOI (Digital Object Identifier) minted for each release.
+Currently, both \texttt{Zenodo} and  \texttt{FigShare}each integrate with \texttt{GitHub} for archiving of material hosted there.
+Ideally, tools like \texttt{datastorr} would also be developed to pull versions from these archives too.
 
@@
 \subsection{List of abbreviations}
-API = An Application Programming Interface provides a set of protocols for exchanging information. DOI = Digital Object Identifier. VDD = Versioned Data Delivery
+API = An Application Programming Interface provides a set of protocols for exchanging information. DOI = Digital Object Identifier.
 

### 2018-10-03 ms.tex
minor edits throughout

@@
 So far, however, data publishing offers limited functions for interacting with evolving datasets - those that continue to grow with time as more records are added, errors fixed, and new data structures are created.
-In this article, we describe a workflow for maintaining and distributing successive versions of an evolving dataset, allowing Users to retrieve and load different versions of a dataset directly into the R platform.
-Our workflow utilises tools and platforms used for development and distribution of successive versions of a open source software, including version control, GitHuB and semantic versioning, and applies these to the analogous process of developing successive versions of an open source datasets.
-Moreover, we argue that this model allows for individual research groups to achieve a dynamic and versioned model of data delivery at no cost and with maximal performance.
+In this article, we describe a workflow for maintaining and distributing successive versions of an evolving dataset, allowing users to retrieve and load different versions of a dataset directly into the R platform.
+Our workflow utilises tools and platforms used for development and distribution of successive versions of a open source software, including version control, GitHub and semantic versioning, and applies these to the analogous process of developing successive versions of an open source datasets.
+Moreover, we argue that this model allows for individual research groups to achieve a dynamic and versioned model of data delivery at no cost.
 \end{abstract}
@@
 For example, a dataset on biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names.
-Moreover, change may not be a temporary state, rather the dataset continue to evolve over extended periods \cite[e.g.][]{Ernest-2018}.
+In some cases, datasets may be expected to continue to evolve over extended periods \cite[e.g.][]{Ernest-2018}.
 Evolving datasets are never ``finished'', and as such there is no ``master'' or ``canonical'' version.
 Rather, as research around a data product grows, there might be many valid versions produced.
-Even datasets that are not initially envisioned as evolving may become so as minor errors are identified and corrected during use.
+%Even datasets that are not initially envisioned as evolving may become so as minor errors are identified and corrected during use.
 
 A common approach taken by those maintaining an evolving dataset is to release sequential versions of the dataset, each containing a snapshot of the dataset at the time of release \cite[e.g.][]{Falster-2015, Pennell-2015a, Yenni-2018, Abolfathi-2018}.
-Ideally, the latest versions of a dataset would be immediately available to global users, along with notes describing the changes when compared to previous versions.
+Ideally, the latest versions of a dataset would be immediately available to all users across the globe, along with notes describing the changes when compared to previous versions.
 For the sake of reproducibility, previous versions of the dataset should also remain available.
@@
 When building a piece of software, Developers maintain a core set of code which produces the binary executable file that is eventually installed on a User's local computer.
-Analogously, Developers of a dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e.
-loaded) on a User's local computer.
+Analogously, Developers of a dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e., loaded) on a User's local computer.
 In either the development of software or data, successive versions -- called ``releases'' -- are distributed as snapshots of the generated product at a particular point in time.
@@
 Increasingly version control has been applied to scientific code and also data management, especially for small-to-medium sized datasets \cite{Ram-2013, Perkel-2016, Lowndes-2017}.
-\texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g.
-``.csv'', ``.tsv'', ``.txt''; with some tricks git can also indicate changes in some other file types such as ``.xlsx'').
+\texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g., ``.csv'', ``.tsv'', ``.txt''; with some tricks git can also indicate changes in some other file types such as ``.xlsx'').
 It  allows users to annotate commits with informative messages detailing the rationale for those changes.
@@
 Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}.
-Multiple Developers can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved.
+Multiple Developers can make changes to different parts of the code (or, in our case, data) and the \texttt{git} system will integrate these together or, when needed, flag where there are conflicts that need to be resolved.
 The proposed system of data delivery thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.
@@
 While our proposed system for data delivery does not guarantee long-term preservation, users can also choose to automatically archive data-versions released on \texttt{GitHub} version in one of several traditional data archives, with a DOI (Digital Object Identifier) minted for each release.
-Currently, both \texttt{Zenodo} and  \texttt{FigShare}each integrate with \texttt{GitHub} for archiving of material hosted there.
+Currently, both \texttt{Zenodo} and  \texttt{FigShare} each integrate with \texttt{GitHub} for archiving of material hosted there.
 Ideally, tools like \texttt{datastorr} would also be developed to pull versions from these archives too.

### 2018-10-05 ms.tex
small wording changes, a bit on semantic versioning, and changing one subsection heading

@@
 
-Publication of a quality dataset -- any unique collection of measurements, stored in one or several files -- is now considered a first-class scientific product.
+Publication of a quality dataset -- a collection of measurements, stored in one or several files -- can now considered a first-class scientific product.
 Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing datasets, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
@@
 Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Force11-2014, Lowndes-2017, Perkel-2016, VanNoorden-2013, Kratz-2015,Wilkinson-2016, Yenni-2018}.
-In particular, in some areas, such as our own area of ecology and evolution, we are only starting to support the fact that many high-quality datasets used in research are evolving entities \cite{Yenni-2018}.
+In particular, in some areas, such as our own area of ecology and evolution, we are only starting to support the fact that many high-quality datasets used in research should be evolving entities \cite{Yenni-2018}.
 
@@
 Rather, as research around a data product grows, there might be many valid versions produced.
+In most cases the most recent version will be the best available information, but there are many reasons to go back to previous versions: to replicate previous analyses or to work on a stable version for downstream analyses or visualisation.
+
 %Even datasets that are not initially envisioned as evolving may become so as minor errors are identified and corrected during use.
@@
 Ideally, the latest versions of a dataset would be immediately available to all users across the globe, along with notes describing the changes when compared to previous versions.
-For the sake of reproducibility, previous versions of the dataset should also remain available.
-Historically, small research groups might have solved the issue of versioning data internally, for example by emailing around the latest version.
-However, as science moves towards more systematic publishing of data, scalable solutions are needed to distribute dataset versions to a wider variety of users.
+For the sake of reproducibility, previous versions of the dataset should be archived and remain available.
+In the recent past, small research groups have solved the issue of versioning data internally and informally, for example by emailing around the latest version.
+However, as science grows and moves towards more systematic publishing of data, scalable solutions are needed to distribute dataset versions to a wider variety of users.
 
@@
 
-Smaller research projects, however, rely on more generic data repositories for publishing of data.
+The vast majority of research projects are smaller and these currently rely on more generic data repositories for publishing of data.
 A common approach for publishing a dataset is to release it under a Digital Object Identifier (DOI) in a standalone data repository, such as DataDryad, Figshare, and Zenodo.
@@
 Second, users need an easy mechanism to discover the existence of new (or all) versions of a dataset.
-Third, users need a mechanism to retrieve different versions.
+Third, users need a mechanism to retrieve specific versions.
 For those using a computational language such as \texttt{R} \cite{R-2017}, all versions of a dataset would ideally be both accessible and discoverable directly from within \texttt{R}.
@@
 
-In brief, the workflow we present here borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versions of a dataset.
+In brief, the workflow we present here borrows best practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versions of a dataset.
 Our approach envisions multiple parties involved in the creation and/or use of a versioned dataset, including \emph{Developers}, \emph{Contributors} and \emph{Users} (Fig.
@@
 \end{itemize}
-Any User of a dataset where releases were tagged using semantic versioning, would immediately know the types of change that might have occurred when requesting different versions and set their there expectations accordingly.
+Any user of a dataset where releases were tagged using semantic versioning, would immediately know the maintainers' view of the types of the magnitude of the changes between versions.
+Note that as with software this taxonomy of changes is a language to more nuanced way to communicate from the maintainer to the user, rather than a set of fixed rules.
+Because users of data--like users of software--do unpredictable things, the maintainer may classify a change as minor, but still could be break some downstream analyses.
 
@@
 
-For efficient usage and to aid reproducibility, many Users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}).
-Code to access a stream of \texttt{GitHub} releases could be written individually by each User, but this creates an unnecessary technological hurdle.
-To make it easier for Users to access versioned data via code, we developed a new package for the \texttt{R} platform, as one of the most prominent platforms for data science \cite{R-2017}.
+For efficient usage and to aid reproducibility, many users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}).
+Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle.
+To make it easier for users to access versioned data via code, we developed a new package for the \texttt{R} platform, as one of the most prominent platforms for data science \cite{R-2017}.
 
@@
 The package functions by fetching each data version once (the first time it is requested), and then caching these files locally for future reuse.
-Moreover, Users can store several versions of a dataset on their computer and unambiguously access different versions with single function.
+Moreover, users can store several versions of a dataset on their computer and unambiguously access different versions with single function.
 
@@
 
-\subsection{Related approaches}
+\subsection{Towards an ecosystem for evolving data}
 
-Our contribution here connects with a growing number of recommendations and technologies supporting the publishing and reuse of data, including multiple versions of datasets.
+Our contribution here connects with a growing number of recommendations and technologies supporting the publishing and reuse of evolving data.
 Such contributions include community guidance on good practice in data citation \cite{Force11-2014} and the FAIR principles for making datasets Findable, Accessible, Interoperable, and Reusable by both machines and humans \cite{Wilkinson-2016}.
@@
 As one of the largest companies hosting computer source code, \texttt{GitHub} provides exceptional performance in this regard -- certainly as good or better than nearly any system scientists might build themselves.
-Thus short-term concerns of reliable and fast performance are almost guaranteed.
+Thus, short-term concerns of reliable and fast performance are almost guaranteed.
 

### 2018-10-10 ms.tex
Further refinements
- consistent phrasing
- minor edits
- update figure
- caps in refs
- add section on python

@@
 
-\title{Datastorr: a workflow and package for delivering successive versions of ``living data'' directly into R}
+\title{Datastorr: a workflow and package for delivering successive versions of ``evolving data'' directly into R}
 
@@
 So far, however, data publishing offers limited functions for interacting with evolving datasets - those that continue to grow with time as more records are added, errors fixed, and new data structures are created.
-In this article, we describe a workflow for maintaining and distributing successive versions of an evolving dataset, allowing users to retrieve and load different versions of a dataset directly into the R platform.
-Our workflow utilises tools and platforms used for development and distribution of successive versions of a open source software, including version control, GitHub and semantic versioning, and applies these to the analogous process of developing successive versions of an open source datasets.
+In this article, we describe a workflow for maintaining and distributing successive versions of an evolving dataset, allowing users to retrieve and load different versions directly into the R platform.
+Our workflow utilises tools and platforms used for development and distribution of successive versions of a open source software, including version control, GitHub and semantic versioning, and applies these to the analogous process of developing successive versions of an open source dataset.
 Moreover, we argue that this model allows for individual research groups to achieve a dynamic and versioned model of data delivery at no cost.
@@
 \begin{itemize}
-\item Datasets are often being expanded and improved; yet current publishing does not accommodate multiple dataset versions.
-\item We present a Version Data Delivery system, for updating and distributing versioned data.
+\item Evolving datasets are those that are often being expanded and improved. Users of evolving datasets require easy ways successive versions.
+\item Using techniques adopted from software development, we present a workflow for maintaining and distributing versions of an evolving dataset.
+A new package \texttt{datastorr} enables fetching and loading successive versions directly in \texttt{R}.
 \item Using semantic versioning to label versions of dataset conveys helps users identify the type of change that has occurred.
@@
 
-Publication of a quality dataset -- a collection of measurements, stored in one or several files -- can now considered a first-class scientific product.
+Publication of a quality dataset -- a collection of measurements, stored in one or several files -- is now considered a first-class scientific product.
 Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing datasets, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
@@
 Rather, as research around a data product grows, there might be many valid versions produced.
-In most cases the most recent version will be the best available information, but there are many reasons to go back to previous versions: to replicate previous analyses or to work on a stable version for downstream analyses or visualisation.
-
-%Even datasets that are not initially envisioned as evolving may become so as minor errors are identified and corrected during use.
+Even datasets that are not initially envisioned as evolving, may become so as minor errors are identified and corrected during use.
+In either case, the most recent version of the dataset will typically contain the best-available information, but there are many reasons to go back to previous versions: to replicate previous analyses or to work on a stable version for downstream analyses or visualisation.
 
 A common approach taken by those maintaining an evolving dataset is to release sequential versions of the dataset, each containing a snapshot of the dataset at the time of release \cite[e.g.][]{Falster-2015, Pennell-2015a, Yenni-2018, Abolfathi-2018}.
-Ideally, the latest versions of a dataset would be immediately available to all users across the globe, along with notes describing the changes when compared to previous versions.
+Ideally, the latest versions of an evolving dataset would be immediately available to all users across the globe, along with notes describing the changes when compared to previous versions.
 For the sake of reproducibility, previous versions of the dataset should be archived and remain available.
@@
 A common approach for publishing a dataset is to release it under a Digital Object Identifier (DOI) in a standalone data repository, such as DataDryad, Figshare, and Zenodo.
-While these platforms did not initially support versioning of datasets, all have subsequently added the ability to support multiple versions of a dataset, either under a single or different DOIs.
+While these platforms did not initially support versioning of datasets, all have subsequently added the ability to support multiple versions of a dataset, either under a single or different DOIs \cite{Nielsen-2017}.
 Yet, while these new features in principle allow users to access multiple versions of a dataset, the release, discovery, and access to multiple versions of a dataset is not always straightforward.
@@
 There are at least three challenges.
-First, dataset developers need a cheap -- ideally free -- and reliable system  to create and distribute versioned datasets with low technical overhead.
-Second, users need an easy mechanism to discover the existence of new (or all) versions of a dataset.
+First, dataset developers need a cheap -- ideally free -- and reliable system  to create and distribute versions of an evolving datasets with low technical overhead.
+Second, users need an easy mechanism to discover the existence of new (or all) versions of an evolving dataset.
 Third, users need a mechanism to retrieve specific versions.
-For those using a computational language such as \texttt{R} \cite{R-2017}, all versions of a dataset would ideally be both accessible and discoverable directly from within \texttt{R}.
+For those using a computational language such as \texttt{R} \cite{R-2017}, all versions of an evolving dataset would ideally be both accessible and discoverable directly from within \texttt{R}.
 
-In this article we outline how emerging technologies from software development  (Table \ref{tab:technologies}) can be used to address these challenges, enabling small research groups to create and maintain a stream of versions for small-to-medium sized datasets (up to 2Gb), and distribute these directly into the \texttt{R} computational environment for a potentially unlimited number of users at zero financial cost.
-To achieve this we developed a new \texttt{R} package called \texttt{datastorr}, which together with other the technologies allows for easy and scalable delivery of streams of versioned data directly into \texttt{R}.
+In this article we outline how emerging technologies from software development  (Table \ref{tab:technologies}) can be used to address these challenges, enabling small research groups to create and maintain a stream of versions for small-to-medium sized datasets (up to 2Gb), and distribute these directly into the \texttt{R} computational environment for a potentially unlimited number of users at zero financial cost and minimal technical overhead.
+To achieve this we developed a new \texttt{R} package called \texttt{datastorr}, which together with other the technologies allows for easy and scalable delivery of successive versions of an evolving dataset directly into \texttt{R}.
 At time of publishing this article, this workflow was being used to distribute versions of datasets across a wide range of topics (Table \ref{tab:examples}), suggesting a potentially wide domain of application.
@@
 \centering
-\caption{Overview of technologies used to maintain, store, and distribute the versioned data as described in this paper.}
+\caption{Overview of technologies used to maintain, store, and distribute versions of an evolving dataset as described in this paper.}
 \vspace{0.2cm}
@@
   \texttt{R} & Widely-used and open-source language for data processing and statistical analysis.\\
-  \texttt{datastorr} & A package in \texttt{R} used to fetch versioned releases of a dataset hosted on \texttt{GitHub}. \\
-  \texttt{semantic versioning} & ???\\
+  \texttt{datastorr} & A package in \texttt{R} used to fetch releases of an evolving dataset hosted on \texttt{GitHub}. \\
+  \texttt{semantic versioning} & The process of assigning unique version numbers in a particular format to successive versions of a digital product; traditionally applied to software but here to an evolving dataset.\\
   \hline
@@
 \includegraphics[width=\linewidth]{figures/Figure-stack.pdf}
-\caption{Overview of the workflow, including Users and technologies, involved in maintaining an distributing a versioned dataset via \texttt{datastorr}.
+\caption{Overview of the workflow, different parties and technologies involved in maintaining an distributing versions of an evolving dataset via \texttt{datastorr}.
 Core features of our approach are shown with black boxes and arrows.
@@
 
-\section{A lightweight, cheap, and scalable workflow for delivering versioned data into \texttt{R}}
+\section{A lightweight, cheap, and scalable workflow for delivering versions of an evolving dataset into \texttt{R}}
 
-In brief, the workflow we present here borrows best practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versions of a dataset.
+In brief, the workflow we present here borrows best practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versions of an evolving dataset.
 Our approach envisions multiple parties involved in the creation and/or use of a versioned dataset, including \emph{Developers}, \emph{Contributors} and \emph{Users} (Fig.
@@
 When building a piece of software, Developers maintain a core set of code which produces the binary executable file that is eventually installed on a User's local computer.
-Analogously, Developers of a dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e., loaded) on a User's local computer.
+Analogously, Developers of an evolving dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e., loaded) on a User's local computer.
 In either the development of software or data, successive versions -- called ``releases'' -- are distributed as snapshots of the generated product at a particular point in time.
 
-The similarity in workflow between software and data allows us to deploy the re-purpose some of the same technological platforms that are used to maintain and distribute versions of a software product to maintain and distribute versions of a dataset (Table \ref{tab:technologies}).
+The similarity in workflow between software and data allows us to deploy the re-purpose some of the same technological platforms that are used to maintain and distribute versions of a software product to maintain and distribute versions of an evolving dataset (Table \ref{tab:technologies}).
 Importantly, these tools are available free of charge for open source projects and already well developed -- ensuring high-level performance and stability.
@@
   \item Raw data files are stored under version control in a \texttt{git} repository --  a free and leading version control system used in software development -- by the dataset Developers.
-All the files that go together to build a single dataset are stored in the repository, together any code used to manipulate these files to create the dataset that is ultimately distributed.
+All the files that go together to build a single dataset are stored in the repository, together with any code used to manipulate these files to create the dataset that is ultimately distributed.
   \item Changes to the raw data files and code are tracked by the Developers using  \texttt{git}'s ability to make ``commits'' -- granular and annotated snapshots of the source files over time.
-  \item The \texttt{git} repository is hosted on \texttt{GitHub} -- a leading platform for hosting, enabling multiple Developers or other Contributors to work collaboratively on improving a dataset.
-  \item Developers use the files in the repository to make a release of the dataset -- a snapshot of the generated data product at a particular commit -- and upload these to \texttt{GitHub}, where they are hosted alongside the raw files and labelled using ``semantic versioning''.
+  \item The \texttt{git} repository is hosted on \texttt{GitHub} -- a leading platform for hosting, enabling multiple Developers or other Contributors to work collaboratively on improving a dataset (Fig.
+\ref{fig:technology_stack}).
+  \item Developers use the files in the repository to make a release of the dataset -- a snapshot of the generated data product at a particular commit -- and upload these to \texttt{GitHub}, where they are hosted alongside the raw files and (optionally) labelled using ``semantic versioning''.
 The version labels indicate both the ordering of versions and the magnitude of change expected between different versions.
@@
 
-\subsection{Hosting and distributing versioned data}
+\subsection{Hosting and distributing versions of an evolving dataset}
 
@@
 % could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
-Semantic versioning allows dataset Developers to communicate to Users the types of changes that have occurred between successive versions of a dataset.}
+Semantic versioning allows dataset Developers to communicate to Users the types of changes that have occurred between successive versions of an evolving dataset.}
 \label{fig:semantic}
@@
 
-Specifically, we suggest applying the theory of semantic versioning, developed for software distribution (\href{http://semver.org/}{semver.org}), to successive releases of a dataset.
+Specifically, we suggest applying the theory of semantic versioning, developed for software distribution (\href{http://semver.org/}{semver.org}), to successive releases of an evolving dataset (Fig \ref{fig:semantic}).
 Semantic versioning uses a tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers.
@@
   \item {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a ``correction'' (or ``patch''), for example a small error correction which is unlikely to break, or change in a substantial way, a user's analyses, although there may be minor changes in the results.
-  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor enhancement'', for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset curator it might change the results of an analysis done with the prior version and so warrants closer inspection.
+  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor enhancement'', for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset Developer it might change the results of an analysis done with the prior version and so warrants closer inspection.
   \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a ``major update'', for example improving the entire structure of the dataset, or adding of new columns.
@@
 \end{itemize}
-Any user of a dataset where releases were tagged using semantic versioning, would immediately know the maintainers' view of the types of the magnitude of the changes between versions.
-Note that as with software this taxonomy of changes is a language to more nuanced way to communicate from the maintainer to the user, rather than a set of fixed rules.
-Because users of data--like users of software--do unpredictable things, the maintainer may classify a change as minor, but still could be break some downstream analyses.
+Any user of an evolving dataset where releases were tagged using semantic versioning, would immediately know the Developer's view of the types of the magnitude of the changes between versions.
+Note that as with software this taxonomy of changes is a language to more nuanced way to communicate from the Developer to the User, rather than a set of fixed rules.
+Because users of data--like users of software--do unpredictable things, the Developer may classify a change as minor, but still could be break some downstream analyses.
 
@@
 
-Our package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), facilitates access to releases of any versioned data hosted on \texttt{GitHub} (Fig.
+Our package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), facilitates access to releases of any evolving dataset hosted on \texttt{GitHub} (Fig.
 \ref{fig:technology_stack}).
@@
 Accessing a different version of the data involves changing only the version number.
-From the user's perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \cite{Duursma-2016,Falster-2016}.
+From the user's perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible [e.g.][]\cite{Duursma-2016,Falster-2018}.
 
-Using \texttt{datastorr}, dataset Developers can set up their own \texttt{R} package to deliver the versioned dataset simply by providing:
+Using \texttt{datastorr}, dataset Developers can set up their own \texttt{R} package to deliver versions of an evolving dataset simply by providing:
 \begin{enumerate}
@@
 The package functions by fetching each data version once (the first time it is requested), and then caching these files locally for future reuse.
-Moreover, users can store several versions of a dataset on their computer and unambiguously access different versions with single function.
+Moreover, users can store several versions of an evolving dataset on their computer and unambiguously access different versions with single function.
 
@@
 \centering
-\caption{Goals and requirements of different parties involved in the creation and use of a versioned dataset.}
+\caption{Goals and requirements of different parties involved in creating and using and evolving dataset.}
 \vspace{0.2cm}
-  \begin{tabular}{p{3cm}p{7cm}p{6cm}}
+  \begin{tabular}{p{3cm}p{5cm}p{8cm}}
   \hline
@@
   \textbf{Group} & \textbf{Primary goal} & \textbf{Requirements} \\ \hline
-  Developer & Create and distribute versions of a dataset & Low technical overhead \\
+  Developer & Create and distribute versions of an evolving dataset & Low technical overhead \\
     & & Low initial and ongoing cost and maintenance \\
@@
     & & Long term preservation \\
-  Contributor & Contribute to future versions of a dataset & Add new data \\
+  Contributor & Contribute to future versions of an evolving dataset & Add new data \\
     & & Report errors in existing data \\
-  Users (all) & Easy access to all versions of a dataset & Background information\\
+  Users (all) & Easy access to all versions of an evolving dataset & Access metadata and background information\\
+    & & Access to all versions of a dataset\\
+    & & Ability to give feedback and contribute \\
     & & Long term stability \\
-    & & Ability to give feedback and contribute \\
-  Users (programmatic) & Reproduce products using specific versions of a dataset & Programmatic access to all versions of a dataset \\
+  Users (programmatic) &  As above, plus\\
+    & & Programmatic access to all versions of an evolving dataset \\
+    & & Reproduce products using specific versions of an evolving dataset \\
     & & Easy installation \\
-    & & Long term stability \\
   \hline
@@
 
-The key issue we are dealing with in this article may be familiar to many readers: many datasets are constantly evolving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is as yet little consensus about how to maintain and distribute multiple versions of a dataset, especially for small research teams.
+The key issue we are dealing with in this article may be familiar to many readers: many datasets are constantly evolving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is as yet little consensus about how to maintain and distribute multiple versions of an evolving dataset, especially for small research teams.
 While such teams could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required are discouraging.
-Moreover, existing data sharing platforms offer a limited set of features for the delivery of successive versions of a dataset.
-This suggests there is a substantial need for an easy, cheap, and scalable solution for maintaining and distributing successive versions of a dataset.
+Moreover, existing data sharing platforms offer a limited set of features for the delivery of successive versions of an evolving dataset.
+This suggests there is a substantial need for an easy, cheap, and scalable solution for maintaining and distributing successive versions of an evolving dataset.
 By adopting open-source and scalable practices from software development, we believe a workable system already largely exists.
 To aid this process, we created the \texttt{datastorr} package to deliver dataset versions directly into the \texttt{R} environment.
-The approach and package is already being used to deliver several versioned datasets spanning a wide range of topics (Table \ref{tab:examples}).
+The approach and package is already being used to deliver versions of several evolving datasets spanning a wide range of topics (Table \ref{tab:examples}).
 Moreover, as it builds off established and open source software and data science platforms (Table \ref{tab:technologies}), the proposed system is already easy to deploy on relatively large scale.
@@
 Our contribution here connects with a growing number of recommendations and technologies supporting the publishing and reuse of evolving data.
-Such contributions include community guidance on good practice in data citation \cite{Force11-2014} and the FAIR principles for making datasets Findable, Accessible, Interoperable, and Reusable by both machines and humans \cite{Wilkinson-2016}.
+Such contributions include community guidance on good practice in data curation \cite{Goodman-2014, Lowndes-2017}, data citation \cite{Force11-2014} and the FAIR principles for making datasets Findable, Accessible, Interoperable, and Reusable by both machines and humans \cite{Wilkinson-2016}.
 In our proposed system, information about appropriate attribution for any dataset  (whatever that is determined to be) should be made readily available, either on the landing page within GitHub, or even better distributed as part of the versioned dataset itself.
@@
 
+While our package \texttt{datastorr} offers an easy way for users of the \texttt{R} ecosystem to directly access dataset version, users of other languages can also access the datasets.
+Moreover, packages similar to \texttt{datastorr} would ideally be developed to make accessing dataset versions as easy as it is with \texttt{datastorr}.
+
 Within the \texttt{R} ecosystem, the \texttt{datastorr} package complements other approaches for creating and delivering datasets.
-One common approach used within \texttt{R} is to embed data directly within an \texttt{R} package, which can then be distributed via the Comprehensive R Archive Network (\smurl{cran.r-project.org/}).
+One common approach used within \texttt{R} is to embed data directly within an \texttt{R} package, which can then be distributed via the Comprehensive R Archive Network (\smurl{cran.r-project.org}).
 Moreover, dedicated packages are being developed to assist dataset Developers in creating data packages \cite{Finak-2018}.
@@
 Notably, datasets must be under 5MB, and only one version of a dataset package can be installed on any given machine at any one time.
-\texttt{datastorr} may therefore offer a viable approach for overcoming these limitations.
+\texttt{datastorr} offers a viable approach for overcoming these limitations.
 
-There are also many emerging or alternative technologies that offer other possible ways to implement a system for storing and distributing versions of a dataset.
+There are also many emerging or alternative technologies that offer other possible ways to implement a system for storing and distributing versions of an evolving dataset.
 Our solution currently emphasises the platform GitHub, but similar functions could be achieved via other \texttt{git} hosts such as \href{http://bitbucket.org}{bitbucket.org} and \href{http://gitlab.com}{gitlab.com}.
@@
 
-The key here is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data, which may be achieved with all of these approaches.
+The key here is not the specific technology, but rather the concept of creating, maintaining, and distributing versions of an evolving dataset, which may be achieved with all of these approaches.
 Indeed, as with every technology the best available approach is certain to evolve, especially as emerging technologies facilitate even better delivery of data in the future.

### 2018-10-10 ms.tex
lower case of names

@@
 In brief, the workflow we present here borrows best practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versions of an evolving dataset.
-Our approach envisions multiple parties involved in the creation and/or use of a versioned dataset, including \emph{Developers}, \emph{Contributors} and \emph{Users} (Fig.
+Our approach envisions multiple parties involved in the creation and/or use of a versioned dataset, including \emph{developers}, \emph{contributors} and \emph{users} (Fig.
 \ref{fig:technology_stack}).
 Each of these will likely have different goals and requirements (see Table \ref{tab:user_requirements}).
-When building a piece of software, Developers maintain a core set of code which produces the binary executable file that is eventually installed on a User's local computer.
-Analogously, Developers of an evolving dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e., loaded) on a User's local computer.
+When building a piece of software, developers maintain a core set of code which produces the binary executable file that is eventually installed on a user's local computer.
+Analogously, developers of an evolving dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e., loaded) on a user's local computer.
 In either the development of software or data, successive versions -- called ``releases'' -- are distributed as snapshots of the generated product at a particular point in time.
@@
 \begin{itemize}
-  \item Raw data files are stored under version control in a \texttt{git} repository --  a free and leading version control system used in software development -- by the dataset Developers.
+  \item Raw data files are stored under version control in a \texttt{git} repository --  a free and leading version control system used in software development -- by the dataset developers.
 All the files that go together to build a single dataset are stored in the repository, together with any code used to manipulate these files to create the dataset that is ultimately distributed.
-  \item Changes to the raw data files and code are tracked by the Developers using  \texttt{git}'s ability to make ``commits'' -- granular and annotated snapshots of the source files over time.
-  \item The \texttt{git} repository is hosted on \texttt{GitHub} -- a leading platform for hosting, enabling multiple Developers or other Contributors to work collaboratively on improving a dataset (Fig.
+  \item Changes to the raw data files and code are tracked by the developers using  \texttt{git}'s ability to make ``commits'' -- granular and annotated snapshots of the source files over time.
+  \item The \texttt{git} repository is hosted on \texttt{GitHub} -- a leading platform for hosting, enabling multiple developers or other contributors to work collaboratively on improving a dataset (Fig.
 \ref{fig:technology_stack}).
-  \item Developers use the files in the repository to make a release of the dataset -- a snapshot of the generated data product at a particular commit -- and upload these to \texttt{GitHub}, where they are hosted alongside the raw files and (optionally) labelled using ``semantic versioning''.
+  \item developers use the files in the repository to make a release of the dataset -- a snapshot of the generated data product at a particular commit -- and upload these to \texttt{GitHub}, where they are hosted alongside the raw files and (optionally) labelled using ``semantic versioning''.
 The version labels indicate both the ordering of versions and the magnitude of change expected between different versions.
-  \item Using the \texttt{datastorr} package, Users can both retrieve a list of all available versions of the dataset, and retrieve particular versions of the dataset on demand, and load them directly into \texttt{R}.
+  \item Using the \texttt{datastorr} package, users can both retrieve a list of all available versions of the dataset, and retrieve particular versions of the dataset on demand, and load them directly into \texttt{R}.
 \end{itemize}
@@
 Here we focus on the platform \texttt{GitHub} (Table \ref{tab:technologies}).
-Hosting of a \texttt{git} repository enables dataset Developers to connect with other potential Contributors and also Users (Fig.
+Hosting of a \texttt{git} repository enables dataset developers to connect with other potential contributors and also users (Fig.
 \ref{fig:technology_stack}).
@@
 Of particular interest for current purposes, is \texttt{GitHub}'s ability to host a stream of releases from the dataset, alongside the \texttt{git} repository containing all the raw files.
-Each release is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset Developer decided to generate a new version of the data for distribution.
-While Users could in principle download the entire \texttt{git} repository, actually all they want (most of the time) are the releases.
+Each release is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset developer decided to generate a new version of the data for distribution.
+While users could in principle download the entire \texttt{git} repository, actually all they want (most of the time) are the releases.
 
-Deciding when to make a new release is at the discretion of the dataset Developer.
-In practice, one makes fewer releases than one does commits into the \texttt{git} repository, though there is nothing stopping Developers from releasing a new version for every commit.
-The flexibility here allows Developers to do internal work between releases and only release the data to Users when the revision represents a clear improvement on the previous release.
+Deciding when to make a new release is at the discretion of the dataset developer.
+In practice, one makes fewer releases than one does commits into the \texttt{git} repository, though there is nothing stopping developers from releasing a new version for every commit.
+The flexibility here allows developers to do internal work between releases and only release the data to users when the revision represents a clear improvement on the previous release.
 
-Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data Users accessing the data: those that interact with the data via point and click downloading, and those that use programmatic interaction (Fig.
+Another important consideration is that websites like \texttt{GitHub} naturally cater for two types of data users accessing the data: those that interact with the data via point and click downloading, and those that use programmatic interaction (Fig.
 \ref{fig:technology_stack}, Table \ref{tab:user_requirements}).
-Specifically, \texttt{GitHub} releases can be downloaded directly by Users or accessed programmatically via the \texttt{GitHub} API.
+Specifically, \texttt{GitHub} releases can be downloaded directly by users or accessed programmatically via the \texttt{GitHub} API.
 
@@
 % could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
-Semantic versioning allows dataset Developers to communicate to Users the types of changes that have occurred between successive versions of an evolving dataset.}
+Semantic versioning allows dataset developers to communicate to users the types of changes that have occurred between successive versions of an evolving dataset.}
 \label{fig:semantic}
@@
 
-To realize the full benefits of a versioned controlled dataset, Users should be able to easily intuit the types of changes that have occurred among versions.
+To realize the full benefits of a versioned controlled dataset, users should be able to easily intuit the types of changes that have occurred among versions.
 Since software development has effectively already dealt with almost identical; problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.
@@
   \item {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a ``correction'' (or ``patch''), for example a small error correction which is unlikely to break, or change in a substantial way, a user's analyses, although there may be minor changes in the results.
-  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor enhancement'', for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset Developer it might change the results of an analysis done with the prior version and so warrants closer inspection.
+  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor enhancement'', for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset developer it might change the results of an analysis done with the prior version and so warrants closer inspection.
   \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a ``major update'', for example improving the entire structure of the dataset, or adding of new columns.
@@
 \end{itemize}
-Any user of an evolving dataset where releases were tagged using semantic versioning, would immediately know the Developer's view of the types of the magnitude of the changes between versions.
-Note that as with software this taxonomy of changes is a language to more nuanced way to communicate from the Developer to the User, rather than a set of fixed rules.
-Because users of data--like users of software--do unpredictable things, the Developer may classify a change as minor, but still could be break some downstream analyses.
+Any user of an evolving dataset where releases were tagged using semantic versioning, would immediately know the developer's view of the types of the magnitude of the changes between versions.
+Note that as with software this taxonomy of changes is a language to more nuanced way to communicate from the developer to the user, rather than a set of fixed rules.
+Because users of data--like users of software--do unpredictable things, the developer may classify a change as minor, but still could be break some downstream analyses.
 
@@
 
-Using \texttt{datastorr}, dataset Developers can set up their own \texttt{R} package to deliver versions of an evolving dataset simply by providing:
+Using \texttt{datastorr}, dataset developers can set up their own \texttt{R} package to deliver versions of an evolving dataset simply by providing:
 \begin{enumerate}
@@
 
-Then as the dataset grows over time, the Developers update the \texttt{git} repository and create a \texttt{GitHub} release with a new version number.
+Then as the dataset grows over time, the developers update the \texttt{git} repository and create a \texttt{GitHub} release with a new version number.
 All the releases are simultaneously available to any user, both point-and-click and programmatically.
@@
     & & Easy workflow for releasing new versions \\
-    & & Enable User feedback in error checking and contributions \\
+    & & Enable user feedback in error checking and contributions \\
     & & Long term preservation \\
@@
 One common approach used within \texttt{R} is to embed data directly within an \texttt{R} package, which can then be distributed via the Comprehensive R Archive Network (\smurl{cran.r-project.org}).
-Moreover, dedicated packages are being developed to assist dataset Developers in creating data packages \cite{Finak-2018}.
+Moreover, dedicated packages are being developed to assist dataset developers in creating data packages \cite{Finak-2018}.
 An advantage of this approach, compared to ours, is that the data are immediately available in the package (whereas our packages only contain instructions for fetching the data).
@@
 Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}.
-Multiple Developers can make changes to different parts of the code (or, in our case, data) and the \texttt{git} system will integrate these together or, when needed, flag where there are conflicts that need to be resolved.
+Multiple developers can make changes to different parts of the code (or, in our case, data) and the \texttt{git} system will integrate these together or, when needed, flag where there are conflicts that need to be resolved.
 The proposed system of data delivery thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.

### 2018-10-10 ms.tex
Refine section on semantic versioning

@@
 Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Force11-2014, Lowndes-2017, Perkel-2016, VanNoorden-2013, Kratz-2015,Wilkinson-2016, Yenni-2018}.
-In particular, in some areas, such as our own area of ecology and evolution, we are only starting to support the fact that many high-quality datasets used in research should be evolving entities \cite{Yenni-2018}.
+In particular, in some areas, such as our own area of ecology and evolution, we are only starting to support the fact that some high-quality datasets may be evolving entities \cite{Yenni-2018}.
 
@@
 A common approach for publishing a dataset is to release it under a Digital Object Identifier (DOI) in a standalone data repository, such as DataDryad, Figshare, and Zenodo.
-While these platforms did not initially support versioning of datasets, all have subsequently added the ability to support multiple versions of a dataset, either under a single or different DOIs \cite{Nielsen-2017}.
+While these platforms did may not all initially support versioning of datasets, they now support multiple versions of a dataset, either under a single or different DOIs \cite{Nielsen-2017}.
 Yet, while these new features in principle allow users to access multiple versions of a dataset, the release, discovery, and access to multiple versions of a dataset is not always straightforward.
@@
 To realize the full benefits of a versioned controlled dataset, users should be able to easily intuit the types of changes that have occurred among versions.
-Since software development has effectively already dealt with almost identical; problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.
+Since software development has effectively already dealt with a similar problem in the labelling of software releases, we suggest there is benefit in adopting the best-practices from that field.
 
-Specifically, we suggest applying the theory of semantic versioning, developed for software distribution (\href{http://semver.org/}{semver.org}), to successive releases of an evolving dataset (Fig \ref{fig:semantic}).
+Specifically, we suggest adapting the process of semantic versioning, developed for software distribution (\href{http://semver.org/}{semver.org}), to labelling of successive releases of an evolving dataset (Fig \ref{fig:semantic}).
 Semantic versioning uses a tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers.
-For example, version ``1.0.0''.
-Changes in the version number then signal the type of changes that occurred in the dataset (Fig \ref{fig:semantic}).
-Applying semantic versioning to data, a change from:
+For example, version ``2.1.2''.
+Software developers increment particular numbers in a systematic way to communicate any changes to the public API for a particular piece of software.
+
+There are two natural advantages of semantic versioning of adapting the process of semantic versioning for dataset development.
+The first is that enables natural ordering of releases.
+The second, is that it enables developers to signal the type and magnitude of change that occurred in the product between successive versions.
+Seeing a series of version numbers, users of an evolving dataset know the developer's view on the type and/or magnitude of change between versions.
+
+Though the analogy to software is not perfect, datasets can also be thought of as having an Interface, which is the structure that allows for interacting with the actual data.
+For example, in tabular data the structure would be the column and row labels, and the data the actual cell contents.
+Changes to the Interface an then be labelled with semantic versioning.
+
+Building off the guidelines for semantic versioning of software, we suggest the following guidelines for labelling a dataset with semantic versioning.
 \begin{itemize}
-  \item {\bf 1.0.0 $\rightarrow$ 1.0.1}: implies a ``correction'' (or ``patch''), for example a small error correction which is unlikely to break, or change in a substantial way, a user's analyses, although there may be minor changes in the results.
-  \item {\bf 1.0.0 $\rightarrow$ 1.1.0}: implies a ``minor enhancement'', for example adding a new study to a meta-analysis dataset (while otherwise maintaining the same dataset structure); this change is large enough that in the view of the dataset developer it might change the results of an analysis done with the prior version and so warrants closer inspection.
-  \item {\bf 1.0.0 $\rightarrow$ 2.0.0}: implies a ``major update'', for example improving the entire structure of the dataset, or adding of new columns.
-These changes are likely to change the results of most user's analyses or to break code that was written to work on the previous version of the dataset.
+  \item Clearly communicate the Interface to the dataset in the metadata or landing page. This includes file types, dataset structure, and element names.
+  \item Use versions beginning with ``0.Y.Z'' to indicate products where the Interface is still in development.
+  \item Version ``1.0.0'' defines the public Interface. 
+  \item Once defined, increment version numbers to communicate any changes to the Interface.
+  \item Increment the "Major" version when you make changes to the Interface that are likely incompatible with any code written to work with previous versions (i.e.
+causes an error).
+Such changes may include revising the file names, the structure of the dataset, or changing element names (e.g.
+column headers).
+  \item Increment the "Minor" version to communicate any changes to the Interface that are likely to be compatible with any code written to work with the previous versions (i.e.
+allows code to run without error).
+Such changes typically involve adding new data within the existing structure, so that the previous dataset version exists as a subset of the new version.
+For tabular data, this includes adding columns or rows.
+On the other hand, removing data should constitute a major version, as records previously relied on may no longer exists.
+  \item Increment the "Patch" version to communicate correction of errors in the actual data, without any changes to the Interface.
+Such changes are unlikely to break, or change in a substantial way, analyses written with the previous version, although there may of course be changes in the results.
+  \item Once a dataset version has been released, do not modify it. Further modifications are released under a new version number.
 \end{itemize}
-Any user of an evolving dataset where releases were tagged using semantic versioning, would immediately know the developer's view of the types of the magnitude of the changes between versions.
-Note that as with software this taxonomy of changes is a language to more nuanced way to communicate from the developer to the user, rather than a set of fixed rules.
-Because users of data--like users of software--do unpredictable things, the developer may classify a change as minor, but still could be break some downstream analyses.
+
+While these guidelines will help users in understanding the types of changes that have occurred between successive versions of a dataset, \emph{any} change in a dataset may alter the results of a users' analysis in non-trivial ways.
+Unlike developers of software, developers of a dataset cannot guarantee full backwards-compatibility, i.e.
+that certain results will remain unchanged in updated versions.
+We suggest responsibility for verifying how different versions of an evolving dataset influence their particular analysis or use thus always remains with the user, even if simply applying a so-called ``patch''.
+While further work -- and likely experience -- is needed to refine the process of semantic versioning for datasets to further develop understanding between data developers and data users of what different changes imply, semantic versioning still provides a more nuanced way to communicate from the developer to the user on the types of change they could expect.
 

### 2019-02-07 ms.tex
Changes to address reviewers concerns

@@
 The sharing and re-use of data has become a cornerstone of modern science.
-Multiple platforms now allow quick and easy data sharing.
-So far, however, data publishing offers limited functions for interacting with evolving datasets - those that continue to grow with time as more records are added, errors fixed, and new data structures are created.
+Multiple platforms now allow easy publication of datasets.
+So far, however, platforms for data sharing offers limited functions for distributing and interacting with evolving datasets - those that continue to grow with time as more records are added, errors fixed, and new data structures are created.
 In this article, we describe a workflow for maintaining and distributing successive versions of an evolving dataset, allowing users to retrieve and load different versions directly into the R platform.
@@
 
-Publication of a quality dataset -- a collection of measurements, stored in one or several files -- is now considered a first-class scientific product.
+Sharing of a quality dataset -- a collection of measurements, stored in one or several files -- is now considered a first-class scientific output.
 Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing datasets, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}.
 Evidence of this trend is seen in the increasing numbers of standalone ``Data papers'' appearing in both standard domain-level journals and specialised data journals.
-Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Force11-2014, Lowndes-2017, Perkel-2016, VanNoorden-2013, Kratz-2015,Wilkinson-2016, Yenni-2018}.
+Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing, the scientific community is still grappling with how to effectively maintain and distribute open-source datasets \cite{Whitlock-2011, Goodman-2014, Force11-2014, Lowndes-2017, Perkel-2016, VanNoorden-2013, Kratz-2015,Wilkinson-2016, Yenni-2018}.
 In particular, in some areas, such as our own area of ecology and evolution, we are only starting to support the fact that some high-quality datasets may be evolving entities \cite{Yenni-2018}.
 
-An evolving (or `living') dataset is one that is subject to recurrent or occasional change.
+An evolving (or `living') dataset is one that is subject to occasional or recurrent change.
 Typical changes may include improving the quality of existing data, adding new data, re-structuring the dataset content, or integrating with other datasets.
@@
 Even datasets that are not initially envisioned as evolving, may become so as minor errors are identified and corrected during use.
-In either case, the most recent version of the dataset will typically contain the best-available information, but there are many reasons to go back to previous versions: to replicate previous analyses or to work on a stable version for downstream analyses or visualisation.
+In either case, the most recent version of the dataset will typically contain the best-available information, but there are still reasons to go back to previous versions: to replicate previous analyses or to work on a stable version for downstream analyses or visualisation.
 
@@
 In the recent past, small research groups have solved the issue of versioning data internally and informally, for example by emailing around the latest version.
-However, as science grows and moves towards more systematic publishing of data, scalable solutions are needed to distribute dataset versions to a wider variety of users.
+However, as science grows and moves towards more systematic sharing of data, scalable solutions are needed to distribute dataset versions to a wider variety of users.
 
 One approach taken by large research consortia has been to create dedicated web servers for archiving and delivering of data.
-Projects such as the Sloan Digital Sky Survey (\smurl{https://www.sdss.org}) have sophisticated infrastructure and processes for storing and distributing successive versions of very large datasets \cite{Abolfathi-2018}.
-The issue of updating and expanding published data has also been addressed in some centralised repositories, like genetic sequences (via \texttt{GenBank}), where new data can be added and there exist abilities to correct errors in existing records.
+Projects such as the Sloan Digital Sky Survey (\smurl{https://www.sdss.org}) have sophisticated infrastructure and processes for distributing successive versions of very large datasets \cite{Abolfathi-2018}.
+The issue of updating data has also been addressed in some centralised repositories, like genetic sequences (via \texttt{GenBank}), where new data can be added and there exist abilities to correct errors in existing records.
 Yet these web databases require a level of funding and technological infrastructure that is beyond most research groups.
 
-The vast majority of research projects are smaller and these currently rely on more generic data repositories for publishing of data.
-A common approach for publishing a dataset is to release it under a Digital Object Identifier (DOI) in a standalone data repository, such as DataDryad, Figshare, and Zenodo.
-While these platforms did may not all initially support versioning of datasets, they now support multiple versions of a dataset, either under a single or different DOIs \cite{Nielsen-2017}.
+The vast majority of research projects are smaller and these currently rely on more generic data repositories for distributing data.
+A common approach for distributing a dataset is to release it under a Digital Object Identifier (DOI) in a standalone data repository, such as DataDryad, Figshare, and Zenodo.
+While these platforms did not all initially support versioning of datasets, they now support multiple versions of a dataset, either under a single or different DOIs \cite{Nielsen-2017}.
 Yet, while these new features in principle allow users to access multiple versions of a dataset, the release, discovery, and access to multiple versions of a dataset is not always straightforward.
 
-We believe more can be done to streamline the delivery of a potentially large number of dataset versions to users, especially for small research teams with limited budgets.
+We believe more can be done to streamline the distribution of a potentially large number of dataset versions to users, especially for small research teams with limited budgets.
 There are at least three challenges.
@@
 Analogously, developers of an evolving dataset maintain a core set of files (the ``code''), which produces an organised dataset that can be ``installed'' (i.e., loaded) on a user's local computer.
-In either the development of software or data, successive versions -- called ``releases'' -- are distributed as snapshots of the generated product at a particular point in time.
+In the development of either software or data, successive versions -- called ``releases'' -- are distributed as snapshots of the generated product at a particular point in time.
 
@@
 \ref{fig:technology_stack}).
-  \item developers use the files in the repository to make a release of the dataset -- a snapshot of the generated data product at a particular commit -- and upload these to \texttt{GitHub}, where they are hosted alongside the raw files and (optionally) labelled using ``semantic versioning''.
+  \item Developers use the files in the repository to make a release of the dataset -- a snapshot of the generated data product at a particular commit -- and upload these to \texttt{GitHub}, where they are hosted alongside the raw files and (optionally) labelled using ``semantic versioning''.
 The version labels indicate both the ordering of versions and the magnitude of change expected between different versions.
   \item Using the \texttt{datastorr} package, users can both retrieve a list of all available versions of the dataset, and retrieve particular versions of the dataset on demand, and load them directly into \texttt{R}.
+  \item Those not using \texttt{R} can also access versions from \texttt{GitHub}.
 \end{itemize}
@@
   \item Developers establish a separate \texttt{git} repository for each dataset to be distributed.
-  \item Saving all files as plain text, so that \texttt{git} can identify line-by-line changes. For example, save tabular data as a ``csv''.
   \item Saving data in their rawest form. In some datasets you might only have a single file. Others may have may files that get manipulated or combined in some way to produce a unified product.
+  \item Where possible, saving all files as plain text, so that \texttt{git} can identify line-by-line changes.
+For example, save tabular data as a ``csv''.
+While this approach works well for small-to-intermediate sized files, those with larger files may prefer to use a compressed format to reduce repository size and bandwidth.
   \item Including in the \texttt{git} repository any code needed to manipulate or compile the raw data files into the final dataset.
@@
 
-Of particular interest for current purposes, is \texttt{GitHub}'s ability to host a stream of releases from the dataset, alongside the \texttt{git} repository containing all the raw files.
+Another notable feature of \texttt{GitHub} is the ability to host a stream of releases from the dataset, alongside the \texttt{git} repository containing all the raw files.
 Each release is linked to a specific commit in the \texttt{git} repository history and occur at points where the dataset developer decided to generate a new version of the data for distribution.
-While users could in principle download the entire \texttt{git} repository, actually all they want (most of the time) are the releases.
+While users could in principle download the entire \texttt{git} repository, most of the time, what they want are the releases.
 
@@
 \caption{
-% could use a more descriptive legend. Maybe walk the reader through it a bit more. Also, I didn’t see (a) and (b) in that figure.
-Semantic versioning allows dataset developers to communicate to users the types of changes that have occurred between successive versions of an evolving dataset.}
+Semantic versioning allows dataset developers to communicate to users the types of changes that have occurred between successive versions of an evolving dataset, using a tri-digit label where increments in a number indicate MAJOR-, MINOR- and PATCH-level changes, respectively.
+See text for further details.}
 \label{fig:semantic}
@@
    \textbf{\texttt{GitHub} repository} & \textbf{Dataset description} \\ \hline
-  \ghsmurl{dfalster/baad.data} & Size dimensions of plants for many species from across the world \cite{Falster-2015}.\\
   \ghsmurl{traitecoevo/taxonlookup} & Taxonomy of world's land plants \cite{Pennell-2015a}.\\
+  \ghsmurl{traitecoevo/growthform} & Growth form of world's land plants \cite{Taseski-2019}.\\
+  \ghsmurl{traitecoevo/baad.data} & Size dimensions of plants for many species from across the world \cite{Falster-2015}.\\
   \ghsmurl{ecohealthalliance/cites} & Trade details from Convention on International Trade in Endangered Species (CITES)\\
   \ghsmurl{madams1/nbadata} & Statistics from the National Basketball Association (NBA) seasons 1996-97 to 2016-17 \\
-  \ghsmurl{madams1/floridainmates} & Dataset from the Florida Department of Corrections on the state's inmate population\\
-  \ghsmurl{traitecoevo/fungaltraits} & Dataset on the traits of fungi \cite{Cornwell-2018}.\\
+  \ghsmurl{madams1/floridainmates} & Statistics on Florida state's inmate population, from Florida Department of Corrections \\
+  \ghsmurl{traitecoevo/fungaltraits} & Traits of world's fungi species \cite{Cornwell-2018}.\\
 
@@
 
-Specifically, we suggest adapting the process of semantic versioning, developed for software distribution (\href{http://semver.org/}{semver.org}), to labelling of successive releases of an evolving dataset (Fig \ref{fig:semantic}).
-Semantic versioning uses a tri-digit notation of the form ``X.Y.Z'' for successive versions, where X, Y, and Z are non-negative integers.
+Specifically, we suggest adapting the process of semantic versioning, developed for labelling successive releases of software (see \href{http://semver.org/}{semver.org}), to labelling of successive releases of an evolving dataset (Fig \ref{fig:semantic}).
+In semantic versioning of software, a tri-digit label of the form ``X.Y.Z'' is applied to each version, where X, Y, and Z are non-negative integers.
 For example, version ``2.1.2''.
-Software developers increment particular numbers in a systematic way to communicate any changes to the public API for a particular piece of software.
+Although everyday practice may differ, the guidelines at \href{http://semver.org/}{semver.org} suggest labels are incremented in a particular way, determined by changes in the public API for the software.
 
-There are two natural advantages of semantic versioning of adapting the process of semantic versioning for dataset development.
+Though the analogy to software is not perfect, datasets can also be thought of as having an ``Interface'', determined by the structure of the dataset, which dictates how users interact with the resource.
+For example, in tabular data the structure is determined by the names of different files, the column labels within each, and the presence of different subgroups within the table (as indicated by labels in particular columns).
+Successive versions of a dataset can then be labelled in a manner analogous to that of software, determined by the structure of the dataset and changes in that structure.
+
+There are two natural advantages of adapting the process of semantic versioning for dataset development.
 The first is that enables natural ordering of releases.
@@
 
-Though the analogy to software is not perfect, datasets can also be thought of as having an Interface, which is the structure that allows for interacting with the actual data.
-For example, in tabular data the structure would be the column and row labels, and the data the actual cell contents.
-Changes to the Interface an then be labelled with semantic versioning.
-
-Building off the guidelines for semantic versioning of software, we suggest the following guidelines for labelling a dataset with semantic versioning.
+Drawing inspiration from the guidelines for semantic versioning of software at \href{http://semver.org/}{semver.org}, we suggest the following guidelines for labelling of a dataset with semantic versioning:
 \begin{itemize}
-  \item Clearly communicate the Interface to the dataset in the metadata or landing page. This includes file types, dataset structure, and element names.
-  \item Use versions beginning with ``0.Y.Z'' to indicate products where the Interface is still in development.
-  \item Version ``1.0.0'' defines the public Interface. 
-  \item Once defined, increment version numbers to communicate any changes to the Interface.
-  \item Increment the "Major" version when you make changes to the Interface that are likely incompatible with any code written to work with previous versions (i.e.
-causes an error).
+  \item Clearly communicate the structure of the dataset in the metadata or landing page. This includes file types, data type, and element names.
+  \item Use versions beginning with ``0.Y.Z'' to indicate products where the structure is still in development.
+  \item Version ``1.0.0'' defines the structure. 
+  \item Once defined, increment version numbers to communicate any changes to the structure.
+  \item Increment the "Major" version when you make changes to the structure that are likely incompatible with any code written to work with previous versions.
 Such changes may include revising the file names, the structure of the dataset, or changing element names (e.g.
 column headers).
-  \item Increment the "Minor" version to communicate any changes to the Interface that are likely to be compatible with any code written to work with the previous versions (i.e.
+Substantial additions of data might also be considered a major change to structure, especially where they add new subgroups to the dataset.
+  \item Increment the "Minor" version to communicate any changes to the structure that are likely to be compatible with any code written to work with the previous versions (i.e.
 allows code to run without error).
-Such changes typically involve adding new data within the existing structure, so that the previous dataset version exists as a subset of the new version.
+Such changes might involve adding new data within the existing structure, so that the previous dataset version exists as a subset of the new version.
 For tabular data, this includes adding columns or rows.
 On the other hand, removing data should constitute a major version, as records previously relied on may no longer exists.
-  \item Increment the "Patch" version to communicate correction of errors in the actual data, without any changes to the Interface.
-Such changes are unlikely to break, or change in a substantial way, analyses written with the previous version, although there may of course be changes in the results.
+  \item Increment the "Patch" version to communicate correction of errors in the actual data, without any changes to the structure.
+Such changes are unlikely to break, or change analyses written with the previous version in a substantial way.
   \item Once a dataset version has been released, do not modify it. Further modifications are released under a new version number.
@@
 
-While these guidelines will help users in understanding the types of changes that have occurred between successive versions of a dataset, \emph{any} change in a dataset may alter the results of a users' analysis in non-trivial ways.
+While it is hoped the guidelines above help users in understanding the types of changes that have occurred between successive versions of a dataset, \emph{any} change in a dataset may alter the results of a users' analysis in non-trivial ways.
 Unlike developers of software, developers of a dataset cannot guarantee full backwards-compatibility, i.e.
@@
 
-For example, \texttt{datastorr} has been used to build several packages (Table \ref{tab:examples}), including \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}.
-The R package \texttt{baad.data} consists of only a few simple functions and associated help files, that were automatically generated with \texttt{datastorr}.
+For example, \texttt{datastorr} has been used to build several packages (Table \ref{tab:examples}), including \texttt{taxonlookup} (\smurl{github.com/traitecoevo/taxonlookup}), which hosts data on the Taxonomy of world's land plants \cite{Pennell-2015a}.
+The R package \texttt{taxonlookup} consists of only a few simple functions and associated help files, that were automatically generated with \texttt{datastorr}.
 For a user, accessing a version of the data is a simple as typing a single line of code (Fig.
@@
 Accessing a different version of the data involves changing only the version number.
-From the user's perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible [e.g.][]\cite{Duursma-2016,Falster-2018}.
+From the user's perspective, the existence of the \texttt{taxonlookup} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data [e.g.][]\cite{Pennell-2015a, Mounce2018} possible.
 
@@
 \begin{enumerate}
-  \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored;
+  \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/taxonlookup'') where releases are stored;
   \item the filename in the release that contains data;
@@
 While such teams could in principle create their own dynamic web interface, the technological hurdles, cost and maintenance required are discouraging.
-Moreover, existing data sharing platforms offer a limited set of features for the delivery of successive versions of an evolving dataset.
-This suggests there is a substantial need for an easy, cheap, and scalable solution for maintaining and distributing successive versions of an evolving dataset.
+Moreover, existing platforms for distributing data offer a limited set of features for the delivery of successive versions of an evolving dataset.
+This suggests there is a need for an easy, cheap, and scalable solution for maintaining and distributing successive versions of an evolving dataset.
 By adopting open-source and scalable practices from software development, we believe a workable system already largely exists.
@@
 
-Our contribution here connects with a growing number of recommendations and technologies supporting the publishing and reuse of evolving data.
+Our contribution here connects with a growing number of recommendations and technologies supporting the sharing and reuse of evolving data.
 Such contributions include community guidance on good practice in data curation \cite{Goodman-2014, Lowndes-2017}, data citation \cite{Force11-2014} and the FAIR principles for making datasets Findable, Accessible, Interoperable, and Reusable by both machines and humans \cite{Wilkinson-2016}.
@@
 
-A central feature of the proposed system is that data are maintained in the cloud.
+A central feature of the proposed system is that data are maintained on the web.
 This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig.
 \ref{fig:technology_stack}).
-Cloud systems thus act as a central point for the collection, curation, and distribution of the data.
-Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}.
+Web platforms thus act as a central point for the collection, curation, and distribution of the data.
+Additionally, one of the greatest benefits of using web platforms like \texttt{GitHub} for development of both software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}.
 Multiple developers can make changes to different parts of the code (or, in our case, data) and the \texttt{git} system will integrate these together or, when needed, flag where there are conflicts that need to be resolved.

### 2019-03-13 ms.tex
Minor changes requested by journal

@@
 %\documentclass[letterpaper]{oup-contemporary}
-\documentclass[a4paper,num-refs]{assets/oup-contemporary}
+\documentclass[a4paper,num-refs]{oup-contemporary}
 
@@
 %%% Author Notes
-\authnote{\authfn{1}daniel.falster@unsw.edu.au}
+\authnote{\authfn{1}daniel.falster@unsw.edu.au, +61-2-93858431\\
+ORCID IDs. Daniel S Falster: 0000-0002-9814-092X, Matthew  W. Pennell: 0000-0002-2886-3970, William K. Cornwell: 0000-0003-4080-4073
+}
 
@@
 
-\section{Availability of source code and requirements}
+\section{Availability of Supporting Data}
+
+Snapshots of the code and example file are archived in the GigaScience GigaDB repository \cite{Falster-2019}.
 
@@
 
+\begin{itemize}
+\item Project name:  \texttt{datastorr example} 
+\item Project home page: \smurl{github.com/richfitz/datastorr.example}.
+\item Operating system(s): Platform independent
+\item Programming language: R
+\item License: MIT
+\end{itemize}
+
+\begin{itemize}
+\item Project name:  \texttt{taxonlookup} 
+\item Project home page: \smurl{github.com/traitecoevo/taxonlookup}.
+\item Operating system(s): Platform independent
+\item Programming language: R
+\item License: MIT
+\end{itemize}
+
 \section{Declarations}
@@
 \subsection{List of abbreviations}
-API = An Application Programming Interface provides a set of protocols for exchanging information. DOI = Digital Object Identifier.
+API = An Application Programming Interface provides a set of protocols for exchanging information. 
+CKAN = Comprehensive Knowledge Archive Network.
+DOI = Digital Object Identifier.
+FAIR =  Findable, Accessible, Interoperable, Reusable.
+Gb =  Gigabyte.
+Mb =  Megabyte.
+OKFN = Open Knowledge International.
 
@@
 
-DSF was funded by the Australian Research Council. MWP was funded by a NSERC Discovery Grant.
+DSF was funded by the Australian Research Council (FT160100113). MWP was funded by a NSERC Discovery Grant (RGPIN-2017-04590).
 

### 2019-08-26 analysis.R
add tinytex and ms folder specifications. create analysis.R file

@@ -1,0 +1,9 @@
+
+# Packages --------------------
+library(tinytex)
+
+# source function scripts -----------
+source("R/build.R")
+
+# Documents ---------------
+pdflatex("ms/manuscript.tex", pdf_file = 'ms/manuscript.pdf')
